### 2018-05-05 README.Rmd
1st commit

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# ScaledPiecesVB

This repository contains the data and code for our paper:

> Horta, P., Cascalheira, J., Bicho, N. (2018). _The role of lithic bipolar technology in Western Iberia’s Upper Paleolithic: the case of Vale Boi (southern Portugal)_. Name of journal/book <https://doi.org/xxx/xxx>

Our pre-print is online here:

> Horta, P., Cascalheira, J., Bicho, N. (2018). _The role of lithic bipolar technology in Western Iberia’s Upper Paleolithic: the case of Vale Boi (southern Portugal)_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

### How to cite

Please cite this compendium as:

> Cascalheira, J., Horta, P., Bicho, N. (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for 'The role of lithic bipolar technology in Western Iberia’s Upper Paleolithic: the case of Vale Boi (southern Portugal)'_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

### How to download or install

You can download the compendium as a zip from from this URL:

<https://github.com/jmcascalheira/ScaledPiecesVB/archive/master.zip>

This repository is organized as an R package using rrtools by Ben Marwick which can be installed from github with:

There are no actual R functions in this package - all the R code is in the Rmd file. I simply used the R package structure to help manage dependencies, to take advantage of continuous integration for automated code testing, and so I didn't have to think too much about how to organise the files.

To download the package source as you see it on GitHub, for offline browsing, use this line at the shell prompt (assuming you have Git installed on your computer):

```{r eval=FALSE}

git clone https://github.com/jmcascalheira/ScaledPiecesVB.git

```

Once the download is complete, open the ShortTermOccupations.Rproj in RStudio to begin working with the package and compendium files.

The package has a number of dependencies on other R packages, and programs outside of R. Installing these can be time-consuming and complicated, so we've included a packrat directory, which contains the source code for all the packages we depend on. If all works well, these will be installed on your computer when you open ShortTermOccupations.Rproj in RStudio.

### Licenses

**Text and figures :**  [CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/)

**Code :** See the [DESCRIPTION](DESCRIPTION) file

**Data :** [CC-0](http://creativecommons.org/publicdomain/zero/1.0/) attribution requested in reuse

### Contributions

We welcome contributions from everyone. Before you get started, please see our [contributor guidelines](CONTRIBUTING.md). Please note that this project is released with a [Contributor Code of Conduct](CONDUCT.md). By participating in this project you agree to abide by its terms.

### Dependencies:

See the colophon section of the docx file in `analysis/paper` for a full list of the packages that this project depends on.

### Contact:

João Cascalheira, Post-Doc Researcher, ICArEHB University of Algarve, Campus de Gambelas 8005-139 Faro PORTUGAL e. jmcascalheira@ualg.pt w. http://www.icarehb.com/cascalheira

### 2018-05-05 analysis/paper/paper.Rmd
1st commit

---
title: "The role of lithic bipolar technology in Western Iberia’s Upper Paleolithic: the case of Vale Boi (southern Portugal)"
author:
  - Pedro Horta (ICArEHB, University of Algarve, Faro, Portugal) pedroehorta@gmail.com
  - João Cascalheira (ICArEHB, University of Algarve, Faro, Portugal) jmcascalheira@ualg.pt
  - Nuno Bicho (ICArEHB, University of Algarve, Faro, Portugal) nbicho@ualg.pt
date: '`r Sys.Date()`'
output: 
    bookdown::word_document2: #subsistute by pdf_document2 or hmtl_document2 as necessary
      fig_caption: yes
      reference_docx: "../templates/template2.docx" 
bibliography: ScaledPiecesVB.bib
csl: "../templates/springer-humanities-author-date.csl"
abstract: |

---

```{r, setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  dpi = 300,
  fig.width = 9,
  fig.path = "../figures/"
  )

#library(ScaledPiecesVB) # Or use devtools::load_all('.', quiet = T) if your code is in script files, rather than as functions in the `/R` diretory
```

```{r load_libraries, echo = FALSE}

library(dplyr)
library(FactoMineR)
library(factoextra)
library(ggpubr)
library(ggplot2)
library(ggrepel)

```

# Introduction

\newline

```{r }

```

\newline

# Materials and Methods

# Results

#Discussion

#Conclusion

#Acknowledgements

##### pagebreak

# References 
<div id="refs"></div>

##### pagebreak

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r colophon, cache = FALSE}
# which R packages and versions?
devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? You may need to change the path value
# if your Rmd is not in analysis/paper/
git2r::repository("../..")
```

### 2018-05-06 analysis/paper/paper.Rmd
significant changes in /R and paper

@@
 
-```{r, setup, echo = FALSE}
+```{r, setup, include = FALSE, echo = FALSE}
 knitr::opts_chunk$set(
@@
   fig.width = 9,
-  fig.path = "../figures/"
-  )
+  fig.path = "../figures/")
 
@@
 
-library(dplyr)
+library(tidyverse)
 library(FactoMineR)
@@
 library(ggrepel)
+library(maptools)
+library(ggmap)
+library(devtools)
+library(legendMap)
+library(tab)
 
@@
 
+Bipolar technology is generically classified into two types of lithic artifacts - bipolar cores and scaled pieces (also known as splintered pieces or pièce esquilée).
+Their distinction, however, has not always been consensual (Hayden, 1980).
+In fact, the definition of scaled piece has suffered significant changes ever since its initial identification in the early 20th century and, to this day, there still seems to be no clear definition for this type of artifact.
+The first definition of scaled pieces was proposed by Baron and Bouysonnie (1906), describing them as a result of bipolar knapping through direct percussion, with the “core” rested in a hard surface, originating splintering in both ends of the tool.
+Since then, several other definitions were adopted and adapted by researchers for a very diverse set of contexts across the world (see e.g.
+Octobon, 1938; Sonneville-Bordes and Perot, 1956: 552; MacDonald; 1968:88; Hayden, 1980; Shott, 1989; Knight 1991:61).
+
+Beyond the classification debate, the biggest issue with this type of tools comes from a functional standpoint, a problem that has also been quite debated (see e.g., Tixier, 1963; Binford and Quimby, 1963; Semenov, 1964; Flood, 1980:308; LeBlanc, 1992; Shott, 1999; Lucas and Hays, 2004; de la Peña, 2011; Igreja & Porraz, 2013).
+The issue surrounding this problem lies within the functional equifinality of these artifacts.
+Contrary to bipolar cores that are most of the times easily identified as cores, scaled pieces have been associated with two distinct types of activities: (1) as intermediate pieces or wedges for working hard raw materials (bone, ivory, antler, etc.); and (2) as cores for the extraction of chips and small bladelets (see e.g.
+Brantingham et al., 2004).
+Although largely debated, the ambiguity of this classification has, however, been ignored in some of the most recent literature, with some authors not acknowledging that bipolar evidence in stone tools might also result from other activities other than only reduction strategies (see e.g.
+Hiscock, 2015).
+
+Most current studies on scaled pieces apply use-wear methods (e.g.
+Vaughan, 2002; Lucas & Hays, 2004; Bosinski, 2007; Gibaja et al.
+2007; Sano, 2012; Igreja & Porraz, 2013, de la Peña, 2011; 2014; 2015a; 2015b, Bader et al., 2015), focusing on the identification of polishes and use-wear patterned stigmas, through both micro and macroscopic analysis of splintered surfaces and its comparison with experimental assemblages.
+Frequently, these studies coincide in interpreting scaled pieces as intermediate elements for the work of hard organic raw materials.
+The study by P.
+de la Peña (2011) is one of the most recent and relevant references in this regard.
+The author presents the results of an experimental program aiming to identify specific wear patterns in the use of bipolar techniques that allow the distinction between different types of activities/worked materials.
+Results indicate that while no visible differences can be identified in the percussion area, significant variation can be observed in the morphology of the areas in contact with the worked material.
+
+As in many other regions and stone age periods across the world (e.g., Diez et.
+al 2009; Igreja & Porraz, 2013; Langjans, 2012; Leakey, 1971: 221; Leblanc, 1992; Shott, 1989, 1999; White, 1968), evidence of bipolar technologies are quite ubiquitous in European Upper Paleolithic contexts.
+In many sites, bipolar elements are classified as scaled pieces, a morphotype that has been putatively considered as one the most common lithic tools in Upper Paleolithic assemblages (de la Peña, 2015).
+Different functional connotations for this type of artifacts have been presented in several western european Upper Paleolithic studies (e.g., Zilhão, 1997; Aubry et al.
+1998; de la Peña 2011; Sano, 2012).
+
+In the case of the westernmost regions of Iberia, scaled pieces are commonly found in archaeological contexts ranging from the Upper Paleolithic to the Neolithic (e.g.
+Aubry et al.
+1998; Bicho 1992; Carvalho 1998; Zilhão 1997).
+While scarce, the majority of Portuguese paleolithic studies (e.g.
+Almeida 2000, Gameiro 2012) do not offer analytical interpretations for the presence of scaled pieces, simply using references that interpret them as bipolar cores for the extraction of small bladelets and chips to be supposedly used in composite tools (as argued by Zilhão, 1997) or as intermediate pieces for working hard materials (Gibaja et al., 2007; Marreiros, 2009).
+Zilhão (1997) for example argues that the presence of scaled pieces throughout the Upper Paleolithic sequence of central Portugal is inversely proportional to the presence of “carinated cores”, and thus it is likely that the former should have worked as a flexible substitute for the latter.
+Carvalho (1998) agrees with this interpretation and considers it also valid for the Portuguese Early Neolithic.
+
+While we agree that some of the elements might have been used as cores, these interpretations seem overly simplistic, and the authors seem to have little to no evidence to back their claims other than the inverse relationship mentioned above.
+This results in part from the lack of comprehensive studies regarding this type of artifacts at a regional scale.
+Additionally, the contexts from where most of these artifacts were recovered did not have good organic preservation and, thus, no direct association between these tools and the exploitation of hard-organic raw materials is possible, and, in face of the inexistence of any absolute dates, their precise chronological attribution is also, most of the times, unreliable.
+
+In this paper, we present a first approach to the characterization of the morpho-technological variability and consequent role of scaled pieces during the Upper Paleolithic of the westernmost regions of Iberia (c.
+32 000 – 10 000 cal BP).
+Using data from the multi-component, thoroughly dated, site of Vale Boi, located in southern Portugal, we present the analysis of a series of technological and morpho-functional attributes of a relatively large set of scaled pieces coming from one of the areas of the site.
+Using these data as a starting point, we then explore the relationship between the variability detected in the production and use of these artifacts with inter-site lithic technological patterns, and with the striking evidence for an intensification and diversification of faunal resources exploitation during the time-span under consideration.
+
 \newline
 
-```{r }
+# Vale Boi
+
+The archaeological site of Vale Boi is located in the western edge of the Algarve region (southern Portugal) (Figure \@ref(fig:VBLocation)).
+The site can be found in a small valley following a river north-south for 2 km until it reaches the Atlantic Ocean.
+Archaeological deposits occupy an estimated area of over 10 000 sq.
+meters across a stepped slope marked at the top by a 10 meter-high limestone cliff face.
+
+A complete sequence of Upper Paleolithic has been identified at Vale Boi, with all the traditionally-defined techno-complexes (Gravettian, Proto-Solutrean, Solutrean and Magdalenian) being identified across the three main excavation areas: the Terrace, the Rockshelter, and the Slope.
+
+The Terrace area is located in the lower part of the hill.
+In this area, the longest archaeological sequence of the site can be found, including the complete Upper Paleolithic sequence but also three Holocene horizons, corresponding to Neolithic, Mesolithic and Epipaleolithic occupations (Bicho et al., 2012; Cascalheira et.
+al, 2017).
+From within the lower levels of this area, an Early Gravettian occupation was discovered, dated to c.
+32 ka cal BP, being one of the earliest radiocarbon dates for anatomically modern humans in Southern Iberia (Bicho et al., 2016; Marreiros et al., 2015).
+
+The Rockshelter area is a collapsed rock shelter located in the upper part of the slope, a couple of meters below the limestone cliff.
+This collapse would have occurred after the Last Glacial Maximum, since below the collapsed debris, several Solutrean occupations can be found, overlaying a sequence of very ephemeral Gravettian horizons (Marreiros, 2009; Cascalheira, 2009; Manne et al., 2012).
+The Solutrean is dated to between c.
+20 ka and 25 ka cal BP (Cascalheira et al., 2012), while the Gravettian is dated between 26 ka and 32 ka cal BP (Marreiros et al., 2015).
+
+Finally, the Slope section, from where the assemblage here presented is coming from, is composed of a series of excavation areas opened across the mid-hill sector of the site.
+These areas exhibit heterogeneous conditions in terms of site formation processes and archaeological preservation (Manne et al., 2012), but all revealed the presence of occupations attributed to the Gravettian, Proto-Solutrean, Solutrean and Magdalenian.
+Like in the previous areas, remains are well preserved, and high frequencies of lithic artifacts, malacological and mammalogical fauna and bone tools were recovered.
+No habitation features were identified in this area and based on the conditions and type of artifacts found, it has been suggested that this area would have functioned as a midden deposit (Bicho et al., 2010, 2012a).
+
+```{r VBLocation, fig.cap=""}
+
+Map()
+
+knitr::include_graphics("../figures/VBMap.png")
 
@@
 
-# Materials and Methods
+## Lithic technology
+
+Vale Boi’s lithic studies have revealed a general tendency for stable technological and functional patterns throughout the Upper Paleolithic (see e.g.
+Bicho et al., 2012; Gibaja and Bicho, 2006; Bradtmöller et.
+al, 2015; Marreiros, 2009, 2013, 2015; Cascalheira, 2009, 2013).
+This can be explained, partially, by raw material availability.
+Most raw materials were procured locally or regionally from deposits located at no more than 20 km away from the site (Bicho et al., 2010a).
+Chert was the most used rock type for more complex retouched tools production, while quartz and greywacke were mostly used for flake extraction and simple retouched tools production.
+Other raw materials can be found at the site, but much more restricted, both diachronically and functionally within each techno-complex and each archaeological horizon.
+Schist, for instance, shows up in some occupations almost exclusively connected to mobile art.
+Dolerite can only be found in the Proto-Solutrean levels, while chalcedony is mostly limited to the Proto-Solutrean and Solutrean levels.
+
+Chert is the most abundant knapped rock type in Vale Boi.
+Throughout the several occupations, chert was considered the best rock type for knapping, which is evident by the way it was explored with much more complex strategies than either greywacke or quartz.
+While it can be said that chert exploration strategies were more elaborate, with exception of the bifacial technology during the Solutrean, they were still simple unidirectional reduction sequences producing mostly small debitage.
+Elongated products are found at very low frequencies across all chronologies.
+When present, retouched tools are mostly notches, denticulates, end-scrapers and, if considered, scaled pieces.
+One possible reason for these simplistic strategies seems to be an overall low knapping quality, since nodules are quite small, and frequently show a high degree of tectonically derived fractures (Pereira et al.
+2016).
+
+Two distinct types of quartz were identified at the site.
+The first is a thick grain, low-quality type, mostly inadequate for knapping.
+Still, this kind of quartz is present in large quantities and is most likely associated with stone boiling and grease rendering activities (Manne & Bicho, 2009; Manne, 2010; Bicho et al, 2012; Manne et al.
+2012, 2014).
+The second type of quartz is characterized by finer-grain small pebbles with yellowish cortex.
+This type of quartz was knapped using rather simple strategies, mostly for small flake production, which in turn were used to produce simple and versatile tools (e.g.
+scaled pieces).
+
+Greywacke was used for the rare production of flakes and choppers but shows up in the site mostly in the form of large slabs in which the identification of concavely shaped impact marks is thought to result from their use as anvils (Manne et.
+al, 2012, Manne & Bicho, 2009).
+The importance of greywacke anvils is indicated by its high frequency, with hundreds of slabs found in many levels throughout all occupations (Bicho et al, 2012; Manne et al.
+2012).
+Greywacke was also knapped for flake extraction, using expedient unidirectional reduction strategies.
+
+\newline
+
+## Subsistence patterns
+
+Organic preservation is good at Vale Boi.
+Faunal remains can be frequently found in every occupation, from both terrestrial and marine sources.
+Marine fauna at the site is marked by the presence of mollusks, crustaceans, some fish vertebrae and, in rare occasions, marine mammal remains (Manne, 2010).
+Regarding terrestrial fauna, three species dominate the vertebrate group: rabbit (Oryctolagus cuniculus), red deer (Cervus elaphus) and horse (Equus caballus).
+Smaller amounts of wild ass (Equus hydruntinus), aurochs (Bos primigenius), ibex (Capra pyrenaica) and wild boar (Sus scrofa) are also present (Manne, 2010, 2012, 2014).
+One of the most interesting patterns within the faunal assemblages is that a large portion of ungulate remains present specific types of fracture that have been associated with bone marrow extraction activities.
+Red deer and horse remains frequently show evidence of opposed cone fractures, trituration and smashing (Manne & Bicho, 2009; Manne, 2010, 2012, 2014).
+
+Ungulates would have been hunted and processed in a similar fashion throughout the Upper Paleolithic.
+While there were conditions for the whole bones to be preserved, these bones are frequently anthropically broken.
+These fragmentation patterns seem to be linked to grease rendering activities (Manne, 2010, 2012, 2014).
+The main goal of grease rendering is to obtain grease, through heat exposure from animal bones, as it has a very high caloric value.
+Other than this, this grease could have many uses with the addition of being easily stored and transported (Manne, 2012).
+The spongeous bone parts would be fragmented and deposited in a hole, covered with animal pelts full of water, after which, pre-heated rock fragments were added.
+The high temperature of the rocks would make the water boil and therefore separate the grease from the spongeous bones creating a highly nutritional stew (Manne, 2014).
+After being cooled the grease would accumulate at the top where it could be easily removed, transported and stored.
+Unlike simple bone marrow extraction, this method involved large preparation, including water transport, fire, and heat production, rock heating, and finally the storing of the grease (Manne, 2014).
+
+These grease rendering techniques are thought to be quite common at the site since the parts of the bones with higher amounts of fat are missing despite the good preservation of the rest of the remains (Manne et al.
+2012).
+Fragmented ungulate remains show up in the site connected to large amounts of thermo-altered quartz, anvils, hard hammers and scaled pieces.
+This suggests that red deer and horse bones were processed and afterward intensively grease rendered (Manne & Bicho, 2009; Manne, 2010, 2012, 2014).
+To confirm this, a single scaled piece was found stuck to a cracked phalanx in a Gravettian horizon (Manne, 2012).
+
+\newline
+
+# Methods
+
+## Scaled pieces attribute analysis
+
+Attribute analysis was split into two main groups, each corresponding to two distinct types of features: (1) technological attributes and (2) morpho-functional attributes.
+
+In the first group, a series of variables traditionally used in lithic studies (e.g.
+Andrefsky, 2005; Inizan et.
+al, 1999) were recorded (Table \@ref(tab:tech_attributes)), aiming to characterize patterns of blank choice for the application of bipolar technology.
+
+For the second group of variables a macroscopic approach, building upon the work of de la Peña (2011) and Fischer et.
+al (1984) was adopted (Table \@ref(tab:morph_attributes)).
+Studies by Vaughan (2002), Sano (2012), and Gibaja et al.
+(2007) indicate that since use-wear traces are formed before the splintering, the removal of small chips removes most of the polishes and use-wear traces left by the contact with the static element (either a stone anvil or a hard-organic materials).
+This prevented us to pursue a microscopic use-wear approach in this study.
+
+\newline
+
+```{r tech_attributes}
+
+```
+
+\newline
+
+```{r morph_attributes}
+
+```
+
+Macroscopic morpho-functional attributes were separately analyzed for each damaged platform, aiming to detect patterns of morphological change that occurred in artifacts during use.
+Following the work of de la Peña (2011, 2015) and Gonzalez-Urquijo & Ibanez-Estévez (1994), we expected that these attributes would be indicative of which function the artifacts had.
+For instance, according to de la Peña (2011, 2015), for pieces used as wedges, the shape of the passive platform remains the same and the other varies according to the type of material worked on the active platform.
+Furthermore, these pieces would have irregular shapes, variable scar size, and irregular scar distribution.
+On the other hand, pieces used as bipolar cores would have squared or rectangular shapes, symmetric straight platforms, and a higher frequency of scars on the striking edge than on the active edge.
+The addition of other attributes expanding on the work by Gonzalez-Urquijo & Ibanez-Estévez (1994) was made following the same reasoning.
+We expected, thus, to be able to differentiate pieces used as wedges from pieces used as cores, as both groups would show distinct combinations of attributes.
+To assist us with this differentiation we tested the presence of the referred patterns within our assemblage using descriptive and multivariate statistical analysis and comparing it with the data described by the referred authors.
+
+\newline
 
 # Results
+
+A total of 139 scaled pieces were analyzed, of which 42.4% come from Gravettian, and 45.3% from Solutrean levels, as shown in (Table(@ref(tab:assemblage)).The assemblage was divided into 4 main chronological groups: Magdalenian, Solutrean, Proto-Solutrean, and Gravettian.
+
+As previously mentioned, as far as the technological data is concerned, one of our main objectives was to characterize and understand the choice of blanks involved in bipolar technology.
+We did not include the metric data, as most of the data from the original blank are mostly gone due to reduction and use.
+Overall, the technological analysis revealed some trends that lasted throughout the Upper Paleolithic, in agreement with the general patterns of lithic technology in the site (see above).
+
+\newline
+
+```{r assemblage, echo = FALSE}
+
+var_list <- c("Chronology")
+
+assemb <- cross_tb(dataset, "RawMaterial", var_list)
+
+assemb <- assemb %>%
+  as.tibble() %>% 
+  dplyr::select(" " = Variable, Chert, Quartz, Chalcedony, Total = Overall)
+
+knitr::kable(assemb, caption= "")
+
+```
+
+\newline
+
+Across all techno-complexes, blanks used were either flakes or unclassifiable fragments, but the choice seems to be different for each raw material.
+For quartz, in the Gravettian assemblage, the blank types are almost equally split between flakes and unclassifiable fragments (Table(@ref(tab:technogravettian)).
+In other periods, flakes were the preferred type of blank (Tables(@ref(tab:technosolutrean)(@ref(tab:technoproto)(@ref(tab:technomagdalenian)).
+Regarding chert, in every occupation flakes dominate the assemblages, followed by a reduced number (4) of unclassifiable fragments.
+Chalcedony blanks were exclusively flakes.
+
+\newline
+
+```{r technology, echo=FALSE}
+
+# Pull dataset
+techno_data <- read_data()
+
+# Set variables
+var_list <- c("Blank","CrossSection","Profile","Blankshape","X.Cortex","ButtType","Profile")
+
+# Filter by chronology
+filter <- split(techno_data, techno_data$Chronology)
+
+# Get cross-tables by chronology using cross_tb() function
+Gravettian <- cross_tb(filter$Gravettian, "RawMaterial", var_list) 
+Gravettian <- select(Gravettian, " ", Chert, Quartz, Total)
+
+Solutrean <- cross_tb(filter$Solutrean, "RawMaterial", var_list)
+
+`Proto-Solutrean` <- cross_tb(filter$`Proto-Solutrean`, "RawMaterial", var_list)
+
+Magdalenian <- cross_tb(filter$Magdalenian, "RawMaterial", var_list)
+Magdanelian <- select(Magdalenian, " ", Chert, Quartz, Total)
+
+```
+
+\newline
+
+```{r technogravettian, echo=FALSE}
+
+knitr::kable(Gravettian, caption= "")
+
+```
+
+\newline
+
+```{r technoproto, echo=FALSE}
+
+knitr::kable(`Proto-Solutrean`, caption= "")
+
+```
+
+\newline
+
+```{r technoSolutrean, echo=FALSE}
+
+knitr::kable(Solutrean, caption= "")
+
+```
+
+\newline
+
+```{r technomagdalenian, echo=FALSE}
+
+knitr::kable(Magdalenian, caption= "")
+
+```
+
+\newline
+
+Technological and morphological characteristics of the flake blanks present very similar patterns across time and amongst raw materials.
+The blanks sought after would have the following characteristics: straight profile, parallel edges, no-cortex and trapezoidal or triangular shaped transversal sections.
+
+Other characteristics of the assemblage are the small frequency of retouched pieces, fire alterations, and of the original striking platforms.
+Still, when present, retouch was located in the lateral part of the artifacts, similar to what would define a side-scraper.
+In a very small number of cases, striking platforms were detected, mostly with a flat morphology.
+The absence of the original blank platforms is to be expected in this type of artifact, mostly due to the functional use of the pieces, and consequent removal of the platform, rather than an actual choice.
+
+When identifiable, the technological and typological axes tend to coincide, although in many cases the technological axis could not be identified, especially in quartz pieces.
+The longitudinal section shows a large variability of shapes, independent of raw material or chronology.
+Similarly, the dorsal pattern of previous removals was rarely identified, although this, like with the case of the butt absence, might occur due to the functional stigmas.
+
+The patterns of use and rotation of damaged platforms seem to be the same across all chronologies.
+In every assemblage quartz pieces were exclusively used in one single axis, exhibiting only two damaged platforms.
+On the other end, some (3,38%) chert and chalcedony artifacts exhibit multiple functional axis, with three or four damaged platforms.
+This seems to indicate different strategies of curation for coarse and fine grain raw materials.
+
+In scaled pieces, metrics are a difficult set of variables to clearly show as results.
+These attributes are impacted by both the initial blank size as well as by the functional use and reduction.
+The pieces showed lengths ranging from 10 mm to 45 mm, widths ranging from 8 mm to 35 mm and thicknesses ranging from 3 mm to 20 mm.
+In general, quartz scaled pieces tend to be larger and thicker than chert and chalcedony pieces.
+On the other hand, chert pieces tend to be wider than the rest.
+
+```{r meanplots, echo = FALSE, fig.cap=""}
+
+length <- mean_plot(techno_data, "Chronology", "AxisLength", "RawMaterial")
+width <- mean_plot(techno_data, "Chronology", "AxisWidth", "RawMaterial")
+thickness <- mean_plot(techno_data, "Chronology", "Thickness", "RawMaterial")
+
+meanplots <- ggarrange(length, width, thickness, ncol = 1, nrow = 3, labels = c("A", "B", "C"), common.legend = TRUE, legend = "right")
+
+meanplots
+
+```
 

### 2018-05-06 R/scaled_pieces_functions.R

@@ -1,0 +1,112 @@
+
+###########################################################################
+# READ DATA ----------------------------------------------------------------
+
+read_data <- function(){
+  tech_data <- read.csv("../data/raw_data/technology_dataset.csv")
+
+}
+
+
+###########################################################################
+# MAPS --------------------------------------------------------------------
+
+require(maptools)
+require(ggplot2)
+require(ggmap)
+require(devtools)
+require(legendMap)
+
+Map <- function(){
+
+  VBLocation <- c(lon = -8.808621, lat = 37.089902)
+
+  VBMap <- get_googlemap(center = c(lon = -8.101284, lat = 39.486585),
+                         zoom = 7,
+                         scale = 2,
+                         maptype = "roadmap")
+
+  VBMap <-
+    ggmap(VBMap) +
+    geom_point(data = data.frame(t(VBLocation)),
+               aes(lon,
+                   lat),
+               size = 5,
+               colour = "black") +
+    geom_text(data = data.frame(t(VBLocation)), aes(x = lon, y = lat, label = "Vale Boi"),
+              size = 4, colour = "black", vjust = -1, hjust = 0.5)
+
+
+  VBMap <- VBMap +
+    legendMap::scale_bar(lon = -11.3,
+                         lat = 37,
+                         distance_lon = 50,
+                         distance_lat = 8,
+                         distance_legend = 20,
+                         legend_colour = "white",
+                         dist_unit = "km",
+                         orientation = TRUE,
+                         arrow_length = 50,
+                         arrow_distance = 50,
+                         arrow_north_size = 3)
+
+  ggsave("VBMap.png")
+
+}
+
+
+###########################################################################
+# CROSS TABLES ------------------------------------------------------------
+
+
+
+cross_tb <- function(dataset,x,y) {
+
+require(tab)
+require(tidyverse)
+
+  ct <- tabmulti(dataset, x, y,
+                  cell = "n",
+                  parenth = "col.percent",
+                  p.include = FALSE,
+                  n.headings = FALSE,
+                  freq.text.label = "none")
+
+  ct <- ct %>%
+    as.tibble() %>%
+    dplyr::select(" " = Variable, Chert, Quartz, Chalcedony, Total = Overall)
+
+}
+
+
+mean_tb <- function(dataset,x,y) {
+
+  require(tab)
+  require(tidyverse)
+
[…]

### 2018-05-08 analysis/paper/paper.Rmd
modifications in stat tests for cross-tables

@@
       reference_docx: "../templates/template2.docx" 
-bibliography: ScaledPiecesVB.bib
+bibliography: ShortTermOccupations.bib
 csl: "../templates/springer-humanities-author-date.csl"
@@
   fig.path = "../figures/")
+
+source('~/R_DATA/compendiums/ScaledPiecesVB/R/scaled_pieces_functions.R')
 
@@
 library(tab)
+
+all_data <- read_data()
+techno_data <- all_data$techno_data
+morpho_data <- all_data$morpho_data
 
@@
 
-```{r VBLocation, fig.cap=""}
+```{r VBLocation, echo = FALSE, fig.cap="Location of the site of Vale Boi"}
+
+source('~/R_DATA/compendiums/ScaledPiecesVB/R/scaled_pieces_functions.R')
 
@@
 
-```{r tech_attributes}
+```{r tech_attributes, echo = FALSE}
 
@@
 
-```{r morph_attributes}
+```{r morph_attributes, echo = FALSE}
 
 ```
+
+\newline
 
@@
 
+## Reproducibility and open source materials
+
+All analyses and data processing were accomplished in R (version `r getRversion()`) [@r_core_team_r:_2013].
+Following recent concerns on the reproducibility of archaeological analysis we include the entire R code used for all the analysis and visualizations contained in this paper in our supplemental online material (SOM) at https://dx.doi.org/… To produce those files we followed the procedures described by Marwick et al.
+[-@marwick_packaging_2017] for the creation of research compendiums to enhance the reproducibility of research.
+The files provided contain all the raw data used in our analysis as well as a custom R package [@wickham_r_2015] holding the code use for all analysis and to produce all tables and figures.
+To enable maximum re-use, our code is released under the MIT license, our data as CC-0, and our figures as CC-BY, [for more information see @marwick_computational_2016].
+
+\newline
+
 # Results
 
-A total of 139 scaled pieces were analyzed, of which 42.4% come from Gravettian, and 45.3% from Solutrean levels, as shown in (Table(@ref(tab:assemblage)).The assemblage was divided into 4 main chronological groups: Magdalenian, Solutrean, Proto-Solutrean, and Gravettian.
+A total of `r summarise(techno_data, n())` scaled pieces were analyzed, of which `r round(techno_data %>% filter(Chronology == "Gravettian") %>% summarise(n = n()) %>% mutate(n = (n / 139)*100), digits = 2)`% come from Gravettian, and `r round(techno_data %>% filter(Chronology == "Solutrean") %>% summarise(n = n()) %>% mutate(n = (n / 139)*100), digits = 2)`% from Solutrean levels, as shown in (Table \@ref(tab:assemblage)).The assemblage was divided into 4 main chronological groups: Magdalenian, Solutrean, Proto-Solutrean, and Gravettian.
 
@@
 
-assemb <- cross_tb(dataset, "RawMaterial", var_list)
-
-assemb <- assemb %>%
-  as.tibble() %>% 
-  dplyr::select(" " = Variable, Chert, Quartz, Chalcedony, Total = Overall)
+assemb <- cross_tb(techno_data, "RawMaterial", var_list)
 
@@
 Across all techno-complexes, blanks used were either flakes or unclassifiable fragments, but the choice seems to be different for each raw material.
-For quartz, in the Gravettian assemblage, the blank types are almost equally split between flakes and unclassifiable fragments (Table(@ref(tab:technogravettian)).
-In other periods, flakes were the preferred type of blank (Tables(@ref(tab:technosolutrean)(@ref(tab:technoproto)(@ref(tab:technomagdalenian)).
-Regarding chert, in every occupation flakes dominate the assemblages, followed by a reduced number (4) of unclassifiable fragments.
-Chalcedony blanks were exclusively flakes.
+For quartz, in the Gravettian assemblage, the blank types are almost equally split between flakes and unclassifiable pieces (Table \@ref(tab:technogravettian)).
+In other periods, flakes were the preferred type of blank (Tables \@ref(tab:technosolutrean), \@ref(tab:technoproto) and \@ref(tab:technomagdalenian)).
+Regarding chert, in every occupation flakes dominate the assemblages, followed by a reduced number (n = `r techno_data %>% filter(RawMaterial == "Chert" & Blank == "Non_identifiable") %>% summarise(n())`) of unclassifiable blanks.
+Chalcedony blanks are exclusively flakes.
 
@@
 
-```{r technology, echo=FALSE}
-
-# Pull dataset
-techno_data <- read_data()
+```{r technology, include = FALSE, echo=FALSE}
 
 # Set variables
-var_list <- c("Blank","CrossSection","Profile","Blankshape","X.Cortex","ButtType","Profile")
+var_list <- c("Blank","CrossSection","Profile","Blankshape","Cortex","ButtType","Profile")
 
@@
 Gravettian <- cross_tb(filter$Gravettian, "RawMaterial", var_list) 
-Gravettian <- select(Gravettian, " ", Chert, Quartz, Total)
+Gravettian <- select(Gravettian, " ", Chert, Quartz, Total) #to remove chalcedony
 
@@
 Magdalenian <- cross_tb(filter$Magdalenian, "RawMaterial", var_list)
-Magdanelian <- select(Magdalenian, " ", Chert, Quartz, Total)
+Magdalenian <- select(Magdalenian, " ", Chert, Quartz, Total) #to remove chalcedony
 
@@
 
-knitr::kable(Gravettian, caption= "")
+knitr::kable(Gravettian, caption= "Technological attribute frequencies by raw materials for the Gravettian sample")
 
@@
 
-knitr::kable(`Proto-Solutrean`, caption= "")
+knitr::kable(`Proto-Solutrean`, caption= "Technological attribute frequencies by raw materials for the Proto-Solutrean sample")
 
@@
 
-```{r technoSolutrean, echo=FALSE}
+```{r technosolutrean, echo=FALSE}
 
-knitr::kable(Solutrean, caption= "")
+knitr::kable(Solutrean, caption= "Technological attribute frequencies by raw materials for the Solutrean sample")
 
@@
 
-knitr::kable(Magdalenian, caption= "")
+knitr::kable(Magdalenian, caption= "Technological attribute frequencies by raw materials for the Magdalenian sample")
 
@@
 Technological and morphological characteristics of the flake blanks present very similar patterns across time and amongst raw materials.
-The blanks sought after would have the following characteristics: straight profile, parallel edges, no-cortex and trapezoidal or triangular shaped transversal sections.
+The blanks sought after would have the following characteristics: straight profile, parallel edges, no cortical surfacem and trapezoidal or triangular shaped cross-sections.
 
-Other characteristics of the assemblage are the small frequency of retouched pieces, fire alterations, and of the original striking platforms.
+Other characteristics of the assemblage are the low frequency of retouch (n = `r techno_data %>% filter(RetouchedBlank == "Yes") %>% summarise(n())`), fire alterations (n = `r techno_data %>% filter(Fire != "No_Traces") %>% summarise(n())`), and of the original striking platforms (n = `r techno_data %>% filter(ButtType != "Absent") %>% summarise(n())`).
 Still, when present, retouch was located in the lateral part of the artifacts, similar to what would define a side-scraper.
-In a very small number of cases, striking platforms were detected, mostly with a flat morphology.
+In a very small number of cases when striking platforms were detected, these were mostly flat.
 The absence of the original blank platforms is to be expected in this type of artifact, mostly due to the functional use of the pieces, and consequent removal of the platform, rather than an actual choice.
-
-When identifiable, the technological and typological axes tend to coincide, although in many cases the technological axis could not be identified, especially in quartz pieces.
+The large absence of platforms may, on the other hand, be the result of the use of the technological axis as the main functional axis.
+In fact, when identifiable, technological and typological axes coincide in `r inline_perc(techno_data, AxisCoincident, "Yes")`% of the cases.
 The longitudinal section shows a large variability of shapes, independent of raw material or chronology.
-Similarly, the dorsal pattern of previous removals was rarely identified, although this, like with the case of the butt absence, might occur due to the functional stigmas.
+Similarly, the dorsal pattern of previous removals was rarely identified, although this, like with the case of the platform absence, might occur due to the functional stigmas and be dependent on the the intensity of use for each artifact.
 
@@
 In every assemblage quartz pieces were exclusively used in one single axis, exhibiting only two damaged platforms.
-On the other end, some (3,38%) chert and chalcedony artifacts exhibit multiple functional axis, with three or four damaged platforms.
+On the other end, some (3,38%) chert and chalcedony artifacts exhibit multiple functional axis, with three or four damaged platforms (Figure \@ref(fig:nplatforms)).
 This seems to indicate different strategies of curation for coarse and fine grain raw materials.
+
+```{r nplatforms, fig.cap=""}
+
+n_axis <- morpho_data %>% 
+  group_by(Chronology,RawMaterial,DamagedPlatforms) %>% 
+  summarise(count=n()) %>% 
+  mutate(perc=100*(count/sum(count)))
+n_axis
+
+ggplot(n_axis, aes(x = factor(RawMaterial), y = perc, fill = factor(DamagedPlatforms))) +
+  geom_bar(stat="identity", width = 0.7) +
+  labs(x = "RawMaterial", y = "%", fill = "Number of Damaged Platforms") +
+  facet_wrap(~Chronology) +
+  theme_bw() +
+  theme(strip.text.x = element_text(size = 14), text = element_text(size=14))
+
+```
 
@@
 These attributes are impacted by both the initial blank size as well as by the functional use and reduction.
-The pieces showed lengths ranging from 10 mm to 45 mm, widths ranging from 8 mm to 35 mm and thicknesses ranging from 3 mm to 20 mm.
-In general, quartz scaled pieces tend to be larger and thicker than chert and chalcedony pieces.
-On the other hand, chert pieces tend to be wider than the rest.
+Consideranging the typological axes (see Table \@ref(tab:tech_attributes)) for definition) length ranging from `r min(techno_data$AxisLength, na.rm = TRUE)` mm to `r max(techno_data$AxisLength, na.rm = TRUE)` mm, width ranging from `r min(techno_data$AxisWidth, na.rm = TRUE)` mm to `r max(techno_data$AxisWidth, na.rm = TRUE)` mm and thickness ranging from `r min(techno_data$Thickness, na.rm = TRUE)` mm to `r max(techno_data$Thickness, na.rm = TRUE)` mm.
+In general, mean values tend to be similiar between raw materials, with some differences occuring within the Proto-Solutrean and Magdalenian assemblages, most certainly as a result of the small samples analysed for each of these periods (Figure \@ref(fig:meanplots)).
 
-```{r meanplots, echo = FALSE, fig.cap=""}
+```{r meanplots, echo = FALSE, fig.cap="Barplots of means for Length, Width and Thickness of scaled pieces by raw material and across the four chronological phases.
+Error bars represent standard deviations."}
 

### 2018-05-08 R/scaled_pieces_functions.R

@@ -3,7 +3,13 @@
 # READ DATA ----------------------------------------------------------------
 
 read_data <- function(){
-  tech_data <- read.csv("../data/raw_data/technology_dataset.csv")
+  #tech_attributes <- read.csv("../data/raw_data/technology_attributes.csv")
+  #morpho_attributes <- read.csv("../data/raw_data/morpho_attributes.csv")
+  techno_data <- read.csv("../data/raw_data/technology_dataset.csv")
+  morpho_data <- read.csv("../data/raw_data/morpho_dataset.csv")
+
+  return(list(morpho_data = morpho_data,
+              techno_data = techno_data))
 
 }
 
@@ -11,13 +17,13 @@ read_data <- function(){
 ###########################################################################
 # MAPS --------------------------------------------------------------------
 
-require(maptools)
-require(ggplot2)
-require(ggmap)
-require(devtools)
-require(legendMap)
-
 Map <- function(){
+
+  require(maptools)
+  require(ggplot2)
+  require(ggmap)
+  require(devtools)
+  require(legendMap)
 
   VBLocation <- c(lon = -8.808621, lat = 37.089902)
 
@@ -50,7 +56,7 @@ Map <- function(){
                          arrow_distance = 50,
                          arrow_north_size = 3)
 
-  ggsave("VBMap.png")
+  ggsave("../figures/VBMap.png")
 
 }
 
@@ -79,6 +85,10 @@ require(tidyverse)
 }
 
 
+###########################################################################
+# MEAN TABLE --------------------------------------------------------------
+
+
 mean_tb <- function(dataset,x,y) {
 
   require(tab)
@@ -99,10 +109,16 @@ mean_tb <- function(dataset,x,y) {
 }
 
 
+
+###########################################################################
+# MEAN PLOTS --------------------------------------------------------------
+
 mean_plot <- function(dataset, x, y, z){
 
+require(ggpubr)
+
   ggbarplot(techno_data, x = x, y = y,
-            add = c("mean_se", "jitter"), size = 1,
+            add = c("mean_sd", "jitter"), size = 1,
             color = z, palette = c("#00AFBB", "#E7B800", "#FC4E07"),
             position = position_dodge(0.8)) +
   theme_gray()
@@ -110,3 +126,21 @@ mean_plot <- function(dataset, x, y, z){
 }
 
 
+
+
+
[…]

### 2018-05-11 analysis/paper/paper.Rmd
function to recode data

@@
 As previously mentioned, as far as the technological data is concerned, one of our main objectives was to characterize and understand the choice of blanks involved in bipolar technology.
-We did not include the metric data, as most of the data from the original blank are mostly gone due to reduction and use.
+We did not include the metric data, as most of the data from the original blanks are mostly gone due to reduction and use.
 Overall, the technological analysis revealed some trends that lasted throughout the Upper Paleolithic, in agreement with the general patterns of lithic technology in the site (see above).
@@
 
+\newline
+
 ```{r nplatforms, fig.cap=""}
@@
 
-In scaled pieces, metrics are a difficult set of variables to clearly show as results.
-These attributes are impacted by both the initial blank size as well as by the functional use and reduction.
-Consideranging the typological axes (see Table \@ref(tab:tech_attributes)) for definition) length ranging from `r min(techno_data$AxisLength, na.rm = TRUE)` mm to `r max(techno_data$AxisLength, na.rm = TRUE)` mm, width ranging from `r min(techno_data$AxisWidth, na.rm = TRUE)` mm to `r max(techno_data$AxisWidth, na.rm = TRUE)` mm and thickness ranging from `r min(techno_data$Thickness, na.rm = TRUE)` mm to `r max(techno_data$Thickness, na.rm = TRUE)` mm.
+\newline
+
+As in other classes of tools, scaled pieces' metric attributes are impacted by both the initial blank size as well as by the intensity of their use.
+Further, because some pieces were used in multiple axes, direct comparisons for length and width of the typological axes (see Table \@ref(tab:tech_attributes)) for definition) cannot be straighforwardly performed.
+Since the majority of the analyzed pieces presents a rectangular edge delineation, we use an approximation of the area (i.e.
+Length x Width), and thickness, to characterize size variability across chronologies and raw materials.
+
+Area calculations revealed a maximum size of `r round(max(techno_data$Area, na.rm = TRUE), digits = 2)` mm and a minimum value of `r round(max(techno_data$Area, na.rm = TRUE), digits = 2)` mm.
+For thickness the maximum is `r round(max(techno_data$Thickness, na.rm = TRUE), digits = 2)` mm and the minimum is `r round(min(techno_data$Thickness, na.rm = TRUE), digits = 2)` mm.
 In general, mean values tend to be similiar between raw materials, with some differences occuring within the Proto-Solutrean and Magdalenian assemblages, most certainly as a result of the small samples analysed for each of these periods (Figure \@ref(fig:meanplots)).
+Across techno-complexes, however, no significant statistical differences were detected when using an ANOVA test for both Area (`r anova_apa(aov(Area~Chronology, techno_data), "Chronology")` and  Thickness (`r anova_apa(aov(Thickness~Chronology, techno_data), "Chronology")`).
 
-```{r meanplots, echo = FALSE, fig.cap="Barplots of means for Length, Width and Thickness of scaled pieces by raw material and across the four chronological phases.
+\newline
+
+```{r meanplots, echo = FALSE, fig.cap="Barplots of means for Area (Length x Width) and Thickness of scaled pieces by raw material and across the four chronological phases.
 Error bars represent standard deviations."}
 
-length <- mean_plot(techno_data, "Chronology", "AxisLength", "RawMaterial")
-width <- mean_plot(techno_data, "Chronology", "AxisWidth", "RawMaterial")
+techno_data <- techno_data %>% 
+  mutate(Area = AxisWidth*AxisLength)
+
+area <- mean_plot(techno_data, "Chronology", "Area", "RawMaterial")
 thickness <- mean_plot(techno_data, "Chronology", "Thickness", "RawMaterial")
 
-meanplots <- ggarrange(length, width, thickness, ncol = 1, nrow = 3, labels = c("A", "B", "C"), common.legend = TRUE, legend = "right")
+meanplots <- ggarrange(area, thickness, ncol = 1, nrow = 2, common.legend = TRUE, legend = "right")
 
 meanplots
+
+```
+
+\newline
+
+With few exceptions (platform angle and facial scar distribution), most of the morpho-functional attributes show a higher degree of variability than technological attributes, and no significant patterns could be clearly noticed.
+However, two attributes were relatively constant across all occupations: most damaged platforms present less than 45º angles; and, most pieces have bifacial scars on one damaged platform and unifacial on the other.
+On a closer inspection, we find that most of the differences within the morpho-functional variables occur between raw materials rather than between chronologies.
+
+```{r }
+
+# Filter attributes
+morpho_data <- all_data$morpho_data
+morpho_data <- morpho_data %>% 
+  filter(DamagedPlatforms == 2) %>% 
+  select(RawMaterial,
+         ScarDistribution_PlatformA, ScarDistribution_PlatformB,
+         ScarArrangement_PlatformA,ScarArrangement_PlatformB,
+         ScarExtention_PlatformA,ScarExtention_PlatformB,
+         ScarFacialDistribution_PlatformA, ScarFacialDistribution_PlatformB,
+         ScarEdgeDelineation_PlatformA, ScarEdgeDelineation_PlatformB,
+         Angle_PlatformA, Angle_PlatformB)
+
+# Recode based on 10% frequency rule
+condense()
+
+# Join attributes
+m2 <- m2 %>% 
+  mutate(Damage_Degree = str_c(Damage_Degree_PlatformA,Damage_Degree_PlatformB, sep = "/")) %>% 
+  mutate(Scar_Shape = str_c(ScarShape_PlatformA,ScarShape_PlatformB, sep = "/")) %>%
+  mutate(Scar_Distribution = str_c(ScarDistribution_PlatformA,ScarDistribution_PlatformA, sep = "/")) %>%
+  mutate(Scar_Arrangement = str_c(ScarArrangement_PlatformA,ScarArrangement_PlatformB, sep = "/")) %>%
+  mutate(Scar_Extension = str_c(ScarExtention_PlatformA,ScarExtention_PlatformB, sep = "/")) %>%
+  mutate(Scar_FacialDistribution = str_c(ScarFacialDistribution_PlatformA,ScarFacialDistribution_PlatformB, sep = "/")) %>%
+  mutate(Scar_EdgeDelineation = str_c(ScarEdgeDelineation_PlatformA,ScarEdgeDelineation_PlatformB, sep = "/")) %>%
+  mutate(Angle = str_c(Angle_PlatformA,Angle_PlatformB, sep = "/"))
 

### 2018-05-11 R/scaled_pieces_functions.R

@@ -3,10 +3,10 @@
 # READ DATA ----------------------------------------------------------------
 
 read_data <- function(){
-  #tech_attributes <- read.csv("../data/raw_data/technology_attributes.csv")
+  #tech_attributes <- readcsv("../data/raw_data/technology_attributes.csv")
   #morpho_attributes <- read.csv("../data/raw_data/morpho_attributes.csv")
-  techno_data <- read.csv("../data/raw_data/technology_dataset.csv")
-  morpho_data <- read.csv("../data/raw_data/morpho_dataset.csv")
+  techno_data <- read_csv("../data/raw_data/technology_dataset.csv")
+  morpho_data <- read_csv("../data/raw_data/morpho_dataset.csv")
 
   return(list(morpho_data = morpho_data,
               techno_data = techno_data))
@@ -78,9 +78,9 @@ require(tidyverse)
                   n.headings = FALSE,
                   freq.text.label = "none")
 
-  ct <- ct %>%
-    as.tibble() %>%
-    dplyr::select(" " = Variable, Chert, Quartz, Chalcedony, Total = Overall)
+  #ct <- ct %>%
+   # as.tibble() %>%
+    #dplyr::select(" " = Variable, Chert, Quartz, Chalcedony, Total = Overall)
 
 }
 
@@ -97,14 +97,14 @@ mean_tb <- function(dataset,x,y) {
   mt <- tabmulti(dataset, x, y,
                  cell = "n",
                  parenth = "col.percent",
-                 p.include = FALSE,
+                 p.include = TRUE,
                  n.headings = FALSE,
                  freq.text.label = "none",
                  means.text.label = "none")
 
-  mt <- mt %>%
-    as.tibble() %>%
-    dplyr::select(" " = Variable, Chert, Quartz, Chalcedony, Overall)
+ # mt <- mt %>%
+  #  as.tibble() %>%
+   # dplyr::select(" " = Variable, Chert, Quartz, Chalcedony, Overall)
 
 }
 
@@ -117,16 +117,18 @@ mean_plot <- function(dataset, x, y, z){
 
 require(ggpubr)
 
-  ggbarplot(techno_data, x = x, y = y,
+  ggbarplot(dataset, x = x, y = y,
             add = c("mean_sd", "jitter"), size = 1,
             color = z, palette = c("#00AFBB", "#E7B800", "#FC4E07"),
-            position = position_dodge(0.8)) +
+            position = position_dodge(0.8),
+            xlab = "") +
[…]

### 2018-05-13 analysis/paper/paper.Rmd
advances in chi-square functions

--- 2018-05-08 analysis/paper/paper.Rmd
@@
 
-In scaled pieces, metrics are a difficult set of variables to clearly show as results.
-These attributes are impacted by both the initial blank size as well as by the functional use and reduction.
-Consideranging the typological axes (see Table \@ref(tab:tech_attributes)) for definition) length ranging from `r min(techno_data$AxisLength, na.rm = TRUE)` mm to `r max(techno_data$AxisLength, na.rm = TRUE)` mm, width ranging from `r min(techno_data$AxisWidth, na.rm = TRUE)` mm to `r max(techno_data$AxisWidth, na.rm = TRUE)` mm and thickness ranging from `r min(techno_data$Thickness, na.rm = TRUE)` mm to `r max(techno_data$Thickness, na.rm = TRUE)` mm.
+\newline
+
+As in other classes of stone tools, scaled pieces' metric attributes are impacted by both the initial blank size as well as by the intensity of their use.
+Specifically, because some scaled pieces are used in multiple axes, direct comparisons for length and width of the typological axes (see Table \@ref(tab:tech_attributes)) for definition) cannot be straighforwardly performed.
+Since the majority of the analyzed pieces presented a rectangular outline, we use the area of a rectangle (i.e.
+Length x Width) as an approximation for the overall dimensions of the artifacts.
+Area calculations revealed a maximum of `r round(max(techno_data$Area, na.rm = TRUE), digits = 2)` mm and a minimum of `r round(max(techno_data$Area, na.rm = TRUE), digits = 2)` mm.
+For thickness the maximum is `r round(max(techno_data$Thickness, na.rm = TRUE), digits = 2)` mm and the minimum is `r round(min(techno_data$Thickness, na.rm = TRUE), digits = 2)` mm.
 In general, mean values tend to be similiar between raw materials, with some differences occuring within the Proto-Solutrean and Magdalenian assemblages, most certainly as a result of the small samples analysed for each of these periods (Figure \@ref(fig:meanplots)).
+Across techno-complexes, however, no significant statistical differences were detected when using an ANOVA test for both Area (`r anova_apa(aov(Area~Chronology, techno_data), "Chronology")` and  Thickness (`r anova_apa(aov(Thickness~Chronology, techno_data), "Chronology")`).
 
-```{r meanplots, echo = FALSE, fig.cap="Barplots of means for Length, Width and Thickness of scaled pieces by raw material and across the four chronological phases.
+When plotted against the number of damaged platforms, results for the Area variable reveal that pieces with four damaged platforms are amongst the largest, and that there is not a visible difference between the thickness of the pieces comprising each group.
+
+\newline
+
+```{r meanplots, echo = FALSE, fig.cap="Barplots of means for Area (Length x Width) and Thickness of scaled pieces by raw material and across the four chronological phases.
 Error bars represent standard deviations."}
 
-length <- mean_plot(techno_data, "Chronology", "AxisLength", "RawMaterial")
-width <- mean_plot(techno_data, "Chronology", "AxisWidth", "RawMaterial")
+techno_data <- techno_data %>% 
+  mutate(Area = AxisWidth*AxisLength)
+
+area <- mean_plot(techno_data, "Chronology", "Area", "RawMaterial")
 thickness <- mean_plot(techno_data, "Chronology", "Thickness", "RawMaterial")
 
-meanplots <- ggarrange(length, width, thickness, ncol = 1, nrow = 3, labels = c("A", "B", "C"), common.legend = TRUE, legend = "right")
+meanplots <- ggarrange(area, thickness, ncol = 1, nrow = 2, common.legend = TRUE, legend = "right")
 
 meanplots
+
+```
+
+\newline
+
+```{r plotbyplatforms, echo = FALSE, fig.caption = ""}
+
+area_by_platforms <- box_plot(techno_data, "DamagedPlatforms", "Area", "", "Area (Length * Width)")
+thickness_by_platforms <- box_plot(techno_data, "DamagedPlatforms", "Thickness", "Number of Damaged Platforms", "Thickness (mm)")
+
+boxplots <- ggarrange(area_by_platforms, thickness_by_platforms, ncol = 1, nrow = 2, common.legend = TRUE, legend = "right")
+
+boxplots
+
+```
+
+\newline
+
+With few exceptions, most of the morpho-functional attributes show a higher degree of variability than technological attributes, and no significant patterns could be clearly noticed.
+In fact, only two attributes were relatively constant across all occupations: most damaged platforms present less than 45º angles; and, a large percentage of pieces with two damaged platforms present bifacial scars on one of the platforms and unifacial scars on the opposite.
+
+On a closer inspection, we find that most of the differences within the morpho-functional variables occur between raw materials rather than between chronologies.
+
+```{r }
+
+# Filter attributes
+morpho_data <- all_data$morpho_data
+morpho_data <- morpho_data %>% 
+  filter(DamagedPlatforms == 2)
+  select(RawMaterial, Chronology,
+         ScarDistribution_PlatformA, ScarDistribution_PlatformB,
+         ScarArrangement_PlatformA,ScarArrangement_PlatformB,
+         ScarExtention_PlatformA,ScarExtention_PlatformB,
+         ScarFacialDistribution_PlatformA, ScarFacialDistribution_PlatformB,
+         ScarEdgeDelineation_PlatformA, ScarEdgeDelineation_PlatformB,
+         Angle_PlatformA, Angle_PlatformB)
+
+var_list <- c("DamageDegree", "ScarShape", "ScarDistribution", "ScarArrangement", "ScarExtension", "ScarFacialDistribution")
+
+# Filter by chronology
+filter <- split(morpho_data_2plat, morpho_data$Chronology)
+
+# Get cross-tables by chronology using cross_tb() function
+Gravettian <- cross_tb(filter$Gravettian, "RawMaterial", var_list) 
+Gravettian <- select(Gravettian, " ", Chert, Quartz, Total) #to remove chalcedony
+
+# Recode based on 10% frequency rule
+condense(morpho_data)
+
+# Join attributes
+morpho_data_2plat <- morpho_data %>% 
+  mutate(DamageDegree = str_c(DamageDegree_PlatformA,DamageDegree_PlatformB, sep = "/")) %>% 
+  mutate(ScarShape = str_c(ScarShape_PlatformA,ScarShape_PlatformB, sep = "/")) %>%
+  mutate(ScarDistribution = str_c(ScarDistribution_PlatformA,ScarDistribution_PlatformB, sep = "/")) %>%
+  mutate(ScarArrangement = str_c(ScarArrangement_PlatformA,ScarArrangement_PlatformB, sep = "/")) %>%
+  mutate(ScarExtension = str_c(ScarExtention_PlatformA,ScarExtention_PlatformB, sep = "/")) %>%
+  mutate(ScarFacialDistribution = str_c(ScarFacialDistribution_PlatformA,ScarFacialDistribution_PlatformB, sep = "/")) %>%
+  mutate(Scar_EdgeDelineation = str_c(ScarEdgeDelineation_PlatformA,ScarEdgeDelineation_PlatformB, sep = "/")) %>%
+  mutate(Angle = str_c(Angle_PlatformA,Angle_PlatformB, sep = "/"))
 

### 2018-05-13 R/chi_square_functions.R

@@ -1,0 +1,76 @@
+# Filter 2 damaged platforms
+
+morpho_data_platA <- morpho_data %>%
+  filter(DamagedPlatforms == "2" & RawMaterial != "Chalcedony") %>%
+  select(4,5,7:18)
+
+cnames <- colnames(morpho_data_platA)
+cnames <- cnames %>%
+  str_replace("_PlatformA", "")
+names(morpho_data_platA) <- cnames
+
+
+morpho_data_platB <- morpho_data %>%
+  filter(DamagedPlatforms == "2" & RawMaterial != "Chalcedony") %>%
+  select(4,5,19:30)
+
+cnames <- colnames(morpho_data_platB)
+cnames <- cnames %>%
+  str_replace("_PlatformB", "")
+names(morpho_data_platB) <- cnames
+
+
+# Bind both tables into morpho_data_2plat
+
+morpho_data_2plat <- bind_rows(morpho_data_platA, morpho_data_platB)
+morpho_data_2plat <- select_if(morpho_data_2plat, is.character)
+
+
+# Run condense() to combine all attributes with presence lower than 10%
+
+var_list <- colnames(morpho_data_2plat)
+
+morpho_data_2plat <- condense_to_other(morpho_data_2plat, "DamageDegree")
+morpho_data_2plat <- condense_to_other(morpho_data_2plat, "ScarShape")
+morpho_data_2plat <- condense_to_other(morpho_data_2plat, "ScarDistribution")
+morpho_data_2plat <- condense_to_other(morpho_data_2plat, "ScarArrangement")
+morpho_data_2plat <- condense_to_other(morpho_data_2plat, "ScarExtention")
+morpho_data_2plat <- condense_to_other(morpho_data_2plat, "ScarFacialDistribution")
+morpho_data_2plat <- condense_to_other(morpho_data_2plat, "ScarEdgeDelineation")
+morpho_data_2plat <- condense_to_other(morpho_data_2plat, "Angle")
+
+# Run condense() to combine all attributes with presence lower than 10%
+
+morpho_data_2plat <- condense_to_NA(morpho_data_2plat, "DamageDegree")
+morpho_data_2plat <- condense_to_NA(morpho_data_2plat, "ScarShape")
+morpho_data_2plat <- condense_to_NA(morpho_data_2plat, "ScarDistribution")
+morpho_data_2plat <- condense_to_NA(morpho_data_2plat, "ScarArrangement")
[…]

### 2018-05-13 R/scaled_pieces_functions.R

@@ -71,16 +71,19 @@ cross_tb <- function(dataset,x,y) {
 require(tab)
 require(tidyverse)
 
+dataset <- as.data.frame(dataset)
+
   ct <- tabmulti(dataset, x, y,
                   cell = "n",
                   parenth = "col.percent",
                   p.include = FALSE,
+                  listwise.deletion = TRUE,
                   n.headings = FALSE,
                   freq.text.label = "none")
 
-  #ct <- ct %>%
-   # as.tibble() %>%
-    #dplyr::select(" " = Variable, Chert, Quartz, Chalcedony, Total = Overall)
+  ct <- ct %>%
+   as.tibble() %>%
+   dplyr::select(" " = Variable, Chert, Quartz, Chalcedony, Total = Overall)
 
 }
 
@@ -122,10 +125,28 @@ require(ggpubr)
             color = z, palette = c("#00AFBB", "#E7B800", "#FC4E07"),
             position = position_dodge(0.8),
             xlab = "") +
-  scale_x_discrete(limits=c("Gravettian", "Proto-Solutrean","Solutrean", "Magdalenian")) +
-  theme_gray()
+  scale_x_discrete(limits=c("Gravettian", "Proto-Solutrean","Solutrean", "Magdalenian"))
 
 }
+
+
+###########################################################################
+# BOXPLOT WITH JITTER -----------------------------------------------------
+
+box_plot <- function(df, x_var, y_var, x_label, y_label){
+  b_plot<-ggplot(df,aes_string(x= x_var, y=y_var)) +
+    geom_boxplot(outlier.shape = NA, fill = "gray")
+
+  b_plot <- b_plot +
+    geom_jitter(position=position_jitter(width=.2, height=0), aes(color = RawMaterial), size = 4) +
+    scale_color_manual(values = c("#00AFBB", "#E7B800", "#FC4E07"), name = "Raw Material") +
+    labs(x = x_label, y = y_label)+
+    scale_fill_discrete(guide=FALSE) +
+    theme(text = element_text(size = 20), axis.text = element_text(size = 20))
+
+  return(b_plot)
+}
+
 
 ###########################################################################
 # INLINE PERCENTAGE -------------------------------------------------------
@@ -170,7 +191,6 @@ compare_list <- function(dataset,x){
 }
 
 
-
 ###########################################################################
 # MEAN PLOTS WITH STAT RESULTS ---------------------------------------------
 
[…]

### 2018-05-17 analysis/paper/paper.Rmd
references added

@@
       reference_docx: "../templates/template2.docx" 
-bibliography: ShortTermOccupations.bib
+bibliography: ScaledPiecesVB.bib
 csl: "../templates/springer-humanities-author-date.csl"
@@
 ```{r, setup, include = FALSE, echo = FALSE}
+
 knitr::opts_chunk$set(
@@
   fig.path = "../figures/")
+  cache = TRUE
 
@@
 
+library(knitr)
 library(tidyverse)
@@
 
-Bipolar technology is generically classified into two types of lithic artifacts - bipolar cores and scaled pieces (also known as splintered pieces or pièce esquilée).
-Their distinction, however, has not always been consensual (Hayden, 1980).
+Bipolar technology is generically classified into two types of lithic artifacts - bipolar cores and scaled pieces (also known as splintered pieces or *pièce esquilée*).
+Their distinction, however, has not always been consensual [@hayden_confusion_1980].
 In fact, the definition of scaled piece has suffered significant changes ever since its initial identification in the early 20th century and, to this day, there still seems to be no clear definition for this type of artifact.
-The first definition of scaled pieces was proposed by Baron and Bouysonnie (1906), describing them as a result of bipolar knapping through direct percussion, with the “core” rested in a hard surface, originating splintering in both ends of the tool.
-Since then, several other definitions were adopted and adapted by researchers for a very diverse set of contexts across the world (see e.g.
-Octobon, 1938; Sonneville-Bordes and Perot, 1956: 552; MacDonald; 1968:88; Hayden, 1980; Shott, 1989; Knight 1991:61).
+The first definition of scaled pieces was proposed by Bardon and Bouyssonie [-@bardon_outils_1906], describing them as a result of bipolar knapping through direct percussion, with the “core” rested in a hard surface, originating splintering in both ends of the tool.
+Since then, several other definitions were adopted and adapted by researchers for a very diverse set of contexts across the world [e.g.
+@hayden_confusion_1980; @macdonald_debert:_1985; @octobon_contribution_1938; @shott_bipolar_1989; @sonneville-bordes_lexique_1956].
 
-Beyond the classification debate, the biggest issue with this type of tools comes from a functional standpoint, a problem that has also been quite debated (see e.g., Tixier, 1963; Binford and Quimby, 1963; Semenov, 1964; Flood, 1980:308; LeBlanc, 1992; Shott, 1999; Lucas and Hays, 2004; de la Peña, 2011; Igreja & Porraz, 2013).
+Beyond the classification debate, the biggest issue with this type of tools comes from a functional standpoint, a problem that has also been quite debated [e.g.
+@binford_indian_1963; @de_la_pena_sobre_2011; @flood_moth_1980; @igreja_functional_2013; @leblanc_wedges_1992; @lucas_les_2004; @shott_bipolar_1999; @tixier_typologie_1963].
 The issue surrounding this problem lies within the functional equifinality of these artifacts.
-Contrary to bipolar cores that are most of the times easily identified as cores, scaled pieces have been associated with two distinct types of activities: (1) as intermediate pieces or wedges for working hard raw materials (bone, ivory, antler, etc.); and (2) as cores for the extraction of chips and small bladelets (see e.g.
-Brantingham et al., 2004).
-Although largely debated, the ambiguity of this classification has, however, been ignored in some of the most recent literature, with some authors not acknowledging that bipolar evidence in stone tools might also result from other activities other than only reduction strategies (see e.g.
-Hiscock, 2015).
+Contrary to bipolar cores that are most of the times easily identified as cores, scaled pieces have been associated with two distinct types of activities: (1) as intermediate pieces or wedges for working hard raw materials (e.g.
+bone, ivory, antler); and (2) as cores for the extraction of chips and small bladelets [@brantingham_early_2004].
+Although largely debated, the ambiguity of this classification has, however, been ignored in some of the most recent literature, with some authors not acknowledging that bipolar evidence in stone tools might also result from other activities other than only reduction strategies [see e.g.
+@hiscock_dynamics_2015].
 
-Most current studies on scaled pieces apply use-wear methods (e.g.
-Vaughan, 2002; Lucas & Hays, 2004; Bosinski, 2007; Gibaja et al.
-2007; Sano, 2012; Igreja & Porraz, 2013, de la Peña, 2011; 2014; 2015a; 2015b, Bader et al., 2015), focusing on the identification of polishes and use-wear patterned stigmas, through both micro and macroscopic analysis of splintered surfaces and its comparison with experimental assemblages.
+Most current studies on scaled pieces apply use-wear methods [e.g.
+@bader_lithic_2015; @de_la_pena_interpretation_2015; @de_la_pena_new_2014; @de_la_pena_qualitative_2015; @de_la_pena_sobre_2011; @gibaja_tecnologiy_2007; @igreja_functional_2013; @lucas_les_2004; @sano_functional_2012], focusing on the identification of polishes and use-wear patterned stigmas, through both micro and macroscopic analysis of splintered surfaces and its comparison with experimental assemblages.
 Frequently, these studies coincide in interpreting scaled pieces as intermediate elements for the work of hard organic raw materials.
 The study by P.
-de la Peña (2011) is one of the most recent and relevant references in this regard.
-The author presents the results of an experimental program aiming to identify specific wear patterns in the use of bipolar techniques that allow the distinction between different types of activities/worked materials.
+de la Peña [-@de_la_pena_sobre_2011; -@de_la_pena_qualitative_2015] is one of the most recent and relevant references in this regard.
+The author presents the results of an experimental program aiming to identify specific wear patterns in the use of bipolar techniques that allow the distinction between different types of activities and worked materials.
 Results indicate that while no visible differences can be identified in the percussion area, significant variation can be observed in the morphology of the areas in contact with the worked material.
 
-As in many other regions and stone age periods across the world (e.g., Diez et.
-al 2009; Igreja & Porraz, 2013; Langjans, 2012; Leakey, 1971: 221; Leblanc, 1992; Shott, 1989, 1999; White, 1968), evidence of bipolar technologies are quite ubiquitous in European Upper Paleolithic contexts.
-In many sites, bipolar elements are classified as scaled pieces, a morphotype that has been putatively considered as one the most common lithic tools in Upper Paleolithic assemblages (de la Peña, 2015).
-Different functional connotations for this type of artifacts have been presented in several western european Upper Paleolithic studies (e.g., Zilhão, 1997; Aubry et al.
-1998; de la Peña 2011; Sano, 2012).
+As in many other regions and stone age periods across the world [e.g.
+@diez-martin_were_2009; @igreja_functional_2013; @langejans_middle_2012; @white_fabricators_1968], evidence of bipolar technologies are quite ubiquitous in European Upper Paleolithic contexts.
+In many sites, bipolar elements are classified as scaled pieces, a morphotype that has been putatively considered as one the most common lithic tools in Upper Paleolithic assemblages [@de_la_pena_qualitative_2015], but different functional connotations for this type of artifacts have been presented in several western european Upper Paleolithic studies [e.g.
+@de_la_pena_sobre_2011; @sano_functional_2012; @zilhao_o_1997].
 
-In the case of the westernmost regions of Iberia, scaled pieces are commonly found in archaeological contexts ranging from the Upper Paleolithic to the Neolithic (e.g.
-Aubry et al.
-1998; Bicho 1992; Carvalho 1998; Zilhão 1997).
-While scarce, the majority of Portuguese paleolithic studies (e.g.
-Almeida 2000, Gameiro 2012) do not offer analytical interpretations for the presence of scaled pieces, simply using references that interpret them as bipolar cores for the extraction of small bladelets and chips to be supposedly used in composite tools (as argued by Zilhão, 1997) or as intermediate pieces for working hard materials (Gibaja et al., 2007; Marreiros, 2009).
-Zilhão (1997) for example argues that the presence of scaled pieces throughout the Upper Paleolithic sequence of central Portugal is inversely proportional to the presence of “carinated cores”, and thus it is likely that the former should have worked as a flexible substitute for the latter.
-Carvalho (1998) agrees with this interpretation and considers it also valid for the Portuguese Early Neolithic.
+In the case of the westernmost regions of Iberia, scaled pieces are commonly found in archaeological contexts ranging from the Upper Paleolithic to the Neolithic [e.g.
+@bicho_technological_2000; @carvalho_talhe_1998; @zilhao_o_1997].
+While scarce, the majority of portuguese paleolithic studies [e.g.
+@almeida_terminal_2000; @goncalves_variabilite_2012] do not offer analytical interpretations for the presence of scaled pieces, simply using references that interpret them as bipolar cores for the extraction of small bladelets and chips to be supposedly used in composite tools [e.g.
+@zilhao_o_1997] or as intermediate pieces for working hard materials [@gibaja_tecnologiy_2007; @marreiros_as_2009].
+Zilhão [-@zilhao_o_1997] for example argues that the presence of scaled pieces throughout the Upper Paleolithic sequence of central Portugal is inversely proportional to the presence of “carinated cores”, and thus it is likely that the former should have worked as a flexible substitute for the latter.
+Carvalho [-@carvalho_talhe_1998] agrees with this interpretation and considers it also valid for the portuguese early Neolithic.
 
@@
 In this paper, we present a first approach to the characterization of the morpho-technological variability and consequent role of scaled pieces during the Upper Paleolithic of the westernmost regions of Iberia (c.
-32 000 – 10 000 cal BP).
+32–10 ka cal BP).
 Using data from the multi-component, thoroughly dated, site of Vale Boi, located in southern Portugal, we present the analysis of a series of technological and morpho-functional attributes of a relatively large set of scaled pieces coming from one of the areas of the site.
@@
 
-A complete sequence of Upper Paleolithic has been identified at Vale Boi, with all the traditionally-defined techno-complexes (Gravettian, Proto-Solutrean, Solutrean and Magdalenian) being identified across the three main excavation areas: the Terrace, the Rockshelter, and the Slope.
+A rather complete Upper Paleolithic sequence has been identified at Vale Boi, with all the traditionally-defined techno-complexes (Gravettian, Proto-Solutrean, Solutrean and Magdalenian) being identified across the three main excavation areas: the Terrace, the Rockshelter, and the Slope.
 
 The Terrace area is located in the lower part of the hill.
-In this area, the longest archaeological sequence of the site can be found, including the complete Upper Paleolithic sequence but also three Holocene horizons, corresponding to Neolithic, Mesolithic and Epipaleolithic occupations (Bicho et al., 2012; Cascalheira et.
-al, 2017).
+In this area, the longest archaeological sequence of the site can be found, including the complete Upper Paleolithic sequence but also three Holocene horizons, corresponding to Neolithic, Mesolithic and Epipaleolithic occupations [@bicho_ledge:_2012; @cascalheira_cross-scale_2017].
 From within the lower levels of this area, an Early Gravettian occupation was discovered, dated to c.
-32 ka cal BP, being one of the earliest radiocarbon dates for anatomically modern humans in Southern Iberia (Bicho et al., 2016; Marreiros et al., 2015).
+32 ka cal BP, being one of the earliest radiocarbon dates for anatomically modern humans in Southern Iberia [@bicho_ecodynamics_2013; @marreiros_lithic_2015].
 
 The Rockshelter area is a collapsed rock shelter located in the upper part of the slope, a couple of meters below the limestone cliff.
-This collapse would have occurred after the Last Glacial Maximum, since below the collapsed debris, several Solutrean occupations can be found, overlaying a sequence of very ephemeral Gravettian horizons (Marreiros, 2009; Cascalheira, 2009; Manne et al., 2012).
+This collapse would have occurred after the Last Glacial Maximum, since below the collapsed debris, several Solutrean occupations can be found, overlaying a sequence of very ephemeral Gravettian horizons [@cascalheira_vale_2012; @manne_intensive_2012; @marreiros_as_2009].
 The Solutrean is dated to between c.
-20 ka and 25 ka cal BP (Cascalheira et al., 2012), while the Gravettian is dated between 26 ka and 32 ka cal BP (Marreiros et al., 2015).
+20 ka and 25 ka cal BP [@cascalheira_chronological_2015], while the Gravettian is dated between 26 ka and 32 ka cal BP [@marreiros_lithic_2015].
 
 Finally, the Slope section, from where the assemblage here presented is coming from, is composed of a series of excavation areas opened across the mid-hill sector of the site.
-These areas exhibit heterogeneous conditions in terms of site formation processes and archaeological preservation (Manne et al., 2012), but all revealed the presence of occupations attributed to the Gravettian, Proto-Solutrean, Solutrean and Magdalenian.
+These areas exhibit heterogeneous conditions in terms of site formation processes and archaeological preservation [@manne_intensive_2012], but all revealed the presence of occupations attributed to the Gravettian, Proto-Solutrean, Solutrean and Magdalenian.
 Like in the previous areas, remains are well preserved, and high frequencies of lithic artifacts, malacological and mammalogical fauna and bone tools were recovered.
-No habitation features were identified in this area and based on the conditions and type of artifacts found, it has been suggested that this area would have functioned as a midden deposit (Bicho et al., 2010, 2012a).
+No habitation features were identified in this area and based on the conditions and type of artifacts found, it has been suggested that this area would have mostly functioned as a midden deposit [@bicho_ledge:_2012; @bicho_paleolithique_2010].
 
@@
 
-Vale Boi’s lithic studies have revealed a general tendency for stable technological and functional patterns throughout the Upper Paleolithic (see e.g.
-Bicho et al., 2012; Gibaja and Bicho, 2006; Bradtmöller et.
-al, 2015; Marreiros, 2009, 2013, 2015; Cascalheira, 2009, 2013).
+Vale Boi’s lithic studies have revealed a general tendency for stable technological and functional patterns throughout the Upper Paleolithic [e.g.
+@bicho_ledge:_2012; @bradtmoller_lithic_2016; @cascalheira_cross-scale_2017; @cascalheira_influencia_2013; @cascalheira_vale_2012; @gibaja_funcion_2011; @marreiros_as_2009; @marreiros_lithic_2013; @marreiros_lithic_2015].
 This can be explained, partially, by raw material availability.
-Most raw materials were procured locally or regionally from deposits located at no more than 20 km away from the site (Bicho et al., 2010a).
+Most raw materials were procured locally, or regionally from deposits located at no more than 20 km away from the site [@bicho_paleolithique_2010].
 Chert was the most used rock type for more complex retouched tools production, while quartz and greywacke were mostly used for flake extraction and simple retouched tools production.
-Other raw materials can be found at the site, but much more restricted, both diachronically and functionally within each techno-complex and each archaeological horizon.
+Other raw materials can be found at the site, but much more restricted, both diachronically and functionally, within each techno-complex.
 Schist, for instance, shows up in some occupations almost exclusively connected to mobile art.
-Dolerite can only be found in the Proto-Solutrean levels, while chalcedony is mostly limited to the Proto-Solutrean and Solutrean levels.
+Dolerite was only  found in the Proto-Solutrean levels [@belmiro_o_2017], while chalcedony is mostly limited to the Proto-Solutrean and Solutrean levels [@cascalheira_vale_2012].
 
@@
 Throughout the several occupations, chert was considered the best rock type for knapping, which is evident by the way it was explored with much more complex strategies than either greywacke or quartz.
-While it can be said that chert exploration strategies were more elaborate, with exception of the bifacial technology during the Solutrean, they were still simple unidirectional reduction sequences producing mostly small debitage.
-Elongated products are found at very low frequencies across all chronologies.
-When present, retouched tools are mostly notches, denticulates, end-scrapers and, if considered, scaled pieces.
-One possible reason for these simplistic strategies seems to be an overall low knapping quality, since nodules are quite small, and frequently show a high degree of tectonically derived fractures (Pereira et al.
-2016).
+While it can be said that chert exploration strategies were more elaborate, the most common strategies were still simple unidirectional or bidirectional reduction sequences, producing mostly flakes [@marreiros_flake_2012].
+Elongated products are found at low frequencies across all chronologies.
+When present, retouched tools are mostly notches, denticulates, end-scrapers, and scaled pieces.
+One possible reason for these simplistic strategies seems to be an overall low knapping quality, since nodules are quite small, and frequently show a high degree of tectonically derived fractures [@pereira_territory_2016].
 
@@
 The first is a thick grain, low-quality type, mostly inadequate for knapping.
-Still, this kind of quartz is present in large quantities and is most likely associated with stone boiling and grease rendering activities (Manne & Bicho, 2009; Manne, 2010; Bicho et al, 2012; Manne et al.
-2012, 2014).
+Still, this kind of quartz is present in large quantities and is most likely associated with stone boiling and grease rendering activities [@bicho_ledge:_2012; @manne_early_2014; @manne_intensive_2012; @manne_upper_2010; @manne_vale_2009].
 The second type of quartz is characterized by finer-grain small pebbles with yellowish cortex.
-This type of quartz was knapped using rather simple strategies, mostly for small flake production, which in turn were used to produce simple and versatile tools (e.g.
-scaled pieces).
+This type of quartz was knapped using rather simple strategies, mostly for small flake production, which in turn were used to produce simple and versatile tools.
 
-Greywacke was used for the rare production of flakes and choppers but shows up in the site mostly in the form of large slabs in which the identification of concavely shaped impact marks is thought to result from their use as anvils (Manne et.
-al, 2012, Manne & Bicho, 2009).
-The importance of greywacke anvils is indicated by its high frequency, with hundreds of slabs found in many levels throughout all occupations (Bicho et al, 2012; Manne et al.
-2012).
-Greywacke was also knapped for flake extraction, using expedient unidirectional reduction strategies.
+Greywacke is the third most frequent raw material at Vale Boi, showing up in the site mostly in the form of large slabs, in which the identification of concavely shaped impact marks is thought to result from their use as anvils.
+The importance of greywacke anvils is indicated by its high frequency, with hundreds of slabs found in many levels throughout all occupations [@bicho_ledge:_2012; @manne_intensive_2012].
+Greywacke was also knapped for flake extraction, using expedient unidirectional reduction strategies [@cascalheira_influencia_2013; @marreiros_as_2009].
 
@@
 
-Organic preservation is good at Vale Boi.
-Faunal remains can be frequently found in every occupation, from both terrestrial and marine sources.
-Marine fauna at the site is marked by the presence of mollusks, crustaceans, some fish vertebrae and, in rare occasions, marine mammal remains (Manne, 2010).
-Regarding terrestrial fauna, three species dominate the vertebrate group: rabbit (Oryctolagus cuniculus), red deer (Cervus elaphus) and horse (Equus caballus).
-Smaller amounts of wild ass (Equus hydruntinus), aurochs (Bos primigenius), ibex (Capra pyrenaica) and wild boar (Sus scrofa) are also present (Manne, 2010, 2012, 2014).
+Organic preservation is fairly good at Vale Boi.
+Faunal remains can be frequently found in all occupations, both of terrestrial and marine resources.
+Marine fauna at the site is marked by the presence of mollusks, crustaceans, some fish vertebrae and, in rare occasions, marine mammal remains [@manne_upper_2010].
+Regarding terrestrial fauna, three species dominate the vertebrate group: rabbit (*Oryctolagus cuniculus*), red deer (*Cervus elaphus*) and horse (*Equus caballus*).
+Smaller amounts of wild ass (*Equus hydruntinus*), aurochs (*Bos primigenius*), ibex (Capra pyrenaica) and wild boar (Sus scrofa) are also present [@manne_early_2014; @manne_intensive_2012; @manne_upper_2010].
 One of the most interesting patterns within the faunal assemblages is that a large portion of ungulate remains present specific types of fracture that have been associated with bone marrow extraction activities.
-Red deer and horse remains frequently show evidence of opposed cone fractures, trituration and smashing (Manne & Bicho, 2009; Manne, 2010, 2012, 2014).
+Red deer and horse remains frequently show evidence of opposed cone fractures, trituration and smashing [@manne_early_2014; @manne_intensive_2012; @manne_upper_2010; @manne_vale_2009].
 
@@
 While there were conditions for the whole bones to be preserved, these bones are frequently anthropically broken.
-These fragmentation patterns seem to be linked to grease rendering activities (Manne, 2010, 2012, 2014).
+These fragmentation patterns seem to be linked to grease rendering activities [@manne_early_2014; @manne_intensive_2012; @manne_upper_2010].
 The main goal of grease rendering is to obtain grease, through heat exposure from animal bones, as it has a very high caloric value.
-Other than this, this grease could have many uses with the addition of being easily stored and transported (Manne, 2012).
+Other than this, this grease could have many uses with the addition of being easily stored and transported [@manne_intensive_2012].
 The spongeous bone parts would be fragmented and deposited in a hole, covered with animal pelts full of water, after which, pre-heated rock fragments were added.
-The high temperature of the rocks would make the water boil and therefore separate the grease from the spongeous bones creating a highly nutritional stew (Manne, 2014).
+The high temperature of the rocks would make the water boil and therefore separate the grease from the spongeous bones creating a highly nutritional stew.
 After being cooled the grease would accumulate at the top where it could be easily removed, transported and stored.
-Unlike simple bone marrow extraction, this method involved large preparation, including water transport, fire, and heat production, rock heating, and finally the storing of the grease (Manne, 2014).
+Unlike simple bone marrow extraction, this method involved large preparation, including water transport, fire, and heat production, rock heating, and finally the storing of the grease [@manne_early_2014].
 
-These grease rendering techniques are thought to be quite common at the site since the parts of the bones with higher amounts of fat are missing despite the good preservation of the rest of the remains (Manne et al.
-2012).
+These grease rendering techniques are thought to be quite common at the site since the parts of the bones with higher amounts of fat are missing despite the good preservation of the rest of the remains.
 Fragmented ungulate remains show up in the site connected to large amounts of thermo-altered quartz, anvils, hard hammers and scaled pieces.
-This suggests that red deer and horse bones were processed and afterward intensively grease rendered (Manne & Bicho, 2009; Manne, 2010, 2012, 2014).
-To confirm this, a single scaled piece was found stuck to a cracked phalanx in a Gravettian horizon (Manne, 2012).
+This suggests that red deer and horse bones were processed and afterward intensively grease rendered.
+To confirm this, a single scaled piece was found stuck to a cracked phalanx in a Gravettian horizon [@manne_early_2014; @manne_intensive_2012; @manne_upper_2010; @manne_vale_2009].
 
@@
 Attribute analysis was split into two main groups, each corresponding to two distinct types of features: (1) technological attributes and (2) morpho-functional attributes.
+In the first group, a series of variables traditionally used in lithic studies [e.g.
+@andrefsky_jr_lithics:_2005; @inizan_technology_1999] were recorded (Table \@ref(tab:tech_attributes)), aiming to characterize patterns of blank choice for the application of bipolar technology.
+For the second group of variables, a macroscopic approach building upon the work of de la Peña [-@de_la_pena_sobre_2011] and Fischer et.
+al [-@fischer_macro_1984] was adopted (Table \@ref(tab:morph_attributes)).
+Studies by Sano [-@sano_functional_2012], and Gibaja et al.
+[-@gibaja_tecnologiy_2007] indicate that since use-wear traces are formed before the splintering, the removal of small chips removes most of the polishes and use-wear traces left by the contact with the static element (either a stone anvil or a hard-organic materials).
+This, together with the large presence of quartz artifacts in our sample, prevented us to pursue a microscopic use-wear approach for this study in particular.
 
-In the first group, a series of variables traditionally used in lithic studies (e.g.
-Andrefsky, 2005; Inizan et.
-al, 1999) were recorded (Table \@ref(tab:tech_attributes)), aiming to characterize patterns of blank choice for the application of bipolar technology.
+\newline
 
-For the second group of variables a macroscopic approach, building upon the work of de la Peña (2011) and Fischer et.
-al (1984) was adopted (Table \@ref(tab:morph_attributes)).
-Studies by Vaughan (2002), Sano (2012), and Gibaja et al.
-(2007) indicate that since use-wear traces are formed before the splintering, the removal of small chips removes most of the polishes and use-wear traces left by the contact with the static element (either a stone anvil or a hard-organic materials).
-This prevented us to pursue a microscopic use-wear approach in this study.
+Macroscopic morpho-functional attributes were separately analyzed for each damaged platform, aiming to detect patterns of morphological change that occurred in artifacts during use.
+Following the work of de la Peña [-@de_la_pena_qualitative_2015; -@de_la_pena_sobre_2011] and Gonzalez-Urquijo & Ibanez-Estévez [-@gonzalez-urquijo_metodologialisis_1994], we expected that these attributes would be indicative of which function the artifacts had.
+For instance, according to de la Peña [-@de_la_pena_qualitative_2015; -@de_la_pena_sobre_2011], for pieces used as wedges the delineation of the damaged platforms are constantly asymetrical and only the hammered edge clearly shows the typical *écaille* retouch.
+Furthermore, these pieces would have irregular shapes, variable scar size, and irregular scar distribution.
+On the other hand, pieces used as bipolar cores would have squared or rectangular shapes, symmetric straight damaged platforms, and a higher frequency of scars on the hammered edge than on the opposed edge.
+The addition of other attributes expanding on the work by Gonzalez-Urquijo & Ibanez-Estévez [-@gonzalez-urquijo_metodologialisis_1994] was made following the same reasoning.
+We expected, thus, to be able to differentiate pieces used as wedges from pieces used as cores, as both groups would show distinct combinations of attributes.
+To assist us with this differentiation we tested the presence of the referred patterns within our assemblage using descriptive and multivariate statistical analysis and comparing it with the data described by the referred authors.
 
@@
 
-Macroscopic morpho-functional attributes were separately analyzed for each damaged platform, aiming to detect patterns of morphological change that occurred in artifacts during use.
-Following the work of de la Peña (2011, 2015) and Gonzalez-Urquijo & Ibanez-Estévez (1994), we expected that these attributes would be indicative of which function the artifacts had.
-For instance, according to de la Peña (2011, 2015), for pieces used as wedges, the shape of the passive platform remains the same and the other varies according to the type of material worked on the active platform.
-Furthermore, these pieces would have irregular shapes, variable scar size, and irregular scar distribution.
-On the other hand, pieces used as bipolar cores would have squared or rectangular shapes, symmetric straight platforms, and a higher frequency of scars on the striking edge than on the active edge.
-The addition of other attributes expanding on the work by Gonzalez-Urquijo & Ibanez-Estévez (1994) was made following the same reasoning.
-We expected, thus, to be able to differentiate pieces used as wedges from pieces used as cores, as both groups would show distinct combinations of attributes.
-To assist us with this differentiation we tested the presence of the referred patterns within our assemblage using descriptive and multivariate statistical analysis and comparing it with the data described by the referred authors.
-
-\newline
-
 ## Reproducibility and open source materials
@@
 
-A total of `r summarise(techno_data, n())` scaled pieces were analyzed, of which `r round(techno_data %>% filter(Chronology == "Gravettian") %>% summarise(n = n()) %>% mutate(n = (n / 139)*100), digits = 2)`% come from Gravettian, and `r round(techno_data %>% filter(Chronology == "Solutrean") %>% summarise(n = n()) %>% mutate(n = (n / 139)*100), digits = 2)`% from Solutrean levels, as shown in (Table \@ref(tab:assemblage)).The assemblage was divided into 4 main chronological groups: Magdalenian, Solutrean, Proto-Solutrean, and Gravettian.
-
-As previously mentioned, as far as the technological data is concerned, one of our main objectives was to characterize and understand the choice of blanks involved in bipolar technology.
-We did not include the metric data, as most of the data from the original blanks are mostly gone due to reduction and use.
-Overall, the technological analysis revealed some trends that lasted throughout the Upper Paleolithic, in agreement with the general patterns of lithic technology in the site (see above).
+A total of `r summarise(techno_data, n())` scaled pieces were analyzed, of which `r round(techno_data %>% filter(Chronology == "Gravettian") %>% summarise(n = n()) %>% mutate(n = (n / 139)*100), 2)`% come from Gravettian, and `r round(techno_data %>% filter(Chronology == "Solutrean") %>% summarise(n = n()) %>% mutate(n = (n / 139)*100), 2)`% from Solutrean levels, as shown in (Table \@ref(tab:assemblage)).
+In terms of raw materials, the great majority of pieces were either made on quartz or chert, with chalcedony being represented only by `r round(techno_data %>% filter(RawMaterial == "Chalcedony") %>% summarise(n = n()), 2)` artifacts.
 
@@
 
-knitr::kable(assemb, caption= "")
+knitr::kable(assemb, caption= "Frequencies of scaled pieces by raw material and chronology")
 
@@
 \newline
-
+As previously mentioned, as far as the technological data is concerned, one of our main objectives was to characterize the choice of blanks involved in bipolar technology.
+Overall, the technological analysis revealed some trends that lasted throughout the Upper Paleolithic, in agreement with the general patterns of lithic technology recorded for the site.
 Across all techno-complexes, blanks used were either flakes or unclassifiable fragments, but the choice seems to be different for each raw material.
@@
 Regarding chert, in every occupation flakes dominate the assemblages, followed by a reduced number (n = `r techno_data %>% filter(RawMaterial == "Chert" & Blank == "Non_identifiable") %>% summarise(n())`) of unclassifiable blanks.
-Chalcedony blanks are exclusively flakes.
+The small sample of Chalcedony blanks is exclusively represented by flakes.
+
+Technological and morphological characteristics of the flake blanks present very similar patterns across time and amongst raw materials.
+The blanks sought after would have straight profiles, parallel edges, no cortical surfaces and trapezoidal or triangular shaped cross-sections.
+
+Other characteristics of the assemblage are the low frequency of retouch (n = `r techno_data %>% filter(RetouchedBlank == "Yes") %>% summarise(n())`), of fire alterations (n = `r techno_data %>% filter(Fire != "No_Traces") %>% summarise(n())`), and of the presence of original striking platforms (n = `r techno_data %>% filter(ButtType != "Absent") %>% summarise(n())`).
+Still, when present, retouch was located in the lateral part of the artifacts, similar to what would define a side-scraper.
+In a very small number of cases, when striking platforms were detected, these were mostly flat.
+The absence of the original blank platforms is to be expected in this type of artifact, mostly due to the functional use of the pieces, and consequent removal of the platform, rather than an actual choice.
+The large absence of platforms may, on the other hand, be the result of the use of the technological axis as the main functional axis.
+In fact, when identifiable, technological and typological axes coincide in `r inline_perc(techno_data, AxisCoincident, "Yes")`% of the cases.
+The longitudinal section shows a large variability of shapes, independent of raw material or chronology.
+Similarly, the dorsal pattern of previous removals was rarely identified, although this, like with the case of platform absence, might occur due to the functional stigmas and be dependent on the intensity of use for each artifact.
 
@@
 Gravettian <- cross_tb(filter$Gravettian, "RawMaterial", var_list) 
-Gravettian <- select(Gravettian, " ", Chert, Quartz, Total) #to remove chalcedony
+Gravettian <- Gravettian %>% 
+  as.tibble() %>% 
+  select(" " = Variable, Chert, Quartz, Total = Overall)
 
 Solutrean <- cross_tb(filter$Solutrean, "RawMaterial", var_list)
+Solutrean <- Solutrean %>% 
+  as.tibble() %>% 
+  select(" " = Variable, Chert, Quartz, Chalcedony, Total = Overall)
 
 `Proto-Solutrean` <- cross_tb(filter$`Proto-Solutrean`, "RawMaterial", var_list)
+`Proto-Solutrean` <- `Proto-Solutrean` %>% 
+  as.tibble() %>% 
+  select(" " = Variable, Chert, Quartz, Chalcedony, Total = Overall)
 
 Magdalenian <- cross_tb(filter$Magdalenian, "RawMaterial", var_list)
-Magdalenian <- select(Magdalenian, " ", Chert, Quartz, Total) #to remove chalcedony
+Magdalenian <- Magdalenian %>% 
+  as.tibble() %>% 
+  select(" " = Variable, Chert, Quartz, Total = Overall)
 
@@
 
+```{r mestric_statistics, echo = FALSE}
+
+techno_data <- techno_data %>% 
+  mutate(Area = AxisWidth*AxisLength)
+
+anova.test.area <- anova.test(Area ~ Chronology, techno_data)
+
+anova.test.thickness <- anova.test(Thickness ~ Chronology, techno_data)
+
+```
+
 \newline
 
-Technological and morphological characteristics of the flake blanks present very similar patterns across time and amongst raw materials.
-The blanks sought after would have the following characteristics: straight profile, parallel edges, no cortical surfacem and trapezoidal or triangular shaped cross-sections.
+As in other classes of stone tools, scaled pieces' metric attributes are impacted by both the initial blank size as well as by the intensity of their use.
+Specifically, because some scaled pieces are used in multiple axes, direct comparisons for length and width of the typological axes (see Table \@ref(tab:tech_attributes)) for definition) cannot be straighforwardly performed.
+Since the majority of the analyzed pieces presented a rectangular outline, we use the area of a rectangle (i.e.
+Length x Width) as an approximation for the overall dimensions of the artifacts.
+Area calculations revealed a maximum of `r round(max(techno_data$Area, na.rm = TRUE), digits = 2)` mm and a minimum of `r round(max(techno_data$Area, na.rm = TRUE), digits = 2)` mm.
+For thickness the maximum is `r round(max(techno_data$Thickness, na.rm = TRUE), digits = 2)` mm and the minimum is `r round(min(techno_data$Thickness, na.rm = TRUE), digits = 2)` mm.
+In general, mean values tend to be similiar between raw materials, with some differences occuring within the Proto-Solutrean and Magdalenian assemblages, most certainly as a result of the small samples analysed for each of these periods (Figure \@ref(fig:meanplots)).
+Across techno-complexes, however, no significant statistical differences were detected when using an ANOVA test for both the Area ($F (`r anova.test.area[1]`, `r anova.test.area[2]`) = `r anova.test.area[3]`$, $p = `r anova.test.area[4]`)$ and Thickness ($F (`r anova.test.thickness[1]`, `r anova.test.thickness[2]`) = `r anova.test.thickness[3]`$, $p = `r anova.test.thickness[4]`)$ variables.
 
-Other characteristics of the assemblage are the low frequency of retouch (n = `r techno_data %>% filter(RetouchedBlank == "Yes") %>% summarise(n())`), fire alterations (n = `r techno_data %>% filter(Fire != "No_Traces") %>% summarise(n())`), and of the original striking platforms (n = `r techno_data %>% filter(ButtType != "Absent") %>% summarise(n())`).
-Still, when present, retouch was located in the lateral part of the artifacts, similar to what would define a side-scraper.
-In a very small number of cases when striking platforms were detected, these were mostly flat.
-The absence of the original blank platforms is to be expected in this type of artifact, mostly due to the functional use of the pieces, and consequent removal of the platform, rather than an actual choice.
-The large absence of platforms may, on the other hand, be the result of the use of the technological axis as the main functional axis.
-In fact, when identifiable, technological and typological axes coincide in `r inline_perc(techno_data, AxisCoincident, "Yes")`% of the cases.
-The longitudinal section shows a large variability of shapes, independent of raw material or chronology.
-Similarly, the dorsal pattern of previous removals was rarely identified, although this, like with the case of the platform absence, might occur due to the functional stigmas and be dependent on the the intensity of use for each artifact.
+\newline
+
+```{r meanplots, echo = FALSE, fig.cap="Barplots of means for Area (Length x Width) and Thickness of scaled pieces by raw material and across the four chronological phases.
+Error bars represent standard deviations."}
+
+area <- mean_plot(techno_data, "Chronology", "Area", "RawMaterial")
+thickness <- mean_plot(techno_data, "Chronology", "Thickness", "RawMaterial")
+
+meanplots <- ggarrange(area, thickness, ncol = 1, nrow = 2, common.legend = TRUE, legend = "right")
+
+meanplots
+
+```
+
+\newline
 
@@
 In every assemblage quartz pieces were exclusively used in one single axis, exhibiting only two damaged platforms.
-On the other end, some (3,38%) chert and chalcedony artifacts exhibit multiple functional axis, with three or four damaged platforms (Figure \@ref(fig:nplatforms)).
+On the other end, some chert and chalcedony artifacts exhibit multiple functional axis, with three or four damaged platforms (Figure \@ref(fig:nplatforms)).
 This seems to indicate different strategies of curation for coarse and fine grain raw materials.
+However, when plotted against metric data, results for the Area variable, on the one hand, reveal that pieces with four damaged platforms are amongst the largest, and, on the other, there is not a visible difference between the thickness of the pieces comprising each group.
+<!-- why is this relevant?
+-->
 
@@
 
-```{r nplatforms, fig.cap=""}
+```{r nplatforms, echo=FALSE, fig.cap=""}
 
@@
 
-As in other classes of stone tools, scaled pieces' metric attributes are impacted by both the initial blank size as well as by the intensity of their use.
-Specifically, because some scaled pieces are used in multiple axes, direct comparisons for length and width of the typological axes (see Table \@ref(tab:tech_attributes)) for definition) cannot be straighforwardly performed.
-Since the majority of the analyzed pieces presented a rectangular outline, we use the area of a rectangle (i.e.
-Length x Width) as an approximation for the overall dimensions of the artifacts.
-Area calculations revealed a maximum of `r round(max(techno_data$Area, na.rm = TRUE), digits = 2)` mm and a minimum of `r round(max(techno_data$Area, na.rm = TRUE), digits = 2)` mm.
-For thickness the maximum is `r round(max(techno_data$Thickness, na.rm = TRUE), digits = 2)` mm and the minimum is `r round(min(techno_data$Thickness, na.rm = TRUE), digits = 2)` mm.
-In general, mean values tend to be similiar between raw materials, with some differences occuring within the Proto-Solutrean and Magdalenian assemblages, most certainly as a result of the small samples analysed for each of these periods (Figure \@ref(fig:meanplots)).
-Across techno-complexes, however, no significant statistical differences were detected when using an ANOVA test for both Area (`r anova_apa(aov(Area~Chronology, techno_data), "Chronology")` and  Thickness (`r anova_apa(aov(Thickness~Chronology, techno_data), "Chronology")`).
-
-When plotted against the number of damaged platforms, results for the Area variable reveal that pieces with four damaged platforms are amongst the largest, and that there is not a visible difference between the thickness of the pieces comprising each group.
-
-\newline
-
-```{r meanplots, echo = FALSE, fig.cap="Barplots of means for Area (Length x Width) and Thickness of scaled pieces by raw material and across the four chronological phases.
-Error bars represent standard deviations."}
-
-techno_data <- techno_data %>% 
-  mutate(Area = AxisWidth*AxisLength)
-
-area <- mean_plot(techno_data, "Chronology", "Area", "RawMaterial")
-thickness <- mean_plot(techno_data, "Chronology", "Thickness", "RawMaterial")
-
-meanplots <- ggarrange(area, thickness, ncol = 1, nrow = 2, common.legend = TRUE, legend = "right")
-
-meanplots
-
-```
-
-\newline
-
 ```{r plotbyplatforms, echo = FALSE, fig.caption = ""}
@@
 
-With few exceptions, most of the morpho-functional attributes show a higher degree of variability than technological attributes, and no significant patterns could be clearly noticed.
+With few exceptions, most of the morpho-functional attributes show a higher degree of variability than technological attributes.
 In fact, only two attributes were relatively constant across all occupations: most damaged platforms present less than 45º angles; and, a large percentage of pieces with two damaged platforms present bifacial scars on one of the platforms and unifacial scars on the opposite.
@@
 
-```{r }
+\newline
 
-# Filter attributes
-morpho_data <- all_data$morpho_data
-morpho_data <- morpho_data %>% 
-  filter(DamagedPlatforms == 2)
-  select(RawMaterial, Chronology,
-         ScarDistribution_PlatformA, ScarDistribution_PlatformB,
-         ScarArrangement_PlatformA,ScarArrangement_PlatformB,
-         ScarExtention_PlatformA,ScarExtention_PlatformB,
-         ScarFacialDistribution_PlatformA, ScarFacialDistribution_PlatformB,
-         ScarEdgeDelineation_PlatformA, ScarEdgeDelineation_PlatformB,
-         Angle_PlatformA, Angle_PlatformB)
+```{r morphoattributes, echo = FALSE}
 
-var_list <- c("DamageDegree", "ScarShape", "ScarDistribution", "ScarArrangement", "ScarExtension", "ScarFacialDistribution")
-
-# Filter by chronology
-filter <- split(morpho_data_2plat, morpho_data$Chronology)
-
-# Get cross-tables by chronology using cross_tb() function
-Gravettian <- cross_tb(filter$Gravettian, "RawMaterial", var_list) 
-Gravettian <- select(Gravettian, " ", Chert, Quartz, Total) #to remove chalcedony
-
-# Recode based on 10% frequency rule
-condense(morpho_data)
-
-# Join attributes
-morpho_data_2plat <- morpho_data %>% 
-  mutate(DamageDegree = str_c(DamageDegree_PlatformA,DamageDegree_PlatformB, sep = "/")) %>% 
-  mutate(ScarShape = str_c(ScarShape_PlatformA,ScarShape_PlatformB, sep = "/")) %>%
-  mutate(ScarDistribution = str_c(ScarDistribution_PlatformA,ScarDistribution_PlatformB, sep = "/")) %>%
-  mutate(ScarArrangement = str_c(ScarArrangement_PlatformA,ScarArrangement_PlatformB, sep = "/")) %>%
-  mutate(ScarExtension = str_c(ScarExtention_PlatformA,ScarExtention_PlatformB, sep = "/")) %>%
-  mutate(ScarFacialDistribution = str_c(ScarFacialDistribution_PlatformA,ScarFacialDistribution_PlatformB, sep = "/")) %>%
-  mutate(Scar_EdgeDelineation = str_c(ScarEdgeDelineation_PlatformA,ScarEdgeDelineation_PlatformB, sep = "/")) %>%
-  mutate(Angle = str_c(Angle_PlatformA,Angle_PlatformB, sep = "/"))
+opposed_plat_morpho()
 
 ```
+
+\newline
 

### 2018-05-17 R/chi_square_functions.R

@@ -1,8 +1,10 @@
-# Filter 2 damaged platforms
+## ---- chisquare
+
+# Filter 2 damaged platforms and change platA and platB
 
 morpho_data_platA <- morpho_data %>%
   filter(DamagedPlatforms == "2" & RawMaterial != "Chalcedony") %>%
-  select(4,5,7:18)
+  select(RawMaterial:Angle_PlatformA)
 
 cnames <- colnames(morpho_data_platA)
 cnames <- cnames %>%
@@ -12,7 +14,7 @@ names(morpho_data_platA) <- cnames
 
 morpho_data_platB <- morpho_data %>%
   filter(DamagedPlatforms == "2" & RawMaterial != "Chalcedony") %>%
-  select(4,5,19:30)
+  select(RawMaterial,Width_PlatformB:Angle_PlatformB)
 
 cnames <- colnames(morpho_data_platB)
 cnames <- cnames %>%
@@ -20,15 +22,13 @@ cnames <- cnames %>%
 names(morpho_data_platB) <- cnames
 
 
-# Bind both tables into morpho_data_2plat
+# Bind both platforms into one df
 
 morpho_data_2plat <- bind_rows(morpho_data_platA, morpho_data_platB)
 morpho_data_2plat <- select_if(morpho_data_2plat, is.character)
 
 
-# Run condense() to combine all attributes with presence lower than 10%
-
-var_list <- colnames(morpho_data_2plat)
+# Run condense() to combine all attributes with presence lower than 10% into "Other" (need function!)
 
 morpho_data_2plat <- condense_to_other(morpho_data_2plat, "DamageDegree")
 morpho_data_2plat <- condense_to_other(morpho_data_2plat, "ScarShape")
@@ -39,7 +39,7 @@ morpho_data_2plat <- condense_to_other(morpho_data_2plat, "ScarFacialDistributio
 morpho_data_2plat <- condense_to_other(morpho_data_2plat, "ScarEdgeDelineation")
 morpho_data_2plat <- condense_to_other(morpho_data_2plat, "Angle")
 
-# Run condense() to combine all attributes with presence lower than 10%
+# Run condense() to remove all attributes with presence lower than 10% (need function!)
 
 morpho_data_2plat <- condense_to_NA(morpho_data_2plat, "DamageDegree")
[…]

### 2018-05-17 R/scaled_pieces_functions.R

@@ -71,19 +71,19 @@ cross_tb <- function(dataset,x,y) {
 require(tab)
 require(tidyverse)
 
+dataset <- dataset %>% mutate_if(is.character,as.factor)
 dataset <- as.data.frame(dataset)
 
   ct <- tabmulti(dataset, x, y,
                   cell = "n",
                   parenth = "col.percent",
                   p.include = FALSE,
-                  listwise.deletion = TRUE,
                   n.headings = FALSE,
                   freq.text.label = "none")
 
-  ct <- ct %>%
-   as.tibble() %>%
-   dplyr::select(" " = Variable, Chert, Quartz, Chalcedony, Total = Overall)
+  #ct <- ct %>%
+   #as.tibble() %>%
+   #dplyr::select(" " = Variable, Chert, Quartz, Chalcedony, Total = Overall)
 
 }
 
@@ -230,6 +230,30 @@ condense_to_other <- function(df, x){
   return(df)
 }
 
+####### Alternative to all columns:
+
+condense_to_other_2 <- function(df){
+
+  nm <- colnames(df)
+
+  for (i in nm){
+    m <- df %>%
+      count_(i) %>%
+      mutate(perc = n/sum(n)) %>%
+      filter(perc < 0.10)
+
+    z <- as.character(t(m[1]))
+
+    for(x in z){
+
+      df[[i]] <- ifelse(df[[i]] == x , "Other", df[[i]])
+
+    }
+  }
+  return(df)
+}
+
+
 ########################################################################################
 # RECODE TO NA BASED ON 10% FREQUENCY -----------------------------------------------
 
@@ -251,6 +275,30 @@ condense_to_NA <- function(df, x){
 }
 
 
+####### Alternative to all columns:
+
+condense_to_NA_2 <- function(df){
+
+  nm <- colnames(df)
+
+  for (i in nm){
+    m <- df %>%
+      count_(i) %>%
+      mutate(perc = n/sum(n)) %>%
+      filter(perc < 0.10)
+
+    z <- as.character(t(m[1]))
+
+    for(x in z){
+
+      df[[i]] <- ifelse(df[[i]] == x, NA, df[[i]])
+
+    }
+  }
+  return(df)
+}
+
+
 ###########################################################################
 # CHI_SQUARE --------------------------------------------------------------
 
@@ -260,6 +308,261 @@ CHI <- function(x, y) {
[…]

### 2018-05-21 analysis/paper/paper.Rmd
experimenting with MCA

@@
 While scarce, the majority of portuguese paleolithic studies [e.g.
-@almeida_terminal_2000; @goncalves_variabilite_2012] do not offer analytical interpretations for the presence of scaled pieces, simply using references that interpret them as bipolar cores for the extraction of small bladelets and chips to be supposedly used in composite tools [e.g.
+@almeida_terminal_2000; @goncalves_variabilite_2012] do not offer analytic interpretations for the presence of scaled pieces, simply using references that interpret them as bipolar cores for the extraction of small bladelets and chips to be supposedly used in composite tools [e.g.
 @zilhao_o_1997] or as intermediate pieces for working hard materials [@gibaja_tecnologiy_2007; @marreiros_as_2009].
@@
 This results in part from the lack of comprehensive studies regarding this type of artifacts at a regional scale.
-Additionally, the contexts from where most of these artifacts were recovered did not have good organic preservation and, thus, no direct association between these tools and the exploitation of hard-organic raw materials is possible, and, in face of the inexistence of any absolute dates, their precise chronological attribution is also, most of the times, unreliable.
+Additionally, the contexts from where most of these artifacts were recovered did not have good organic preservation and, thus, no direct association between these tools and the exploitation of hard-organic raw materials is possible, and, in face of the nonexistence of any absolute dates, their precise chronological attribution is also, most of the times, unreliable.
 
@@
 Following the work of de la Peña [-@de_la_pena_qualitative_2015; -@de_la_pena_sobre_2011] and Gonzalez-Urquijo & Ibanez-Estévez [-@gonzalez-urquijo_metodologialisis_1994], we expected that these attributes would be indicative of which function the artifacts had.
-For instance, according to de la Peña [-@de_la_pena_qualitative_2015; -@de_la_pena_sobre_2011], for pieces used as wedges the delineation of the damaged platforms are constantly asymetrical and only the hammered edge clearly shows the typical *écaille* retouch.
+For instance, according to de la Peña [-@de_la_pena_qualitative_2015; -@de_la_pena_sobre_2011], for pieces used as wedges the delineation of the damaged platforms are constantly asymmetrical and only the hammered edge clearly shows the typical *écaille* retouch.
 Furthermore, these pieces would have irregular shapes, variable scar size, and irregular scar distribution.
 On the other hand, pieces used as bipolar cores would have squared or rectangular shapes, symmetric straight damaged platforms, and a higher frequency of scars on the hammered edge than on the opposed edge.
-The addition of other attributes expanding on the work by Gonzalez-Urquijo & Ibanez-Estévez [-@gonzalez-urquijo_metodologialisis_1994] was made following the same reasoning.
+The addition of other attributes drawing upon the work by Gonzalez-Urquijo & Ibanez-Estévez [-@gonzalez-urquijo_metodologialisis_1994] was made following the same reasoning.
 We expected, thus, to be able to differentiate pieces used as wedges from pieces used as cores, as both groups would show distinct combinations of attributes.
@@
 \newline
-As previously mentioned, as far as the technological data is concerned, one of our main objectives was to characterize the choice of blanks involved in bipolar technology.
+
+As previously mentioned, concerning technological data, one of our main objectives was to characterize the choice of blanks involved in bipolar technology.
 Overall, the technological analysis revealed some trends that lasted throughout the Upper Paleolithic, in agreement with the general patterns of lithic technology recorded for the site.
@@
 
-Technological and morphological characteristics of the flake blanks present very similar patterns across time and amongst raw materials.
+Technological and morphological characteristics of the flake blanks present very similar patterns across time and among raw materials.
 The blanks sought after would have straight profiles, parallel edges, no cortical surfaces and trapezoidal or triangular shaped cross-sections.
@@
 Other characteristics of the assemblage are the low frequency of retouch (n = `r techno_data %>% filter(RetouchedBlank == "Yes") %>% summarise(n())`), of fire alterations (n = `r techno_data %>% filter(Fire != "No_Traces") %>% summarise(n())`), and of the presence of original striking platforms (n = `r techno_data %>% filter(ButtType != "Absent") %>% summarise(n())`).
-Still, when present, retouch was located in the lateral part of the artifacts, similar to what would define a side-scraper.
-In a very small number of cases, when striking platforms were detected, these were mostly flat.
+Still, when present, retouch is located in the lateral part of the artifacts, similar to what would define a side-scraper.
+In a very small number of cases, when striking platforms were detected, these are mostly flat.
 The absence of the original blank platforms is to be expected in this type of artifact, mostly due to the functional use of the pieces, and consequent removal of the platform, rather than an actual choice.
-The large absence of platforms may, on the other hand, be the result of the use of the technological axis as the main functional axis.
+The large absence of platforms may, also, be the result of the use of the technological axis as the main functional axis.
 In fact, when identifiable, technological and typological axes coincide in `r inline_perc(techno_data, AxisCoincident, "Yes")`% of the cases.
@@
 As in other classes of stone tools, scaled pieces' metric attributes are impacted by both the initial blank size as well as by the intensity of their use.
-Specifically, because some scaled pieces are used in multiple axes, direct comparisons for length and width of the typological axes (see Table \@ref(tab:tech_attributes)) for definition) cannot be straighforwardly performed.
+Specifically, because some scaled pieces are used in multiple axes, direct comparisons for length and width of the typological axes (see Table \@ref(tab:tech_attributes)) for definition) cannot be straightforwardly performed.
 Since the majority of the analyzed pieces presented a rectangular outline, we use the area of a rectangle (i.e.
@@
 For thickness the maximum is `r round(max(techno_data$Thickness, na.rm = TRUE), digits = 2)` mm and the minimum is `r round(min(techno_data$Thickness, na.rm = TRUE), digits = 2)` mm.
-In general, mean values tend to be similiar between raw materials, with some differences occuring within the Proto-Solutrean and Magdalenian assemblages, most certainly as a result of the small samples analysed for each of these periods (Figure \@ref(fig:meanplots)).
-Across techno-complexes, however, no significant statistical differences were detected when using an ANOVA test for both the Area ($F (`r anova.test.area[1]`, `r anova.test.area[2]`) = `r anova.test.area[3]`$, $p = `r anova.test.area[4]`)$ and Thickness ($F (`r anova.test.thickness[1]`, `r anova.test.thickness[2]`) = `r anova.test.thickness[3]`$, $p = `r anova.test.thickness[4]`)$ variables.
+In general, mean values tend to be similar between raw materials, with some differences occurring within the Proto-Solutrean and Magdalenian assemblages, most certainly as a result of the small samples analysed for each of these periods (Figure \@ref(fig:meanplots)).
+Across techno-complexes, however, no significant statistical differences were detected when using an ANOVA test for both Area ($F (`r anova.test.area[1]`, `r anova.test.area[2]`) = `r anova.test.area[3]`$, $p = `r anova.test.area[4]`)$ and Thickness ($F (`r anova.test.thickness[1]`, `r anova.test.thickness[2]`) = `r anova.test.thickness[3]`$, $p = `r anova.test.thickness[4]`)$.
 
@@
 In every assemblage quartz pieces were exclusively used in one single axis, exhibiting only two damaged platforms.
-On the other end, some chert and chalcedony artifacts exhibit multiple functional axis, with three or four damaged platforms (Figure \@ref(fig:nplatforms)).
-This seems to indicate different strategies of curation for coarse and fine grain raw materials.
-However, when plotted against metric data, results for the Area variable, on the one hand, reveal that pieces with four damaged platforms are amongst the largest, and, on the other, there is not a visible difference between the thickness of the pieces comprising each group.
-<!-- why is this relevant?
--->
+On the other end, a small number of chert and chalcedony artifacts (n= `r techno_data %>% filter(RawMaterial != "Quartz" & DamagedPlatforms == "3" & DamagedPlatforms == "4_or_more") %>% summarize(n())`) exhibit multiple functional axis, with three or four damaged platforms (Figure \@ref(fig:nplatforms)).
+This seems to indicate different strategies of curation for coarse and fine grain raw materials, with fine grain materials being rotated when the first used axis becomes too small and/or the edges get useless for that specific activity.
+However, when plotted against metric data, results for the Area variable reveal that pieces with four damaged platforms are among the largest in the whole assemblages, and that there is not a visible difference between the thickness of the pieces comprising each group.
+This seems to attest that the use of several axis in the same piece is not correlated with curation occuring in later phases of artifact use but instead to a probable difference in raw material performance.
 
@@
   mutate(perc=100*(count/sum(count)))
-n_axis
 
@@
   theme_bw() +
-  theme(strip.text.x = element_text(size = 14), text = element_text(size=14))
+  theme(strip.text.x = element_text(size = 12))+
+  scale_fill_manual(values=c("#00AFBB", "#E7B800", "#FC4E07", "#999999"))
 
@@
 
-With few exceptions, most of the morpho-functional attributes show a higher degree of variability than technological attributes.
-In fact, only two attributes were relatively constant across all occupations: most damaged platforms present less than 45º angles; and, a large percentage of pieces with two damaged platforms present bifacial scars on one of the platforms and unifacial scars on the opposite.
+With few exceptions, most of the morpho-functional attributes show a fairly high degree of variability.
+Similarly to what was registered for the technological attributes most of the differences occur between raw materials rather than between chronologies.
+For this reason the results presented below focus only on the overall and between raw materials variability.
+Figure \@ref(fig:morphoattributes) shows the absolute frequencies of all qualitative variables recorded for each damaged platform.
+To simplify representation, and since we actually do not know which platform was the hammered one, opposed damage platforms are grouped into pairwise categories.
+When present, category 'Other' represents the cluster of attributes whose frequency was less than 10% within each variable.
 
-On a closer inspection, we find that most of the differences within the morpho-functional variables occur between raw materials rather than between chronologies.
+In three (Scar Arrangement, Extension and Edge Delineation) of the six represented variables the category 'Other' is one of the most frequent (above 35%) for both quartz and chert, revealing a very high variability for the combination of attributes within each variable.
+
+For quartz artifacts the most common patterns are the combination of central/total scar distributions, both opposed platforms with angles wider than 45º, and a combination of either unifacial/unifacial or unifacial/bifacial scar distribution.
+Chert pieces, on the other hand, more typically present opposed platforms with damage occupying the whole platform width, and a weaker presence of unifacial/unifacial scar edge facial distribution.
+Still, with exception of Angle and Scar Facial Distribution, the overall trend for the morpho-functional variables is one of high variability, with a very large set of combinations appearing at very low frequencies.
+
+To identify possible patterns of association between each one of the qualitative variables used in our analysis \@ref(fig:morphoattributes), a Multiple Correspondence Analysis was run using the FactoMineR package.
+Results presented in Figure \@ref(fig:mca).
+
+The first two dimensions of the MCA express 21.35% of the total dataset inertia, meaning that 21.35% of the individuals (or variables) cloud total variability is explained by the plane combining the first two dimensions.
+This is a rather small percentage but it is greater than the reference value that equals 18.92%.
+The variability explained by this plane is thus significant.
 

### 2018-05-21 R/scaled_pieces_functions.R

@@ -138,11 +138,11 @@ box_plot <- function(df, x_var, y_var, x_label, y_label){
     geom_boxplot(outlier.shape = NA, fill = "gray")
 
   b_plot <- b_plot +
-    geom_jitter(position=position_jitter(width=.2, height=0), aes(color = RawMaterial), size = 4) +
+    geom_jitter(position=position_jitter(width=.2, height=0), aes(color = RawMaterial), size = 2) +
     scale_color_manual(values = c("#00AFBB", "#E7B800", "#FC4E07"), name = "Raw Material") +
     labs(x = x_label, y = y_label)+
     scale_fill_discrete(guide=FALSE) +
-    theme(text = element_text(size = 20), axis.text = element_text(size = 20))
+    the
 
   return(b_plot)
 }
@@ -504,44 +504,99 @@ opposed_plat_morpho <- function(){
   morpho_data_opposed_platforms <- condense_to_other_2(morpho_data_opposed_platforms)
   morpho_data_opposed_platforms <- condense_to_NA_2(morpho_data_opposed_platforms)
 
+
   # Plot all variables
 
-  Dist <- count(morpho_data_opposed_platforms, RawMaterial, ScarDistribution)
-  Dist <- ggbarplot(Dist, "ScarDistribution", "n",
+  dist_perc <- morpho_data_opposed_platforms %>%
+    group_by(RawMaterial, ScarDistribution) %>%
+    tally() %>%
+    group_by(RawMaterial) %>%
+    mutate(pct = (n / sum(n))*100)
+
+  dist <- ggbarplot(dist_perc, x = "ScarDistribution", y = "pct",
+                    position = position_dodge(0.9),
                     fill = "RawMaterial", color = "RawMaterial",
-                    palette = c("#00AFBB", "#E7B800"), orientation = "horiz")
+                    palette = c("#00AFBB", "#E7B800"),
+                    orientation = "horiz",
+                    xlab = FALSE,
+                    ylab = FALSE)
 
 
-  Arr <- count(morpho_data_opposed_platforms, RawMaterial, ScarArrangement)
-  Arr <- ggbarplot(Arr, "ScarArrangement", "n",
-                   fill = "RawMaterial", color = "RawMaterial",
-                   palette = c("#00AFBB", "#E7B800"), orientation = "horiz")
 
+  arr_perc <- morpho_data_opposed_platforms %>%
+    group_by(RawMaterial, ScarArrangement) %>%
+    tally() %>%
[…]

### 2018-06-07 analysis/paper/paper.Rmd
mca changes

@@
 
-#library(ScaledPiecesVB) # Or use devtools::load_all('.', quiet = T) if your code is in script files, rather than as functions in the `/R` diretory
 ```
@@
 library(tidyverse)
+library(broom)
 library(FactoMineR)
@@
 library(tab)
+library(sjstats)
+library(powerAnalysis)
+library(epitools)
+#library(ScaledPiecesVB)
 
@@
 
-Beyond the classification debate, the biggest issue with this type of tools comes from a functional standpoint, a problem that has also been quite debated [e.g.
+Beyond the classification debate, the biggest issue with this type of tools comes from a functional standpoint, a problem that has also been largely debated [e.g.
 @binford_indian_1963; @de_la_pena_sobre_2011; @flood_moth_1980; @igreja_functional_2013; @leblanc_wedges_1992; @lucas_les_2004; @shott_bipolar_1999; @tixier_typologie_1963].
@@
 While scarce, the majority of portuguese paleolithic studies [e.g.
-@almeida_terminal_2000; @goncalves_variabilite_2012] do not offer analytic interpretations for the presence of scaled pieces, simply using references that interpret them as bipolar cores for the extraction of small bladelets and chips to be supposedly used in composite tools [e.g.
+@almeida_terminal_2000; @goncalves_variabilite_2012] have not presented, so far, any context-specific interpretations for the presence of scaled pieces.
+Most references broadly interpret these as bipolar cores for the extraction of small bladelets and chips to be  used in composite tools [e.g.
 @zilhao_o_1997] or as intermediate pieces for working hard materials [@gibaja_tecnologiy_2007; @marreiros_as_2009].
@@
 While we agree that some of the elements might have been used as cores, these interpretations seem overly simplistic, and the authors seem to have little to no evidence to back their claims other than the inverse relationship mentioned above.
-This results in part from the lack of comprehensive studies regarding this type of artifacts at a regional scale.
+This results, in part, from the lack of comprehensive studies regarding this type of artifacts at a regional scale.
 Additionally, the contexts from where most of these artifacts were recovered did not have good organic preservation and, thus, no direct association between these tools and the exploitation of hard-organic raw materials is possible, and, in face of the nonexistence of any absolute dates, their precise chronological attribution is also, most of the times, unreliable.
@@
 Archaeological deposits occupy an estimated area of over 10 000 sq.
-meters across a stepped slope marked at the top by a 10 meter-high limestone cliff face.
+meters across a stepped slope marked at the top by a 10 meter-high limestone cliff face [@bicho_ledge:_2012; @bicho_paleolithique_2010].
 
@@
 
-source('~/R_DATA/compendiums/ScaledPiecesVB/R/scaled_pieces_functions.R')
-
 Map()
@@
 Vale Boi’s lithic studies have revealed a general tendency for stable technological and functional patterns throughout the Upper Paleolithic [e.g.
-@bicho_ledge:_2012; @bradtmoller_lithic_2016; @cascalheira_cross-scale_2017; @cascalheira_influencia_2013; @cascalheira_vale_2012; @gibaja_funcion_2011; @marreiros_as_2009; @marreiros_lithic_2013; @marreiros_lithic_2015].
+@bicho_ledge:_2012; @bradtmoller_lithic_2016; @cascalheira_cross-scale_2017; @cascalheira_influencia_2013; @cascalheira_vale_2012; @gibaja_funcion_2011; @marreiros_as_2009; @marreiros_lithic_2013; @marreiros_lithic_2015; @marreiros_lithic_2018].
 This can be explained, partially, by raw material availability.
@@
 
-## Reproducibility and open source materials
+## Analysis, reproducibility and open source materials
 
@@
 Still, when present, retouch is located in the lateral part of the artifacts, similar to what would define a side-scraper.
-In a very small number of cases, when striking platforms were detected, these are mostly flat.
+In a very small number of cases, when striking platforms are present, these are mostly flat.
 The absence of the original blank platforms is to be expected in this type of artifact, mostly due to the functional use of the pieces, and consequent removal of the platform, rather than an actual choice.
-The large absence of platforms may, also, be the result of the use of the technological axis as the main functional axis.
+The large absence of platforms may, also, be the result of the use of the technological axis as main functional axis.
 In fact, when identifiable, technological and typological axes coincide in `r inline_perc(techno_data, AxisCoincident, "Yes")`% of the cases.
-The longitudinal section shows a large variability of shapes, independent of raw material or chronology.
+The longitudinal sections show a large variability of shapes, independent of raw materials or chronologies.
 Similarly, the dorsal pattern of previous removals was rarely identified, although this, like with the case of platform absence, might occur due to the functional stigmas and be dependent on the intensity of use for each artifact.
@@
 
-```{r mestric_statistics, echo = FALSE}
+```{r metric_statistics, echo = FALSE}
 
@@
 
-anova.test.area <- anova.test(Area ~ Chronology, techno_data)
+aov_area <- anova.test(techno_data, "Area")
+cohens_area <- cohens.test(aov_area)
+aov_area <- tidy(aov_area)
 
-anova.test.thickness <- anova.test(Thickness ~ Chronology, techno_data)
+aov_thickness <- anova.test(techno_data, "Thickness")
+cohens_thickness <- cohens.test(aov_thickness)
+aov_thickness <-  tidy(aov_thickness)
 
@@
 In general, mean values tend to be similar between raw materials, with some differences occurring within the Proto-Solutrean and Magdalenian assemblages, most certainly as a result of the small samples analysed for each of these periods (Figure \@ref(fig:meanplots)).
-Across techno-complexes, however, no significant statistical differences were detected when using an ANOVA test for both Area ($F (`r anova.test.area[1]`, `r anova.test.area[2]`) = `r anova.test.area[3]`$, $p = `r anova.test.area[4]`)$ and Thickness ($F (`r anova.test.thickness[1]`, `r anova.test.thickness[2]`) = `r anova.test.thickness[3]`$, $p = `r anova.test.thickness[4]`)$.
+Across techno-complexes, however, no significant statistical differences were detected when using an ANOVA test for both Area (*F* (`r aov_area$df[1]`, `r aov_area$df[2]`) = `r aov_area$statistic[1]`, *p* = `r aov_area$p.value[1]`, *d* = `r cohens_area$cohens.f`), and Thickness (*F* (`r aov_thickness$df[1]`, `r aov_thickness$df[2]`) = `r aov_thickness$statistic[1]`, *p* = `r aov_thickness$p.value[1]`, *d* = `r cohens_thickness$cohens.f`).
 
@@
 In every assemblage quartz pieces were exclusively used in one single axis, exhibiting only two damaged platforms.
-On the other end, a small number of chert and chalcedony artifacts (n= `r techno_data %>% filter(RawMaterial != "Quartz" & DamagedPlatforms == "3" & DamagedPlatforms == "4_or_more") %>% summarize(n())`) exhibit multiple functional axis, with three or four damaged platforms (Figure \@ref(fig:nplatforms)).
+On the other end, a small number of chert and chalcedony artifacts (*N* = `r techno_data %>% filter(RawMaterial != "Quartz" & DamagedPlatforms == "3" | DamagedPlatforms == "4_or_more") %>% summarize(n())`) exhibit multiple functional axis, with three or four damaged platforms (Figure \@ref(fig:nplatforms)).
 This seems to indicate different strategies of curation for coarse and fine grain raw materials, with fine grain materials being rotated when the first used axis becomes too small and/or the edges get useless for that specific activity.
@@
 
-ggplot(n_axis, aes(x = factor(RawMaterial), y = perc, fill = factor(DamagedPlatforms))) +
-  geom_bar(stat="identity", width = 0.7) +
-  labs(x = "RawMaterial", y = "%", fill = "Number of Damaged Platforms") +
-  facet_wrap(~Chronology) +
-  theme_bw() +
-  theme(strip.text.x = element_text(size = 12))+
-  scale_fill_manual(values=c("#00AFBB", "#E7B800", "#FC4E07", "#999999"))
+d_platforms <- ggbarplot(n_axis, x = "RawMaterial", y = "perc",
+          color = "DamagedPlatforms", fill = "DamagedPlatforms",
+          palette = c("#00AFBB", "#E7B800", "#FC4E07", "#999999"),
+          facet.by = "Chronology", xlab = "", ylab = "%")
+
+ggpar(d_platforms, legend = "right", legend.title = "Number of Damaged Platforms")
 
@@
 
-area_by_platforms <- box_plot(techno_data, "DamagedPlatforms", "Area", "", "Area (Length * Width)")
+area_by_platforms <- box_plot(techno_data, "DamagedPlatforms", "Area", "Number of Damaged Platforms", "Area (Length * Width)")
 thickness_by_platforms <- box_plot(techno_data, "DamagedPlatforms", "Thickness", "Number of Damaged Platforms", "Thickness (mm)")
@@
 \newline
+```{r chisquare, echo = FALSE}
+
+#Select and arrange morpho-functional variables for each pair of opposed platforms
+morpho_data_opposed_platforms <- opposed_plat_morpho()
+
+#Run chi-square and effect size test for each variable
+chi_distribution <- CHI(morpho_data_opposed_platforms$ScarDistribution,morpho_data_opposed_platforms$RawMaterial)
+
+chi_arrangement <- CHI(morpho_data_opposed_platforms$ScarArrangement,morpho_data_opposed_platforms$RawMaterial)
+
+chi_extension <- CHI(morpho_data_opposed_platforms$ScarExtension,morpho_data_opposed_platforms$RawMaterial)
+
+chi_delineation <- CHI(morpho_data_opposed_platforms$ScarEdgeDelineation,morpho_data_opposed_platforms$RawMaterial)
+
+chi_faciality <- CHI(morpho_data_opposed_platforms$ScarFaciality,morpho_data_opposed_platforms$RawMaterial)
+
+chi_angle <- CHI(morpho_data_opposed_platforms$Angle,morpho_data_opposed_platforms$RawMaterial)
+
+```
+
+\newline
 
 With few exceptions, most of the morpho-functional attributes show a fairly high degree of variability.
-Similarly to what was registered for the technological attributes most of the differences occur between raw materials rather than between chronologies.
-For this reason the results presented below focus only on the overall and between raw materials variability.
-Figure \@ref(fig:morphoattributes) shows the absolute frequencies of all qualitative variables recorded for each damaged platform.
-To simplify representation, and since we actually do not know which platform was the hammered one, opposed damage platforms are grouped into pairwise categories.
-When present, category 'Other' represents the cluster of attributes whose frequency was less than 10% within each variable.
+Similarly to what was registered for the technological attributes, most of the differences occur between raw materials rather than between chronologies.
+For this reason the results presented below focus only on overall and between raw materials variability.
+Figure \@ref(fig:morphoattributes) shows the relative frequencies of all qualitative variables recorded for each damaged platform.
+To simplify representation, and to avoid wrong comparisons between hammered and active platforms, opposed damage platforms are grouped into pairwise categories.
+When present, category 'Other' represents the cluster of attributes which frequency was less than 10% within each variable.
 
 In three (Scar Arrangement, Extension and Edge Delineation) of the six represented variables the category 'Other' is one of the most frequent (above 35%) for both quartz and chert, revealing a very high variability for the combination of attributes within each variable.
+A chi-square test with modified alpha level (Bonferroni correction) to adjust for multiple testing and reduce type-I error, shows significant differences in the quartz and chert Scar Arrangement categories (*X^2^* [`r chi_arrangement$chi_test$parameter`, *N* = `r sum(chi_arrangement$chi_test$observed)`] = `r  round(chi_arrangement$chi_test$statistic, 2)`, *p* = `r round(chi_arrangement$chi_test$p.value, 3)`, *phi* = `r round(chi_arrangement$chi_size_effect$phi, 2)`), and in Scar Faciality categories (*X^2^* [`r chi_faciality$chi_test$parameter`, *N* = `r sum(chi_faciality$chi_test$observed)`] = `r  round(chi_faciality$chi_test$statistic,2)`, *p* = `r round(chi_faciality$chi_test$p.value, 3)`, *phi* = `r round(chi_faciality$chi_size_effect$phi, 2)`).
+The effect size statistic (Phi) suggests, however, a low practical significance for Scar Arrangement and a large pratical significance for Scar Faciality.
 
-For quartz artifacts the most common patterns are the combination of central/total scar distributions, both opposed platforms with angles wider than 45º, and a combination of either unifacial/unifacial or unifacial/bifacial scar distribution.
+For quartz artifacts the most common patterns are the combination of central/total scar distributions, both platforms with angles wider than 45º, and a combination of either unifacial/unifacial or unifacial/bifacial scar distribution.
 Chert pieces, on the other hand, more typically present opposed platforms with damage occupying the whole platform width, and a weaker presence of unifacial/unifacial scar edge facial distribution.
 Still, with exception of Angle and Scar Facial Distribution, the overall trend for the morpho-functional variables is one of high variability, with a very large set of combinations appearing at very low frequencies.
-
-To identify possible patterns of association between each one of the qualitative variables used in our analysis \@ref(fig:morphoattributes), a Multiple Correspondence Analysis was run using the FactoMineR package.
-Results presented in Figure \@ref(fig:mca).
-
-The first two dimensions of the MCA express 21.35% of the total dataset inertia, meaning that 21.35% of the individuals (or variables) cloud total variability is explained by the plane combining the first two dimensions.
-This is a rather small percentage but it is greater than the reference value that equals 18.92%.
-The variability explained by this plane is thus significant.
 
@@
 
-opposed_plat_morpho()
+#PLot categorical morpho-function variables
+
+plot_morpho_var(morpho_data_opposed_platforms)
+
+```
+
+\newline
+
+To identify possible patterns of association among the qualitative variables used in our analysis \@ref(fig:morphoattributes), a Multiple Correspondence Analysis was performed.
+Results are presented in Figures \@ref(fig:mca).
+In this analysis raw material was used as a qualitative supplementary variable.
+
+The first two dimensions of the MCA express 20.92% of the total dataset inertia, meaning that only that percentage of the total variability is explained by the plane combining the first two dimensions.
+An inspection to the screeplot presented in Figure \@ref(fir:screeplot) suggests to restrict the analysis to the description of the first 4 dimensions.
+These dimensions present an amount of inertia slightly greater than that obtained by the 0.95-quantile of random distributions (37.74% against 34.37%).
+Still, it is a rather small percentage, somehow attesting the high variability suggested by our interpretation of Figure \@ref(fig:morphoattributes), and suggesting that patterns identified by previous studies [e.g.
+@de_la_pena_qualitative_2015; @de_la_pena_sobre_2011], in which certain combination of attributes were used to discriminate scaled pieces functionalities, are difficult to apply to the assemblage used in this study.
+
+The high variability of the assemblage is also well noticeable in the biplots presented in Figure \@ref(fig:biplot).
+
+\newline
+
+```{r mca, echo = FALSE}
+
+# Multiple Correspondence Analysis
+
+mca.rawmaterial()
+
+```
+
+\newline
+
+```{r screeplot, echo = FALSE}
+
+# Screeplot
+fviz_screeplot(morpho_data_mca, addlabels = TRUE, ylim = c(0, 15), barfill = "#00AFBB", barcolor = "#00AFBB")
+
+```
+
+\newline
+
+```{r biplot, echo = FALSE}
+
+# Biplot for 4 main MCA dimensions
+
+fviz_screeplot(morpho_data_mca, addlabels = TRUE, ylim = c(0, 15), barfill = "#00AFBB", barcolor = "#00AFBB")
 

### 2018-06-07 R/chi_square_functions.R

@@ -1,83 +1,0 @@
-## ---- chisquare
-
-# Filter 2 damaged platforms and change platA and platB
-
-morpho_data_platA <- morpho_data %>%
-  filter(DamagedPlatforms == "2" & RawMaterial != "Chalcedony") %>%
-  select(RawMaterial:Angle_PlatformA)
-
-cnames <- colnames(morpho_data_platA)
-cnames <- cnames %>%
-  str_replace("_PlatformA", "")
-names(morpho_data_platA) <- cnames
-
-
-morpho_data_platB <- morpho_data %>%
-  filter(DamagedPlatforms == "2" & RawMaterial != "Chalcedony") %>%
-  select(RawMaterial,Width_PlatformB:Angle_PlatformB)
-
-cnames <- colnames(morpho_data_platB)
-cnames <- cnames %>%
-  str_replace("_PlatformB", "")
-names(morpho_data_platB) <- cnames
-
-
-# Bind both platforms into one df
-
-morpho_data_2plat <- bind_rows(morpho_data_platA, morpho_data_platB)
-morpho_data_2plat <- select_if(morpho_data_2plat, is.character)
-
-
-# Run condense() to combine all attributes with presence lower than 10% into "Other" (need function!)
-
-morpho_data_2plat <- condense_to_other(morpho_data_2plat, "DamageDegree")
-morpho_data_2plat <- condense_to_other(morpho_data_2plat, "ScarShape")
-morpho_data_2plat <- condense_to_other(morpho_data_2plat, "ScarDistribution")
-morpho_data_2plat <- condense_to_other(morpho_data_2plat, "ScarArrangement")
-morpho_data_2plat <- condense_to_other(morpho_data_2plat, "ScarExtention")
-morpho_data_2plat <- condense_to_other(morpho_data_2plat, "ScarFacialDistribution")
-morpho_data_2plat <- condense_to_other(morpho_data_2plat, "ScarEdgeDelineation")
-morpho_data_2plat <- condense_to_other(morpho_data_2plat, "Angle")
-
-# Run condense() to remove all attributes with presence lower than 10% (need function!)
-
-morpho_data_2plat <- condense_to_NA(morpho_data_2plat, "DamageDegree")
-morpho_data_2plat <- condense_to_NA(morpho_data_2plat, "ScarShape")
[…]

### 2018-06-07 R/scaled_pieces_functions.R

@@ -118,8 +118,6 @@ mean_tb <- function(dataset,x,y) {
 
 mean_plot <- function(dataset, x, y, z){
 
-require(ggpubr)
-
   ggbarplot(dataset, x = x, y = y,
             add = c("mean_sd", "jitter"), size = 1,
             color = z, palette = c("#00AFBB", "#E7B800", "#FC4E07"),
@@ -135,14 +133,14 @@ require(ggpubr)
 
 box_plot <- function(df, x_var, y_var, x_label, y_label){
   b_plot<-ggplot(df,aes_string(x= x_var, y=y_var)) +
-    geom_boxplot(outlier.shape = NA, fill = "gray")
+    geom_boxplot(outlier.shape = NA)
 
   b_plot <- b_plot +
     geom_jitter(position=position_jitter(width=.2, height=0), aes(color = RawMaterial), size = 2) +
     scale_color_manual(values = c("#00AFBB", "#E7B800", "#FC4E07"), name = "Raw Material") +
     labs(x = x_label, y = y_label)+
-    scale_fill_discrete(guide=FALSE) +
-    the
+    scale_fill_discrete(guide=FALSE)+
+    theme_classic()
 
   return(b_plot)
 }
@@ -232,7 +230,7 @@ condense_to_other <- function(df, x){
 
 ####### Alternative to all columns:
 
-condense_to_other_2 <- function(df){
+condense_to_other_all <- function(df){
 
   nm <- colnames(df)
 
@@ -277,7 +275,7 @@ condense_to_NA <- function(df, x){
 
 ####### Alternative to all columns:
 
-condense_to_NA_2 <- function(df){
+condense_to_NA_all <- function(df){
 
   nm <- colnames(df)
 
@@ -300,133 +298,43 @@ condense_to_NA_2 <- function(df){
 
 
 ###########################################################################
-# CHI_SQUARE --------------------------------------------------------------
+# CHI_SQUARE, EFFECT SIZE AND RATIO----------------------------------------
+
 
 CHI <- function(x, y) {
 
-  test <- chisq.test(table(x, y))
-  result <- c(test$p.value)
+  chi_test <- chisq.test(table(x, y))
+  chi_size_effect <- ES.chisq.assoc(ct=table(x, y))
 
-  return(result)
-
-}
-
[…]

### 2018-06-08 analysis/paper/paper.Rmd
unused functions deleted

@@
 
-source('~/R_DATA/compendiums/ScaledPiecesVB/R/scaled_pieces_functions.R')
+source('~/R_DATA/compendiums/ScaledPiecesVB/R/functions.R')
 
@@
 
-all_data <- read_data()
+all_data <- read.data()
 techno_data <- all_data$techno_data
@@
 
-assemb <- cross_tb(techno_data, "RawMaterial", var_list)
+assemb <- cross.tb(techno_data, "RawMaterial", var_list)
 
@@
 The large absence of platforms may, also, be the result of the use of the technological axis as main functional axis.
-In fact, when identifiable, technological and typological axes coincide in `r inline_perc(techno_data, AxisCoincident, "Yes")`% of the cases.
+In fact, when identifiable, technological and typological axes coincide in `r inline.perc(techno_data, AxisCoincident, "Yes")`% of the cases.
 The longitudinal sections show a large variability of shapes, independent of raw materials or chronologies.
@@
 # Get cross-tables by chronology using cross_tb() function
-Gravettian <- cross_tb(filter$Gravettian, "RawMaterial", var_list) 
+Gravettian <- cross.tb(filter$Gravettian, "RawMaterial", var_list) 
 Gravettian <- Gravettian %>% 
@@
 
-Solutrean <- cross_tb(filter$Solutrean, "RawMaterial", var_list)
+Solutrean <- cross.tb(filter$Solutrean, "RawMaterial", var_list)
 Solutrean <- Solutrean %>% 
@@
 
-`Proto-Solutrean` <- cross_tb(filter$`Proto-Solutrean`, "RawMaterial", var_list)
+`Proto-Solutrean` <- cross.tb(filter$`Proto-Solutrean`, "RawMaterial", var_list)
 `Proto-Solutrean` <- `Proto-Solutrean` %>% 
@@
 
-Magdalenian <- cross_tb(filter$Magdalenian, "RawMaterial", var_list)
+Magdalenian <- cross.tb(filter$Magdalenian, "RawMaterial", var_list)
 Magdalenian <- Magdalenian %>% 
@@
 
-area <- mean_plot(techno_data, "Chronology", "Area", "RawMaterial")
-thickness <- mean_plot(techno_data, "Chronology", "Thickness", "RawMaterial")
+area <- mean.plot(techno_data, "Chronology", "Area", "RawMaterial")
+thickness <- mean.plot(techno_data, "Chronology", "Thickness", "RawMaterial")
 
@@
 
-area_by_platforms <- box_plot(techno_data, "DamagedPlatforms", "Area", "Number of Damaged Platforms", "Area (Length * Width)")
-thickness_by_platforms <- box_plot(techno_data, "DamagedPlatforms", "Thickness", "Number of Damaged Platforms", "Thickness (mm)")
+area_by_platforms <- box.plot(techno_data, "DamagedPlatforms", "Area", "Number of Damaged Platforms", "Area (Length * Width)")
+thickness_by_platforms <- box.plot(techno_data, "DamagedPlatforms", "Thickness", "Number of Damaged Platforms", "Thickness (mm)")
 
@@
 #Select and arrange morpho-functional variables for each pair of opposed platforms
-morpho_data_opposed_platforms <- opposed_plat_morpho()
+morpho_data_opposed_platforms <- opposed.plat.morpho()
 
@@
 
-plot_morpho_var(morpho_data_opposed_platforms)
+plot.morpho.var(morpho_data_opposed_platforms)
 
@@
 
-To identify possible patterns of association among the qualitative variables used in our analysis \@ref(fig:morphoattributes), a Multiple Correspondence Analysis was performed.
+To identify possible patterns of association among the qualitative variables used in our analysis \@ref(fig:morphoattributes), a Multiple Correspondence Analysis (MCA) was performed.
 Results are presented in Figures \@ref(fig:mca).
-In this analysis raw material was used as a qualitative supplementary variable.
+In this analysis, raw material was used as a qualitative supplementary variable.
 
@@
 
-mca.rawmaterial()
+morpho_data_mca <- mca.scaled.pieces()
 

### 2018-06-08 R/functions.R

@@ -2,7 +2,7 @@
 ###########################################################################
 # READ DATA ----------------------------------------------------------------
 
-read_data <- function(){
+read.data <- function(){
   #tech_attributes <- readcsv("../data/raw_data/technology_attributes.csv")
   #morpho_attributes <- read.csv("../data/raw_data/morpho_attributes.csv")
   techno_data <- read_csv("../data/raw_data/technology_dataset.csv")
@@ -12,7 +12,6 @@ read_data <- function(){
               techno_data = techno_data))
 
 }
-
 
 ###########################################################################
 # MAPS --------------------------------------------------------------------
@@ -64,9 +63,7 @@ Map <- function(){
 ###########################################################################
 # CROSS TABLES ------------------------------------------------------------
 
-
-
-cross_tb <- function(dataset,x,y) {
+cross.tb <- function(dataset,x,y) {
 
 require(tab)
 require(tidyverse)
@@ -87,36 +84,10 @@ dataset <- as.data.frame(dataset)
 
 }
 
-
-###########################################################################
-# MEAN TABLE --------------------------------------------------------------
-
-
-mean_tb <- function(dataset,x,y) {
-
-  require(tab)
-  require(tidyverse)
-
-  mt <- tabmulti(dataset, x, y,
-                 cell = "n",
-                 parenth = "col.percent",
-                 p.include = TRUE,
-                 n.headings = FALSE,
-                 freq.text.label = "none",
-                 means.text.label = "none")
-
- # mt <- mt %>%
-  #  as.tibble() %>%
-   # dplyr::select(" " = Variable, Chert, Quartz, Chalcedony, Overall)
-
-}
-
-
-
 ###########################################################################
 # MEAN PLOTS --------------------------------------------------------------
 
-mean_plot <- function(dataset, x, y, z){
+mean.plot <- function(dataset, x, y, z){
 
   ggbarplot(dataset, x = x, y = y,
             add = c("mean_sd", "jitter"), size = 1,
@@ -131,7 +102,7 @@ mean_plot <- function(dataset, x, y, z){
 ###########################################################################
 # BOXPLOT WITH JITTER -----------------------------------------------------
 
[…]

### 2018-07-31 analysis/paper/paper.Rmd
Discussion added

@@
 Their distinction, however, has not always been consensual [@hayden_confusion_1980].
-In fact, the definition of scaled piece has suffered significant changes ever since its initial identification in the early 20th century and, to this day, there still seems to be no clear definition for this type of artifact.
+Besides the fact that bipolar cores are often confused or lumped with scaled pieces [@villa_neandertals_2018], the definition of scaled piece has suffered significant changes ever since its initial identification in the early 20th century and, to this day, there still seems to be no clear definition for this type of artifact.
 The first definition of scaled pieces was proposed by Bardon and Bouyssonie [-@bardon_outils_1906], describing them as a result of bipolar knapping through direct percussion, with the “core” rested in a hard surface, originating splintering in both ends of the tool.
@@
 @diez-martin_were_2009; @igreja_functional_2013; @langejans_middle_2012; @white_fabricators_1968], evidence of bipolar technologies are quite ubiquitous in European Upper Paleolithic contexts.
-In many sites, bipolar elements are classified as scaled pieces, a morphotype that has been putatively considered as one the most common lithic tools in Upper Paleolithic assemblages [@de_la_pena_qualitative_2015], but different functional connotations for this type of artifacts have been presented in several western european Upper Paleolithic studies [e.g.
+In many sites, bipolar elements classified as scaled pieces have been associated with different functionalities [see e.g.
 @de_la_pena_sobre_2011; @sano_functional_2012; @zilhao_o_1997].
@@
 ## Scaled pieces attribute analysis
+
+We define scaled pieces as artifacts of variable size and morphology, showing traces of crushing and splitering of edges at opposite ends.
+These traces are caused by direct percussion at one end and subsequent crushing of the other because it is rested on a hard surface, and can be bifacially or unifacially distributed.
+As will become clear from our results some of the analyzed artifacts did not present splitering/crushing traces in opposed platforms, but we still classified them as scaled pieces.
+This detail has been previously referred by Villa et al.
+[-@villa_neandertals_2018] who noted that some edges may present a flat (instead of intensively shattered) platforms.
 
@@
 
-#Discussion
+#Discussion and conclusions
 
-#Conclusion
+Bipolar technology clearly had an important role on the first modern humans in Western Iberia.
+As mentioned by Shot and Tostevin [-@shott_diversity_2015], a rise in bipolar technology frequency may be related to human mobility patterns and even evolutionary trends [@duke_weaving_2015].
+For the first modern humans, as well as their descendants, this technology would have played a major role in resource exploitation and diversification.
+In Vale Boi, bipolar technology shows up in the archaeological record connected to evidence related to the intensification and diversification of resource exploration such as the above-mentioned grease-rendering techniques.
+These have been shown by the constant presence of impact fractures in ungulate bones, low percentage of bone areas related to higher amounts of grease, and the constant correlation between red deer bone fragments with the amount of bone marrow and grease contained in them.
+Other functional indicators are both the constant presence of thermally altered quartz fragments that might be linked to stone boiling activities, and a large number of greywacke slabs with impact marks that were used as anvils.
+
+In Table 6 we present current data based on previous work on bipolar technology in Westernmost Iberia for the Upper Paleolithic.
+While the number of artifacts is not very high, bipolar technology is present in a significant number of sites.
+Even with the considerable number of sites represented, it is clear the discrepancy in counts when comparing the presence of scaled pieces and bipolar cores, which by itself could be indicative of some sort of functional pattern.
+It is clear, however, that scaled pieces show on average, a higher representation in Vale Boi’s assemblages than on most other sites.
+Further, of the three bipolar cores in all sites, 2 of these come from Vale Boi.
+While currently there are no data on the Typological Index (TI) of the Slope’s scaled pieces, since the rest of the assemblage has not yet been studied, it is fairly safe to assume that the TI would have a particularly high value.
+Bipolar technology is quite common in Western Iberian Upper Paleolithic sites and interpretations made are solely based on a single reference with no statistical or analytical evidence.
+We argue that all factors mentioned in this study must be considered while interpreting each site in this region.
+While in Vale Boi there is clear evidence that the use of these artifacts would not be for bipolar knapping, in other sites the picture may be different, although these interpretations should be in the future backed by some sort of analytical evidence.
+As our results show, the low presence of bipolar cores is quite evident in these sites, while scaled pieces are more often present.
+It can also be argued that the number of pieces in the overall picture is also low, although presenting an accurate picture for each site since most of these sites have been extensively studied.
+
+Our results show that in Vale Boi, scaled pieces have several patterns that stay consistent across all occupations.
+One relevant trend is the fact that while quartz pieces were only explored in a single axis, chert pieces were, sometimes, explored in multiple axes.
+One possible, rather improbable, reason for these patterns might be the fact that quartz would be more easily available than chert.
+Another, more likely, possibility is due to the material properties of each raw material.
+While with quartz it may be possible to continuously use a single axis in a piece, reducing it in only one axis without losing its efficiency, with chert it may be necessary to rotate it in order to continue obtaining usable edges.
+On the other hand, the higher number of axis could simply suggest that chert was differently managed and economized.
+Nevertheless, the simple fact that most of the pieces with multiple axes are on average larger than the rest suggests that this hypothesis is likely incorrect.
+From an economic standpoint, the ideal would be to continuously use the same piece without any risk of loss of the features that made it ideal for this kind of use.
+Throughout the use of these pieces, they suffer successive violent blows, even the platform that rests on a surface suffers some sort of heavy impact.
+In this specific case, we argue that there is little to no control on how the piece’s features change over its lifetime use.
+
+From a paleoanthropological point of view, this type of bipolar technique would have a major role on a day to day basis.
+For carcass processing and bone marrow extraction the use of scaled pieces would bring two major advantages: (1) it would allow for a better control of bone fracture; and, in consequence, (2) it would allow a clear choice of which part of the bone can either be broken or removed.
+It can be argued that both activities can be performed with any kind of hammer.
+However, by using a hard hammer directly on the bone it may either over fracture the bone or even crush it due to the direct propagation of force (and partially damage the marrow).
+The use of a wedge for these activities provides more control, avoiding complete crushing of the bone.
+It also is much easier to cleanly extract bone marrow from a cracked bone than from a smashed bone.
+According to both Leblanc [-@leblanc_wedges_1992] and Shott [-@shott_bipolar_1999] wedges are needed to work bone, antler, and wood and stone wedges are preferable than other types of material.
+For hunter-gatherers, these artifacts would provide multifunctionality from a single piece that could be continuously used until losing their morphological characteristics.
+And as previously mentioned, the fact that the presence of these artifacts is constant throughout Vale Boi’s Upper Paleolithic corroborates the importance that these pieces seem to have had.
+In addition, scaled pieces in Vale Boi may also be related to bone tool production, since Vale Boi is the Portuguese Upper Paleolithic site with more evidence for the production and use of bone tool technology [@evora_osseous_2013].
+
+Several researchers argued that scaled pieces were used as cores for the extraction of chips or small bladelets [e.g.
+@carvalho_talhe_1998; @de_la_pena_sobre_2011; @leblanc_wedges_1992; @shott_bipolar_1989; @shott_bipolar_1999; @zilhao_o_1997].
+Here, we argue that the concept itself is flawed, especially when applied to Vale Boi.
+With the single exception of the early Gravettian, the site’s technology shows no evidence of manufacture or use of tools with such small dimensions (5-8 mm).
+The most common interpretation is that these chips or bladelets would be inserted in composite throwing tools.
+If we look at the average length of Vale Boi’s chips removed from scaled pieces we have an average of 7 mm.
+The functionality of chips of this size is quite arguable, even when inserted into a composite tool, and especially the impact that these would have in a perforation or extension of bleeding scenario opposed to using regular bladelets.
+Also, there is no evidence of bone or antler tools with perforation marks for the insertion of chips, although this kind of weapon could have been exclusively made with perishable materials such as wood.
+Regarding the technology, as previously mentioned, in Vale Boi every phase of the chaîne operatoire is present, which means the knapping was made in situ, and so the number of chips found is quite high.
+Considering that most knapping activities originate chips, the need of a specific tool for the sole purpose of extracting them does not seem viable.
+By putting all these factors in context, the hypothesis that these pieces were used for carcass processing, bone marrow extracting [@manne_intensive_2012] and other similar activities (e.g.
+working wood or antler with the goal of producing utensils) seems to be the most reasonable interpretation.
+
+Our results also suggest that higher degrees of representation of scaled pieces in sites can be originated by factors such as site functionality but also by cultural patterns.
+Regarding site functionality, if we look at the sites with higher counts of scaled pieces (Vale Boi’s Terrace and Slope and Fonte Santa), both are open-air residential sites.
+One example outside of this region is the Southwestern German Magdalenian site of Gönnersdorf, where according to Sano (2012) scaled pieces were the second largest assemblage from the site’s retouched utensils (257 from a total of 1501 utensils), corresponding to c.
+17,12% of the total assemblage.
+This typological index is as high as both Vale Boi’s and Fonte Santa’s and like both of these sites, Gönnersdorf is an open-air residential site.
+However, these data can be misleading, as Casal do Felipe (TI=5,91), Terra do Manuel 1940-42 (TI=1,03), Terra do Manuel 1988-89 (TI=2,13) and Vale Comprido – Encosta (TI=0,9) are also open-air multifunctional sites.
+These sites show the same functionality yet, smaller representations of scaled pieces.
+
+It is quite reasonable that scaled pieces should be linked mostly with residential sites since the activities for which they are applied are conducted in a residential scenario rather than a hunting station, quarry, workshop or temporary shelter.
+While we cannot fully discard the fact that scaled pieces can be used as a single representation of site functionality, they can be a tool for such ends, as argued by Jeske and Sterner-Miller [-@jeske_microwear_2015].
+We, thus, argue that in this region, bipolar technology in the Upper Paleolithic may be also linked to cultural traditions rather than just simply functional ones.
+By considering that most sites aforementioned show similarities from a functional, chronological, faunal, floral and climate standpoints, there is no clear reason for the presence of different relative numbers of scaled pieces at each site other than a combination of both cultural and functional patterns.
+
+\newline
 

### 2018-07-31 analysis/paper/paper.Rmd
Reordered discussion

@@
 
-Bipolar technology clearly had an important role on the first modern humans in Western Iberia.
+Bipolar technology clearly had an important role on the adaptive systems of the first modern humans in Western Iberia and elsewhere.
 As mentioned by Shot and Tostevin [-@shott_diversity_2015], a rise in bipolar technology frequency may be related to human mobility patterns and even evolutionary trends [@duke_weaving_2015].
-For the first modern humans, as well as their descendants, this technology would have played a major role in resource exploitation and diversification.
-In Vale Boi, bipolar technology shows up in the archaeological record connected to evidence related to the intensification and diversification of resource exploration such as the above-mentioned grease-rendering techniques.
-These have been shown by the constant presence of impact fractures in ungulate bones, low percentage of bone areas related to higher amounts of grease, and the constant correlation between red deer bone fragments with the amount of bone marrow and grease contained in them.
-Other functional indicators are both the constant presence of thermally altered quartz fragments that might be linked to stone boiling activities, and a large number of greywacke slabs with impact marks that were used as anvils.
-
-In Table 6 we present current data based on previous work on bipolar technology in Westernmost Iberia for the Upper Paleolithic.
-While the number of artifacts is not very high, bipolar technology is present in a significant number of sites.
-Even with the considerable number of sites represented, it is clear the discrepancy in counts when comparing the presence of scaled pieces and bipolar cores, which by itself could be indicative of some sort of functional pattern.
-It is clear, however, that scaled pieces show on average, a higher representation in Vale Boi’s assemblages than on most other sites.
-Further, of the three bipolar cores in all sites, 2 of these come from Vale Boi.
-While currently there are no data on the Typological Index (TI) of the Slope’s scaled pieces, since the rest of the assemblage has not yet been studied, it is fairly safe to assume that the TI would have a particularly high value.
-Bipolar technology is quite common in Western Iberian Upper Paleolithic sites and interpretations made are solely based on a single reference with no statistical or analytical evidence.
-We argue that all factors mentioned in this study must be considered while interpreting each site in this region.
-While in Vale Boi there is clear evidence that the use of these artifacts would not be for bipolar knapping, in other sites the picture may be different, although these interpretations should be in the future backed by some sort of analytical evidence.
-As our results show, the low presence of bipolar cores is quite evident in these sites, while scaled pieces are more often present.
-It can also be argued that the number of pieces in the overall picture is also low, although presenting an accurate picture for each site since most of these sites have been extensively studied.
+For the first modern humans, as well as their descendants, this technology would have played a major role in the intensive and diverse exploitation of resources.
+In Vale Boi, bipolar technology shows up in the archaeological record connected to evidence related to the intensification and diversification of resource exploration such as the above-mentioned grease obtention techniques.
+These have been shown by the constant presence of impact fractures in ungulate bones, low percentage of bone areas related to higher amounts of grease, and the presence of red deer bone fragments from bones with considerably higher marrow and grease contents.
+Other  indicators are both the constant presence of thermally altered quartz fragments that might be linked to stone boiling activities, and a large number of greywacke slabs with impact marks revealing their use as anvils.
 
@@
 @carvalho_talhe_1998; @de_la_pena_sobre_2011; @leblanc_wedges_1992; @shott_bipolar_1989; @shott_bipolar_1999; @zilhao_o_1997].
-Here, we argue that the concept itself is flawed, especially when applied to Vale Boi.
-With the single exception of the early Gravettian, the site’s technology shows no evidence of manufacture or use of tools with such small dimensions (5-8 mm).
-The most common interpretation is that these chips or bladelets would be inserted in composite throwing tools.
-If we look at the average length of Vale Boi’s chips removed from scaled pieces we have an average of 7 mm.
-The functionality of chips of this size is quite arguable, even when inserted into a composite tool, and especially the impact that these would have in a perforation or extension of bleeding scenario opposed to using regular bladelets.
-Also, there is no evidence of bone or antler tools with perforation marks for the insertion of chips, although this kind of weapon could have been exclusively made with perishable materials such as wood.
-Regarding the technology, as previously mentioned, in Vale Boi every phase of the chaîne operatoire is present, which means the knapping was made in situ, and so the number of chips found is quite high.
-Considering that most knapping activities originate chips, the need of a specific tool for the sole purpose of extracting them does not seem viable.
-By putting all these factors in context, the hypothesis that these pieces were used for carcass processing, bone marrow extracting [@manne_intensive_2012] and other similar activities (e.g.
-working wood or antler with the goal of producing utensils) seems to be the most reasonable interpretation.
+Based on our data we find that this concept is somehow flawed, especially when applied to Vale Boi.
+First, with the single exception of the early Gravettian (during which very small backed bladelets are present - see Marreiros et al [-@marreiros_lithic_2018]), the site’s technology shows no evidence of manufacture or use of tools with such small dimensions as the ones recorded from the scars of scaled pieces (5-8 mm).
+Second, although the most common interpretation is that these chips/bladelets would be inserted in composite throwing tools, the functionality of chips of ~7mm is quite arguable, especially regarding the impact that these would have in a perforation or extension of a bleeding scenario, as opposed to using regular bladelets.
+Third, although Vale Boi has a very rich assemblage of organic tools [@evora_osseous_2013], there is no evidence of bone or antler tools with perforation marks for the insertion of stone implements.
+Finally, as previously mentioned, Vale Boi stone tool technological analyses revealed that every phase of the chaîne operatoire is present, which means that knapping was mostly occurring at the site, which is attested by the high number of chips found.
+Considering that most knapping activities originate chipage, the need of a specific tool for the sole purpose of extracting them does not seem viable to us.
+By putting all these factors in context, the hypothesis that these pieces were used for carcass processing, bone marrow extracting [@manne_intensive_2012] and other similar activities, such as working hard, organic materials with the goal of producing tools seems, thus far, to be the most reasonable interpretation.
 
-Our results also suggest that higher degrees of representation of scaled pieces in sites can be originated by factors such as site functionality but also by cultural patterns.
-Regarding site functionality, if we look at the sites with higher counts of scaled pieces (Vale Boi’s Terrace and Slope and Fonte Santa), both are open-air residential sites.
+In Table 6 we present current data on Upper Paleolithic bipolar technology in Portugal.
+Although the number of scaled pieces and bipolar cores is considerably low in most of the sites, a significant number of *loci* present some kind of bipolar technology evidence.
+Moreover, it is still clear a discrepancy when comparing the presence of scaled pieces versus bipolar cores, the latter being well under represented than the first.
+It is clear, also, that scaled pieces show on average a higher representation in Vale Boi’s assemblages than on most other sites.
+Further, of the three bipolar cores in all sites, 2 of these come from Vale Boi.
+While currently there are no definitive data on the representation of scaled pieces within the retouched tool assemblage of the Slope area, it is fairly safe to assume that it would have a particularly high value.
+
+Bipolar technology is quite common in Upper Paleolithic sites of Western Iberia, but current interpretations are based on empirical observations, rather than on more thoroughly analytical evidence.
+We argue that all factors mentioned in this study must be considered while interpreting each site in this region.
+While in Vale Boi there is clear evidence that the use of these artifacts would not be for bipolar knapping, in other sites the picture may be different.
+
+As our results show, the low presence of bipolar cores is quite evident in these sites, while scaled pieces are more often present.
+It can also be argued that the number of pieces in the overall picture is also low, although presenting an accurate picture for each site since most of these sites have been extensively studied.
+
+Higher degrees of representation of scaled pieces in sites can be originated by factors such as site function, but also by cultural patterns.
+Regarding site functionality, if we look at the *loci* with higher counts of scaled pieces (Vale Boi’s Terrace and Slope and Fonte Santa), both are open-air residential sites.
 One example outside of this region is the Southwestern German Magdalenian site of Gönnersdorf, where according to Sano (2012) scaled pieces were the second largest assemblage from the site’s retouched utensils (257 from a total of 1501 utensils), corresponding to c.

### 2018-08-01 analysis/paper/paper.Rmd
Discussion revised

@@
 
-Bipolar technology clearly had an important role on the adaptive systems of the first modern humans in Western Iberia and elsewhere.
-As mentioned by Shot and Tostevin [-@shott_diversity_2015], a rise in bipolar technology frequency may be related to human mobility patterns and even evolutionary trends [@duke_weaving_2015].
-For the first modern humans, as well as their descendants, this technology would have played a major role in the intensive and diverse exploitation of resources.
-In Vale Boi, bipolar technology shows up in the archaeological record connected to evidence related to the intensification and diversification of resource exploration such as the above-mentioned grease obtention techniques.
+Bipolar technology clearly had an important role on the adaptive systems of the first modern humans in Western Iberia, and elsewhere [@villa_neandertals_2018].
+A rise in bipolar technologies cannot be dissociated from the diverse set of factors that made Anatomically Modern Humans thrive.
+Changes in mobility patterns [@shott_diversity_2015], or the development of a "generalist specialist" ecological approach [@roberts_defining_2018], with particular emphasis on the diversification and intensification of resources use, are among some of those traits.
+It is in the contexts of this latter point that bipolar technology may have played a major role.
+In Vale Boi, since the earliest occupations at c.
+32 ka cal BP, scaled pieces show up in the archaeological record connected to evidence related to an intensification and diversification of resource exploration, particularly the above-mentioned grease/marrow obtention techniques.
 These have been shown by the constant presence of impact fractures in ungulate bones, low percentage of bone areas related to higher amounts of grease, and the presence of red deer bone fragments from bones with considerably higher marrow and grease contents.
 Other indicators are both the constant presence of thermally altered quartz fragments that might be linked to stone boiling activities, and a large number of greywacke slabs with impact marks revealing their use as anvils.
+For carcass processing and bone marrow extraction the use of scaled pieces would allow for a better control of bone fracture.
+In fact, this would be much beteer then using a hard hammer directly on the bone, since this may either over fracture the bone or even crush it due to the lack of precision.
+The use of a wedge for these activities provides more control, avoiding complete crushing of the bone, making it much easier to cleanly extract bone marrow.
 
-Our results show that in Vale Boi, scaled pieces have several patterns that stay consistent across all occupations.
+According to both Leblanc [-@leblanc_wedges_1992] and Shott [-@shott_bipolar_1999] wedges are needed to work bone, antler, and wood and stone wedges are preferable than other types of material.
+For hunter-gatherers, these artifacts would provide multifunctionality from a single piece that could be continuously used until losing their morphological characteristics.
+And as previously mentioned, the fact that the presence of these artifacts is constant throughout Vale Boi’s Upper Paleolithic corroborates the importance that these pieces seem to have had.
+In addition, scaled pieces in Vale Boi may also be related to bone tool production, since Vale Boi is the Portuguese Upper Paleolithic site with more evidence for the production and use of bone tool technology [@evora_osseous_2013].
+
+Importantly, our results show that scaled pieces use at Vale Boi reveal techno-functional patterns that stay consistent across all paleolithic horizons.
+
 One relevant trend is the fact that while quartz pieces were only explored in a single axis, chert pieces were, sometimes, explored in multiple axes.
@@
 In this specific case, we argue that there is little to no control on how the piece’s features change over its lifetime use.
-
-From a paleoanthropological point of view, this type of bipolar technique would have a major role on a day to day basis.
-For carcass processing and bone marrow extraction the use of scaled pieces would bring two major advantages: (1) it would allow for a better control of bone fracture; and, in consequence, (2) it would allow a clear choice of which part of the bone can either be broken or removed.
-It can be argued that both activities can be performed with any kind of hammer.
-However, by using a hard hammer directly on the bone it may either over fracture the bone or even crush it due to the direct propagation of force (and partially damage the marrow).
-The use of a wedge for these activities provides more control, avoiding complete crushing of the bone.
-It also is much easier to cleanly extract bone marrow from a cracked bone than from a smashed bone.
-According to both Leblanc [-@leblanc_wedges_1992] and Shott [-@shott_bipolar_1999] wedges are needed to work bone, antler, and wood and stone wedges are preferable than other types of material.
-For hunter-gatherers, these artifacts would provide multifunctionality from a single piece that could be continuously used until losing their morphological characteristics.
-And as previously mentioned, the fact that the presence of these artifacts is constant throughout Vale Boi’s Upper Paleolithic corroborates the importance that these pieces seem to have had.
-In addition, scaled pieces in Vale Boi may also be related to bone tool production, since Vale Boi is the Portuguese Upper Paleolithic site with more evidence for the production and use of bone tool technology [@evora_osseous_2013].
 
@@
 
-Bipolar technology is quite common in Upper Paleolithic sites of Western Iberia, but current interpretations are based on empirical observations, rather than on more thoroughly analytical evidence.
+Bipolar technology is quite common in Upper Paleolithic contexts of Western Iberia, but current interpretations are based on empirical observations, rather than on more solid analytical evidence.
 We argue that all factors mentioned in this study must be considered while interpreting each site in this region.

### 2018-10-03 analysis/paper/paper.Rmd
final version pre-submission

@@
 abstract: |
+  Scaled or splintered pieces are one of the most common lithic artifacts type in Upper Paleolithic assemblages throughout Europe, especially in its westernmost regions.
+Despite this, even after one century of being identified there is still no consensus on how to define, analyze or even interpret these tools.
+In western Iberia there is a clear lack of comprehensive studies regarding this type of artifacts at a regional scale.
+In this paper we present a first techno-morphological analysis of a sample of scaled pieces from the Upper Paleolithic site of Vale Boi.
+Our first aim was to build upon existing analytical models in order to identify function and possible reduction strategies for these artifacts.
+Our second goal was to critically evaluate the role of these artifacts within western Iberia’s Upper Paleolithic.
+  Our results showed that functional identification of scaled pieces is still not clear.
+By comparing our data with other author’s we found that current models could not be applied to the archaeological record, as the attribute variability is too high.
+Furthermore, in this region we found that higher frequencies of bipolar technology can be found related to residential sites due to both functional and cultural patterns.
+While we still cannot define a specific function for these artifacts (intermediate pieces or wedges for working hard raw materials or cores for the extraction of chips and small bladelets), it is clear that they had a major role in the variability and intensification of resource exploitation during the Upper Paleolithic in western Iberia.
 
@@
 
+\newline
+
 ```{r VBLocation, echo = FALSE, fig.cap="Location of the site of Vale Boi"}
 
-Map()
+# Map()
+# The function above is used to build the map shown below. Since the map is saved in the figures folder of the compendium it is not needed to run this function again.
 
@@
 Regarding terrestrial fauna, three species dominate the vertebrate group: rabbit (*Oryctolagus cuniculus*), red deer (*Cervus elaphus*) and horse (*Equus caballus*).
-Smaller amounts of wild ass (*Equus hydruntinus*), aurochs (*Bos primigenius*), ibex (Capra pyrenaica) and wild boar (Sus scrofa) are also present [@manne_early_2014; @manne_intensive_2012; @manne_upper_2010].
+Smaller amounts of wild ass (*Equus hydruntinus*), aurochs (*Bos primigenius*), ibex (*Capra pyrenaica*) and wild boar (*Sus scrofa*) are also present [@manne_early_2014; @manne_intensive_2012; @manne_upper_2010].
 One of the most interesting patterns within the faunal assemblages is that a large portion of ungulate remains present specific types of fracture that have been associated with bone marrow extraction activities.
@@
 In the first group, a series of variables traditionally used in lithic studies [e.g.
-@andrefsky_jr_lithics:_2005; @inizan_technology_1999] were recorded (Table \@ref(tab:tech_attributes)), aiming to characterize patterns of blank choice for the application of bipolar technology.
+@andrefsky_jr_lithics:_2005; @inizan_technology_1999] were recorded (see Supplementary Materials for full list of variables), aiming to characterize patterns of blank choice for the application of bipolar technology.
 For the second group of variables, a macroscopic approach building upon the work of de la Peña [-@de_la_pena_sobre_2011] and Fischer et.
-al [-@fischer_macro_1984] was adopted (Table \@ref(tab:morph_attributes)).
+al [-@fischer_macro_1984] was adopted (see Supplementary Materials for full list of variables).
 Studies by Sano [-@sano_functional_2012], and Gibaja et al.
@@
 To assist us with this differentiation we tested the presence of the referred patterns within our assemblage using descriptive and multivariate statistical analysis and comparing it with the data described by the referred authors.
-
-\newline
-
-```{r tech_attributes, echo = FALSE}
-
-```
-
-\newline
-
-```{r morph_attributes, echo = FALSE}
-
-```
 
@@
 
-knitr::kable(Gravettian, caption= "Technological attribute frequencies by raw materials for the Gravettian sample")
+knitr::kable(Gravettian, caption= "Technological attribute frequencies by raw materials for the Gravettian sample. Percentages are shown in parentheses.")
 
@@
 
-knitr::kable(`Proto-Solutrean`, caption= "Technological attribute frequencies by raw materials for the Proto-Solutrean sample")
+knitr::kable(`Proto-Solutrean`, caption= "Technological attribute frequencies by raw materials for the Proto-Solutrean sample. Percentages are shown in parentheses.")
 
@@
 
-knitr::kable(Solutrean, caption= "Technological attribute frequencies by raw materials for the Solutrean sample")
+knitr::kable(Solutrean, caption= "Technological attribute frequencies by raw materials for the Solutrean sample. Percentages are shown in parentheses.")
 
@@
 
-knitr::kable(Magdalenian, caption= "Technological attribute frequencies by raw materials for the Magdalenian sample")
+knitr::kable(Magdalenian, caption= "Technological attribute frequencies by raw materials for the Magdalenian sample. Percentages are shown in parentheses.")
 
 ```
+
+\newline
 
@@
 As in other classes of stone tools, scaled pieces' metric attributes are impacted by both the initial blank size as well as by the intensity of their use.
-Specifically, because some scaled pieces are used in multiple axes, direct comparisons for length and width of the typological axes (see Table \@ref(tab:tech_attributes)) for definition) cannot be straightforwardly performed.
+Specifically, because some scaled pieces are used in multiple axes, direct comparisons for length and width of the typological axes (see Suplementary Materials for definition) cannot be straightforwardly performed.
 Since the majority of the analyzed pieces presented a rectangular outline, we use the area of a rectangle (i.e.
 Length x Width) as an approximation for the overall dimensions of the artifacts.
-Area calculations revealed a maximum of `r round(max(techno_data$Area, na.rm = TRUE), digits = 2)` mm and a minimum of `r round(max(techno_data$Area, na.rm = TRUE), digits = 2)` mm.
+Area calculations revealed a maximum of `r round(max(techno_data$Area, na.rm = TRUE), digits = 2)` mm and a minimum of `r round(min(techno_data$Area, na.rm = TRUE), digits = 2)` mm.
 For thickness the maximum is `r round(max(techno_data$Thickness, na.rm = TRUE), digits = 2)` mm and the minimum is `r round(min(techno_data$Thickness, na.rm = TRUE), digits = 2)` mm.
@@
 In every assemblage quartz pieces were exclusively used in one single axis, exhibiting only two damaged platforms.
-On the other end, a small number of chert and chalcedony artifacts (*N* = `r techno_data %>% filter(RawMaterial != "Quartz" & DamagedPlatforms == "3" | DamagedPlatforms == "4_or_more") %>% summarize(n())`) exhibit multiple functional axis, with three or four damaged platforms (Figure \@ref(fig:nplatforms)).
+On the other end, a small number of chert and chalcedony artifacts (n = `r techno_data %>% filter(RawMaterial != "Quartz" & DamagedPlatforms == "3" | DamagedPlatforms == "4_or_more") %>% summarize(n())`) exhibit multiple functional axis, with three or four damaged platforms (Figure \@ref(fig:nplatforms)).
 This seems to indicate different strategies of curation for coarse and fine grain raw materials, with fine grain materials being rotated when the first used axis becomes too small and/or the edges get useless for that specific activity.
-However, when plotted against metric data, results for the Area variable reveal that pieces with four damaged platforms are among the largest in the whole assemblages, and that there is not a visible difference between the thickness of the pieces comprising each group.
+However, when plotted against metric data (Figure \@ref(fig:plotbyplatforms)), results for the Area variable reveal that pieces with four damaged platforms are among the largest in the whole assemblages, and that there is not a visible difference between the thickness of the pieces comprising each group.
 This seems to attest that the use of several axis in the same piece is not correlated with curation occuring in later phases of artifact use but instead to a probable difference in raw material performance.
@@
 
-```{r nplatforms, echo=FALSE, fig.cap=""}
+```{r nplatforms, echo=FALSE, fig.cap="Number of damaged platforms by raw material and chronology."}
 
@@
 
-```{r plotbyplatforms, echo = FALSE, fig.caption = ""}
+```{r plotbyplatforms, echo = FALSE, fig.cap="Boxplot of Area and Thickness distribution for each raw material."}
 
@@
 \newline
+
 ```{r chisquare, echo = FALSE}
@@
 
-```{r morphoattributes, echo = FALSE}
+```{r morphoattributes, echo = FALSE, fig.cap="Frequency of morphological attributes for each raw material.
+Damaged platforms were grouped so that each artifact was only counted once and to avoid wrong comparisons between active a hammered platforms.
+A - Distribution of damage; B - Arrangement of scars; C - Extension of scars; D - Delineation of damaged platforms; E - Facial distribution of scars; F - Angle of damaged platforms."}
 
@@
 To identify possible patterns of association among the qualitative variables used in our analysis \@ref(fig:morphoattributes), a Multiple Correspondence Analysis (MCA) was performed.
-Results are presented in Figures \@ref(fig:mca).
 In this analysis, raw material was used as a qualitative supplementary variable.
@@
 @de_la_pena_qualitative_2015; @de_la_pena_sobre_2011], in which certain combination of attributes were used to discriminate scaled pieces functionalities, are difficult to apply to the assemblage used in this study.
-
-The high variability of the assemblage is also well noticeable in the biplots presented in Figure \@ref(fig:biplot).
 
@@
 # Screeplot
-fviz_screeplot(morpho_data_mca, addlabels = TRUE, ylim = c(0, 15), barfill = "#00AFBB", barcolor = "#00AFBB")
+fviz_screeplot(morpho_data_mca, addlabels = TRUE, ylim = c(0, 15), barfill = "grey50", barcolor = "grey50")
 
@@
 
-```{r biplot, echo = FALSE}
-
-# Biplot for 4 main MCA dimensions
-
-fviz_screeplot(morpho_data_mca, addlabels = TRUE, ylim = c(0, 15), barfill = "#00AFBB", barcolor = "#00AFBB")
-
-```
-
-\newline
-
-#Discussion and conclusions
+# Discussion and conclusions
 
@@
 Changes in mobility patterns [@shott_diversity_2015], or the development of a "generalist specialist" ecological approach [@roberts_defining_2018], with particular emphasis on the diversification and intensification of resources use, are among some of those traits.
-It is in the contexts of this latter point that bipolar technology may have played a major role.
+It is in the context of this latter point that bipolar technology may have played a major role.
 In Vale Boi, since the earliest occupations at c.
-32 ka cal BP, scaled pieces show up in the archaeological record connected to evidence related to an intensification and diversification of resource exploration, particularly the above-mentioned grease/marrow obtention techniques.
-These have been shown by the constant presence of impact fractures in ungulate bones, low percentage of bone areas related to higher amounts of grease, and the presence of red deer bone fragments from bones with considerably higher marrow and grease contents.
-Other indicators are both the constant presence of thermally altered quartz fragments that might be linked to stone boiling activities, and a large number of greywacke slabs with impact marks revealing their use as anvils.
+32 ka cal BP, scaled pieces show up in the archaeological record connected to evidence related to an intensification and diversification of resource exploration, particularly the aforementioned grease/marrow obtention techniques.
+These have been shown by the constant presence of impact fractures in ungulate bones, low percentage of bone areas related to higher amounts of grease, and the presence of red deer bone fragments from bones with considerably higher marrow and grease contents [@manne_early_2014; @manne_intensive_2012; @manne_upper_2010; @manne_vale_2009].
+Other indicators are both the constant presence of thermally altered quartz fragments that might be linked to stone boiling activities, and a large number of greywacke slabs with impact marks revealing their use as anvils [@cascalheira_cross-scale_2017].
 For carcass processing and bone marrow extraction the use of scaled pieces would allow for a better control of bone fracture.
-In fact, this would be much beteer then using a hard hammer directly on the bone, since this may either over fracture the bone or even crush it due to the lack of precision.
+In fact, this should be significantly better than using a hard hammer directly on the bone, since the latter technique may either over fracture the bone or even crush it due to the lack of precision.
 The use of a wedge for these activities provides more control, avoiding complete crushing of the bone, making it much easier to cleanly extract bone marrow.
 
-According to both Leblanc [-@leblanc_wedges_1992] and Shott [-@shott_bipolar_1999] wedges are needed to work bone, antler, and wood and stone wedges are preferable than other types of material.
+In addition, scaled pieces in Vale Boi may also be associated to bone tool production, given that Vale Boi is one of the portuguese Upper Paleolithic sites with significant evidence for the onsite production and use of bone tools [@evora_osseous_2013].
+According to both Leblanc [-@leblanc_wedges_1992] and Shott [-@shott_bipolar_1999] wedges are needed to work bone, antler, and wood, and stone wedges are preferable than other types of material.
 For hunter-gatherers, these artifacts would provide multifunctionality from a single piece that could be continuously used until losing their morphological characteristics.
-And as previously mentioned, the fact that the presence of these artifacts is constant throughout Vale Boi’s Upper Paleolithic corroborates the importance that these pieces seem to have had.
-In addition, scaled pieces in Vale Boi may also be related to bone tool production, since Vale Boi is the Portuguese Upper Paleolithic site with more evidence for the production and use of bone tool technology [@evora_osseous_2013].
 
-Importantly, our results show that scaled pieces use at Vale Boi reveal techno-functional patterns that stay consistent across all paleolithic horizons.
-
+Importantly, our results show that the use of scale pieces at Vale Boi reveals techno-functional patterns that seem to stay fairly similar across all paleolithic horizons.
 One relevant trend is the fact that while quartz pieces were only explored in a single axis, chert pieces were, sometimes, explored in multiple axes.
-One possible, rather improbable, reason for these patterns might be the fact that quartz would be more easily available than chert.
-Another, more likely, possibility is due to the material properties of each raw material.
-While with quartz it may be possible to continuously use a single axis in a piece, reducing it in only one axis without losing its efficiency, with chert it may be necessary to rotate it in order to continue obtaining usable edges.
+One possible reason for these patterns might be the fact that quartz would be more easily available than chert, and thus tool economy would have been different between both materials.
+Another possibility for this is the difference in the physical properties of each raw material.
+While with quartz it may be possible to continuously use a single axis in a piece, reducing it in only one axis without losing its efficiency, with chert it may be necessary to rotate it more often in order to continue obtaining usable edges.
 On the other hand, the higher number of axis could simply suggest that chert was differently managed and economized.
-Nevertheless, the simple fact that most of the pieces with multiple axes are on average larger than the rest suggests that this hypothesis is likely incorrect.
+Nevertheless, the simple fact that most of the pieces with multiple axes are on average larger than the rest, suggests that this hypothesis is likely incorrect.
 From an economic standpoint, the ideal would be to continuously use the same piece without any risk of loss of the features that made it ideal for this kind of use.
@@
 Second, although the most common interpretation is that these chips/bladelets would be inserted in composite throwing tools, the functionality of chips of ~7mm is quite arguable, especially regarding the impact that these would have in a perforation or extension of a bleeding scenario, as opposed to using regular bladelets.
-Third, although Vale Boi has a very rich assemblage of organic tools [@evora_osseous_2013], there is no evidence of bone or antler tools with perforation marks for the insertion of stone implements.
+Third, although Vale Boi has a very rich assemblage of organic tools [@evora_osseous_2013], there is no evidence of bone or antler tools with grooves that might be used for the insertion of stone implements.
 Finally, as previously mentioned, Vale Boi stone tool technological analyses revealed that every phase of the chaîne operatoire is present, which means that knapping was mostly occurring at the site, which is attested by the high number of chips found.
@@
 
-In Table 6 we present current data on Upper Paleolithic bipolar technology in Portugal.
+In Table \@ref(tab:scaledpiecesportugal) we present current data on Upper Paleolithic bipolar technology in Portugal.
 Although the number of scaled pieces and bipolar cores is considerably low in most of the sites, a significant number of *loci* present some kind of bipolar technology evidence.
-Moreover, it is still clear a discrepancy when comparing the presence of scaled pieces versus bipolar cores, the latter being well under represented than the first.
-It is clear, also, that scaled pieces show on average a higher representation in Vale Boi’s assemblages than on most other sites.
+Moreover, it is clear a discrepancy when comparing the presence of scaled pieces versus bipolar cores, the latter being well under represented than the first.
+It is also clear that scaled pieces show on average a higher representation in Vale Boi’s assemblages than on most other sites.
 Further, of the three bipolar cores in all sites, 2 of these come from Vale Boi.
 While currently there are no definitive data on the representation of scaled pieces within the retouched tool assemblage of the Slope area, it is fairly safe to assume that it would have a particularly high value.
+
+\newline
+
+```{r scaledpiecesportugal, echo = FALSE}
+
+scaledpiecesportugal <- read_csv("../data/raw_data/scaled_pieces_UP_portugal.csv")
+
+knitr::kable(scaledpiecesportugal, caption= "Frequencies of scaled pieces in portuguese Upper Paleolithic sites")
+
+```
+
+\newline
 
@@
 One example outside of this region is the Southwestern German Magdalenian site of Gönnersdorf, where according to Sano (2012) scaled pieces were the second largest assemblage from the site’s retouched utensils (257 from a total of 1501 utensils), corresponding to c.
-17,12% of the total assemblage.
+17,12% of the total retouched tools assemblage.
 This typological index is as high as both Vale Boi’s and Fonte Santa’s and like both of these sites, Gönnersdorf is an open-air residential site.
-However, these data can be misleading, as Casal do Felipe (TI=5,91), Terra do Manuel 1940-42 (TI=1,03), Terra do Manuel 1988-89 (TI=2,13) and Vale Comprido – Encosta (TI=0,9) are also open-air multifunctional sites.
-These sites show the same functionality yet, smaller representations of scaled pieces.
+However, these data can be misleading, as Casal do Felipe (5,91%), Terra do Manuel 1940-42 (1,03%), Terra do Manuel 1988-89 (2,13%) and Vale Comprido – Encosta (0,9%) are also open-air multifunctional sites.
+These sites show the same functionality yet, much smaller representations of scaled pieces.
 
-It is quite reasonable that scaled pieces should be linked mostly with residential sites since the activities for which they are applied are conducted in a residential scenario rather than a hunting station, quarry, workshop or temporary shelter.
+It seems reasonable that scaled pieces should be linked mostly with residential sites since the activities for which they are applied are conducted in a residential scenario rather than a hunting station, quarry, workshop or temporary shelter.
 While we cannot fully discard the fact that scaled pieces can be used as a single representation of site functionality, they can be a tool for such ends, as argued by Jeske and Sterner-Miller [-@jeske_microwear_2015].
@@
 
-#Acknowledgements
+# Acknowledgements
 

### 2018-10-03 code.R

@@ -1,0 +1,52 @@
+barplot(res$eig[,2], names.arg = 1:nrow(res$eig))
+drawn <-
+c("122", "95", "66", "11", "69", "7", "36", "27", "101", "100", 
+"109", "110", "70")
+plot.MCA(res, select = drawn, axes = 1:2, choix = 'ind', invisible = c('var', 'quali'), title = '', cex = cex)
+wilks.p <-
+c(RawMaterial = 0.000225130888006352)
+wilks.p
+sample = sample(rownames(res$call$X), length(rownames(res$call$X)))
+res$call$X = res$call$X[sample,]
+res$ind$coord = res$ind$coord[sample[!sample %in% rownames(res$ind.sup$coord)],]
+res$ind.sup$coord = res$ind.sup$coord[sample[sample %in% rownames(res$ind.sup$coord)],]
+drawn <-
+c("122", "95", "66", "11", "69", "7", "36", "27", "101", "100", 
+"109", "110", "70")
+hab <-
+"RawMaterial"
+plotellipses(res, axes = 1:2, invisible = c('var', 'quali'), select = drawn, keepvar = hab, title = '', cex = cex)
+drawn <-
+c("ScarArrangement_Isolated/Isolated", "ScarDistribution_Total/Total", 
+"ScarExtension_Marginal/Marginal", "ScarArrangement_Other", "ScarDistribution_Central/Total", 
+"ScarDistribution_Central/Central", "ScarExtension_Other", "ScarArrangement_Aligned/Isolated"
+)
+plot.MCA(res, selectMod = drawn, axes = 1:2, choix = 'ind', invisible = 'ind', title = '', cex = cex)
+drawn <-
+c("26", "43", "68", "97", "49", "83", "65", "15", "28", "107", 
+"25", "69", "7")
+plot.MCA(res, select = drawn, axes = 3:4, choix = 'ind', invisible = c('var', 'quali'), title = '', cex = cex)
+wilks.p <-
+c(RawMaterial = 0.285066928942493)
+wilks.p
+sample = sample(rownames(res$call$X), length(rownames(res$call$X)))
+res$call$X = res$call$X[sample,]
+res$ind$coord = res$ind$coord[sample[!sample %in% rownames(res$ind.sup$coord)],]
+res$ind.sup$coord = res$ind.sup$coord[sample[sample %in% rownames(res$ind.sup$coord)],]
+drawn <-
[…]

### 2018-10-04 README.Rmd
formatting to submit

@@
 ```
+
+[![Last-changedate](https://img.shields.io/badge/last%20change-`r gsub('-', '--', Sys.Date())`-brightgreen.svg)](https://github.com/jmcascalheira/ScaledPiecesVB/commits/master)
+[![minimal R version](https://img.shields.io/badge/R%3E%3D-3.2.4-brightgreen.svg)](https://cran.r-project.org/) 
+[![Build Status](https://travis-ci.org/jmcascalheira/ScaledPiecesVB.svg?branch=master)](https://travis-ci.org/jmcascalheira/ScaledPiecesVB)
+[![Licence](https://img.shields.io/github/license/mashape/apistatus.svg)](http://choosealicense.com/licenses/mit/)
+[![ORCiD](https://img.shields.io/badge/ORCiD-/0000--0003--0321--8892-green.svg)](http://orcid.org/0000-0003-0321-8892)
 
@@
 Accessed `r format(Sys.Date(), "%d %b %Y")`.
-Online at <https://doi.org/xxx/xxx>
+Online at <https://doi.org/10.17605/OSF.IO/WPXGH>
 
@@
 
-There are no actual R functions in this package - all the R code is in the Rmd file.
-I simply used the R package structure to help manage dependencies, to take advantage of continuous integration for automated code testing, and so I didn't have to think too much about how to organise the files.
-
 To download the package source as you see it on GitHub, for offline browsing, use this line at the shell prompt (assuming you have Git installed on your computer):
@@
 
-Once the download is complete, open the ShortTermOccupations.Rproj in RStudio to begin working with the package and compendium files.
+Once the download is complete, open the ScaledPieces.Rproj in RStudio to begin working with the package and compendium files.
 
@@
 Installing these can be time-consuming and complicated, so we've included a packrat directory, which contains the source code for all the packages we depend on.
-If all works well, these will be installed on your computer when you open ShortTermOccupations.Rproj in RStudio.
+If all works well, these will be installed on your computer when you open ScaledPiecesVB.Rproj in RStudio.
 

### 2018-10-04 code.R

@@ -1,52 +1,0 @@
-barplot(res$eig[,2], names.arg = 1:nrow(res$eig))
-drawn <-
-c("122", "95", "66", "11", "69", "7", "36", "27", "101", "100", 
-"109", "110", "70")
-plot.MCA(res, select = drawn, axes = 1:2, choix = 'ind', invisible = c('var', 'quali'), title = '', cex = cex)
-wilks.p <-
-c(RawMaterial = 0.000225130888006352)
-wilks.p
-sample = sample(rownames(res$call$X), length(rownames(res$call$X)))
-res$call$X = res$call$X[sample,]
-res$ind$coord = res$ind$coord[sample[!sample %in% rownames(res$ind.sup$coord)],]
-res$ind.sup$coord = res$ind.sup$coord[sample[sample %in% rownames(res$ind.sup$coord)],]
-drawn <-
-c("122", "95", "66", "11", "69", "7", "36", "27", "101", "100", 
-"109", "110", "70")
-hab <-
-"RawMaterial"
-plotellipses(res, axes = 1:2, invisible = c('var', 'quali'), select = drawn, keepvar = hab, title = '', cex = cex)
-drawn <-
-c("ScarArrangement_Isolated/Isolated", "ScarDistribution_Total/Total", 
-"ScarExtension_Marginal/Marginal", "ScarArrangement_Other", "ScarDistribution_Central/Total", 
-"ScarDistribution_Central/Central", "ScarExtension_Other", "ScarArrangement_Aligned/Isolated"
-)
-plot.MCA(res, selectMod = drawn, axes = 1:2, choix = 'ind', invisible = 'ind', title = '', cex = cex)
-drawn <-
-c("26", "43", "68", "97", "49", "83", "65", "15", "28", "107", 
-"25", "69", "7")
-plot.MCA(res, select = drawn, axes = 3:4, choix = 'ind', invisible = c('var', 'quali'), title = '', cex = cex)
-wilks.p <-
-c(RawMaterial = 0.285066928942493)
-wilks.p
-sample = sample(rownames(res$call$X), length(rownames(res$call$X)))
-res$call$X = res$call$X[sample,]
-res$ind$coord = res$ind$coord[sample[!sample %in% rownames(res$ind.sup$coord)],]
-res$ind.sup$coord = res$ind.sup$coord[sample[sample %in% rownames(res$ind.sup$coord)],]
-drawn <-
[…]

### 2018-10-04 analysis/paper/SOM.Rmd
formatting to submit

---
title: "Supplementary Materials for 'The role of lithic bipolar technology in Western Iberia’s Upper Paleolithic: the case of Vale Boi (southern Portugal)'"
date: '`r Sys.Date()`'
output:
   bookdown::word_document2:
    fig_caption: no
    reference_docx: "../templates/template2.docx"
bibliography: ScaledPiecesVB.bib
csl: "../templates/springer-socpsych-author-date.csl"
---

```{r setup, include=FALSE, echo=FALSE}

knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  dpi = 300,
  fig.width = 9,
  fig.path = "../figures/")
  cache = TRUE

library(readr)

```

\newline

### Table 1 - Technological attributes used in the analysis of scaled pieces included in this study

```{r techno_attributes, echo=FALSE}

techno_attributes <- read_csv("../data/raw_data/technological_attributes.csv")

techno_attributes[is.na(techno_attributes)] <- " "

knitr::kable(techno_attributes)

```

\newline

###  Table 2 - Morpho-functional attributes used to analyze each damaged platform of a scaled piece. *Adapted from Gonzalez-Urquijo and Ibanez-Estévez [-@gonzalez-urquijo_metodologialisis_1994]. **Adapted from de la Peña [-@de_la_pena_sobre_2011]

```{r morpho_attributes, echo=FALSE}

morpho_attributes <- read_csv("../data/raw_data/functional_attributes.csv")

morpho_attributes[is.na(morpho_attributes)] <- " "

knitr::kable(morpho_attributes)

```

\newline

# References 
<div id="refs"></div>

##### pagebreak

# Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r colophon, cache = FALSE}
# which R packages and versions?
devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? You may need to change the path value
# if your Rmd is not in analysis/paper/
git2r::repository("../..")
```

### 2018-10-04 analysis/paper/_main.Rmd
formatting to submit

--- 2018-10-03 analysis/paper/paper.Rmd
@@
 bibliography: ScaledPiecesVB.bib
-csl: "../templates/springer-humanities-author-date.csl"
+csl: "../templates/springer-socpsych-author-date.csl"
 abstract: |
@@
 While we still cannot define a specific function for these artifacts (intermediate pieces or wedges for working hard raw materials or cores for the extraction of chips and small bladelets), it is clear that they had a major role in the variability and intensification of resource exploitation during the Upper Paleolithic in western Iberia.
+
+figsintext        : no
+figurelist        : no
+tablelist         : no
+footnotelist      : no
+lineno            : no
+keywords          : ""
 
@@
 library(epitools)
+library(wesanderson)
 #library(ScaledPiecesVB)
@@
 Most references broadly interpret these as bipolar cores for the extraction of small bladelets and chips to be  used in composite tools [e.g.
-@zilhao_o_1997] or as intermediate pieces for working hard materials [@gibaja_tecnologiy_2007; @marreiros_as_2009].
-Zilhão [-@zilhao_o_1997] for example argues that the presence of scaled pieces throughout the Upper Paleolithic sequence of central Portugal is inversely proportional to the presence of “carinated cores”, and thus it is likely that the former should have worked as a flexible substitute for the latter.
+@zilhao_o_1997], or as intermediate pieces for working hard materials [@gibaja_tecnologiy_2007; @marreiros_as_2009].
+Zilhão [-@zilhao_o_1997], for example, argues that the presence of scaled pieces throughout the Upper Paleolithic sequence of central Portugal is inversely proportional to the presence of “carinated cores”, and thus it is likely that the former should have worked as a flexible substitute for the latter.
 Carvalho [-@carvalho_talhe_1998] agrees with this interpretation and considers it also valid for the portuguese early Neolithic.
@@
 Schist, for instance, shows up in some occupations almost exclusively connected to mobile art.
-Dolerite was only  found in the Proto-Solutrean levels [@belmiro_o_2017], while chalcedony is mostly limited to the Proto-Solutrean and Solutrean levels [@cascalheira_vale_2012].
+Dolerite is only  found in the Proto-Solutrean levels [@belmiro_o_2017], while chalcedony is mostly limited to the Proto-Solutrean and Solutrean levels [@cascalheira_vale_2012].
 
@@
 These traces are caused by direct percussion at one end and subsequent crushing of the other because it is rested on a hard surface, and can be bifacially or unifacially distributed.
-As will become clear from our results some of the analyzed artifacts did not present splitering/crushing traces in opposed platforms, but we still classified them as scaled pieces.
+As will become clear from our results, some of the analyzed artifacts did not present splitering/crushing traces in opposed platforms, but we still classified them as scaled pieces.
 This detail has been previously referred by Villa et al.
@@
 Studies by Sano [-@sano_functional_2012], and Gibaja et al.
-[-@gibaja_tecnologiy_2007] indicate that since use-wear traces are formed before the splintering, the removal of small chips removes most of the polishes and use-wear traces left by the contact with the static element (either a stone anvil or a hard-organic materials).
+[-@gibaja_tecnologiy_2007] indicate that, since use-wear traces are formed before the splintering, the removal of small chips removes most of the polishes and use-wear traces left by the contact with the static element (either a stone anvil or a hard-organic materials).
 This, together with the large presence of quartz artifacts in our sample, prevented us to pursue a microscopic use-wear approach for this study in particular.
@@
 Macroscopic morpho-functional attributes were separately analyzed for each damaged platform, aiming to detect patterns of morphological change that occurred in artifacts during use.
-Following the work of de la Peña [-@de_la_pena_qualitative_2015; -@de_la_pena_sobre_2011] and Gonzalez-Urquijo & Ibanez-Estévez [-@gonzalez-urquijo_metodologialisis_1994], we expected that these attributes would be indicative of which function the artifacts had.
+Following the work of de la Peña [-@de_la_pena_qualitative_2015; -@de_la_pena_sobre_2011] and Gonzalez-Urquijo and Ibanez-Estévez [-@gonzalez-urquijo_metodologialisis_1994], we expected that these attributes would be indicative of which function the artifacts had.
 For instance, according to de la Peña [-@de_la_pena_qualitative_2015; -@de_la_pena_sobre_2011], for pieces used as wedges the delineation of the damaged platforms are constantly asymmetrical and only the hammered edge clearly shows the typical *écaille* retouch.
@@
 On the other hand, pieces used as bipolar cores would have squared or rectangular shapes, symmetric straight damaged platforms, and a higher frequency of scars on the hammered edge than on the opposed edge.
-The addition of other attributes drawing upon the work by Gonzalez-Urquijo & Ibanez-Estévez [-@gonzalez-urquijo_metodologialisis_1994] was made following the same reasoning.
+The addition of other attributes drawing upon the work by Gonzalez-Urquijo and Ibanez-Estévez [-@gonzalez-urquijo_metodologialisis_1994] was made following the same reasoning.
 We expected, thus, to be able to differentiate pieces used as wedges from pieces used as cores, as both groups would show distinct combinations of attributes.
@@
 
-A total of `r summarise(techno_data, n())` scaled pieces were analyzed, of which `r round(techno_data %>% filter(Chronology == "Gravettian") %>% summarise(n = n()) %>% mutate(n = (n / 139)*100), 2)`% come from Gravettian, and `r round(techno_data %>% filter(Chronology == "Solutrean") %>% summarise(n = n()) %>% mutate(n = (n / 139)*100), 2)`% from Solutrean levels, as shown in (Table \@ref(tab:assemblage)).
+A total of `r summarise(techno_data, n())` scaled pieces were analyzed, of which `r round(techno_data %>% filter(Chronology == "Gravettian") %>% summarise(n = n()) %>% mutate(n = (n / 139)*100), 2)`% come from Gravettian, and `r round(techno_data %>% filter(Chronology == "Solutrean") %>% summarise(n = n()) %>% mutate(n = (n / 139)*100), 2)`% from Solutrean levels, as shown in Table (\@ref(tab:assemblage)).
 In terms of raw materials, the great majority of pieces were either made on quartz or chert, with chalcedony being represented only by `r round(techno_data %>% filter(RawMaterial == "Chalcedony") %>% summarise(n = n()), 2)` artifacts.
@@
 Regarding chert, in every occupation flakes dominate the assemblages, followed by a reduced number (n = `r techno_data %>% filter(RawMaterial == "Chert" & Blank == "Non_identifiable") %>% summarise(n())`) of unclassifiable blanks.
-The small sample of Chalcedony blanks is exclusively represented by flakes.
+The small sample of chalcedony blanks is exclusively represented by flakes.
 
@@
 As in other classes of stone tools, scaled pieces' metric attributes are impacted by both the initial blank size as well as by the intensity of their use.
-Specifically, because some scaled pieces are used in multiple axes, direct comparisons for length and width of the typological axes (see Suplementary Materials for definition) cannot be straightforwardly performed.
+Specifically, because some scaled pieces are used in multiple axes, direct comparisons for length and width of the typological axes cannot be straightforwardly performed.
 Since the majority of the analyzed pieces presented a rectangular outline, we use the area of a rectangle (i.e.
@@
           color = "DamagedPlatforms", fill = "DamagedPlatforms",
-          palette = c("#00AFBB", "#E7B800", "#FC4E07", "#999999"),
-          facet.by = "Chronology", xlab = "", ylab = "%")
+          facet.by = "Chronology", xlab = "", ylab = "%") + 
+  scale_fill_manual(values = wes_palette("Rushmore")[c(1,3,4,5)]) +
+  scale_color_manual(values=wes_palette("Rushmore")[c(1,3,4,5)])
 
@@
 Figure \@ref(fig:morphoattributes) shows the relative frequencies of all qualitative variables recorded for each damaged platform.
-To simplify representation, and to avoid wrong comparisons between hammered and active platforms, opposed damage platforms are grouped into pairwise categories.
-When present, category 'Other' represents the cluster of attributes which frequency was less than 10% within each variable.
+To simplify representation, and to avoid wrong comparisons between hammered and active platforms, opposed damaged platforms were grouped into pairwise categories.
+When present, the category 'Other' represents the cluster of attributes which frequency was less than 10% within each variable.
 
-In three (Scar Arrangement, Extension and Edge Delineation) of the six represented variables the category 'Other' is one of the most frequent (above 35%) for both quartz and chert, revealing a very high variability for the combination of attributes within each variable.
+In three (Scar Arrangement, Scar Extension and Edge Delineation) of the six represented variables, the category 'Other' is one of the most frequent (above 35%) for both quartz and chert, revealing a very high variability for the combination of attributes within each variable.
 A chi-square test with modified alpha level (Bonferroni correction) to adjust for multiple testing and reduce type-I error, shows significant differences in the quartz and chert Scar Arrangement categories (*X^2^* [`r chi_arrangement$chi_test$parameter`, *N* = `r sum(chi_arrangement$chi_test$observed)`] = `r  round(chi_arrangement$chi_test$statistic, 2)`, *p* = `r round(chi_arrangement$chi_test$p.value, 3)`, *phi* = `r round(chi_arrangement$chi_size_effect$phi, 2)`), and in Scar Faciality categories (*X^2^* [`r chi_faciality$chi_test$parameter`, *N* = `r sum(chi_faciality$chi_test$observed)`] = `r  round(chi_faciality$chi_test$statistic,2)`, *p* = `r round(chi_faciality$chi_test$p.value, 3)`, *phi* = `r round(chi_faciality$chi_size_effect$phi, 2)`).
@@
 
-To identify possible patterns of association among the qualitative variables used in our analysis \@ref(fig:morphoattributes), a Multiple Correspondence Analysis (MCA) was performed.
+To identify possible patterns of association among the qualitative variables used in our analysis, a Multiple Correspondence Analysis (MCA) was performed.
 In this analysis, raw material was used as a qualitative supplementary variable.
@@
 The first two dimensions of the MCA express 20.92% of the total dataset inertia, meaning that only that percentage of the total variability is explained by the plane combining the first two dimensions.
-An inspection to the screeplot presented in Figure \@ref(fir:screeplot) suggests to restrict the analysis to the description of the first 4 dimensions.
+An inspection to the screeplot presented in Figure \@ref(fig:screeplot) suggests restricting the analysis to the description of the first 4 dimensions.
 These dimensions present an amount of inertia slightly greater than that obtained by the 0.95-quantile of random distributions (37.74% against 34.37%).
@@
 
-```{r screeplot, echo = FALSE}
+```{r screeplot, echo = FALSE, fig.cap="Mutiple Correspondence Analysis screeplot."}
 
 # Screeplot
-fviz_screeplot(morpho_data_mca, addlabels = TRUE, ylim = c(0, 15), barfill = "grey50", barcolor = "grey50")
+fviz_screeplot(morpho_data_mca, addlabels = TRUE, ylim = c(0, 15), barfill = "#0B775E", barcolor = "#0B775E")
 
@@
 
-Bipolar technology clearly had an important role on the adaptive systems of the first modern humans in Western Iberia, and elsewhere [@villa_neandertals_2018].
+Bipolar technology clearly had an important role on the adaptive systems of the first modern humans in Western Iberia, as well as in other european regions [see e.g.
+@villa_neandertals_2018].
 A rise in bipolar technologies cannot be dissociated from the diverse set of factors that made Anatomically Modern Humans thrive.
@@
 
-In addition, scaled pieces in Vale Boi may also be associated to bone tool production, given that Vale Boi is one of the portuguese Upper Paleolithic sites with significant evidence for the onsite production and use of bone tools [@evora_osseous_2013].
+In addition, scaled pieces in Vale Boi may also be associated to bone tool production, given that Vale Boi is one of the portuguese Upper Paleolithic sites with significant evidence for onsite production and use of bone tools [@evora_osseous_2013].
 According to both Leblanc [-@leblanc_wedges_1992] and Shott [-@shott_bipolar_1999] wedges are needed to work bone, antler, and wood, and stone wedges are preferable than other types of material.
-For hunter-gatherers, these artifacts would provide multifunctionality from a single piece that could be continuously used until losing their morphological characteristics.
+For hunter-gatherers, these artifacts would provide multifunctionality from a single piece that could be continuously used until losing their optimal morphological characteristics.
 
-Importantly, our results show that the use of scale pieces at Vale Boi reveals techno-functional patterns that seem to stay fairly similar across all paleolithic horizons.
+Importantly, our results show that the use of scale pieces at Vale Boi reveals techno-functional patterns that seem to stay fairly similar across all Upper Paleolithic horizons.
 One relevant trend is the fact that while quartz pieces were only explored in a single axis, chert pieces were, sometimes, explored in multiple axes.
@@
 Finally, as previously mentioned, Vale Boi stone tool technological analyses revealed that every phase of the chaîne operatoire is present, which means that knapping was mostly occurring at the site, which is attested by the high number of chips found.
-Considering that most knapping activities originate chipage, the need of a specific tool for the sole purpose of extracting them does not seem viable to us.
+Considering that most knapping activities originate chipage, the need of a specific tool for the sole purpose of extracting them does not seem viable.
 By putting all these factors in context, the hypothesis that these pieces were used for carcass processing, bone marrow extracting [@manne_intensive_2012] and other similar activities, such as working hard, organic materials with the goal of producing tools seems, thus far, to be the most reasonable interpretation.
@@
 Further, of the three bipolar cores in all sites, 2 of these come from Vale Boi.
-While currently there are no definitive data on the representation of scaled pieces within the retouched tool assemblage of the Slope area, it is fairly safe to assume that it would have a particularly high value.
+While currently there are no definitive data on the representation of scaled pieces within the retouched tool assemblage of the Slope area, it is fairly safe to assume that it should be a particularly high value.
 
@@
 Higher degrees of representation of scaled pieces in sites can be originated by factors such as site function, but also by cultural patterns.
-Regarding site functionality, if we look at the *loci* with higher counts of scaled pieces (Vale Boi’s Terrace and Slope and Fonte Santa), both are open-air residential sites.
-One example outside of this region is the Southwestern German Magdalenian site of Gönnersdorf, where according to Sano (2012) scaled pieces were the second largest assemblage from the site’s retouched utensils (257 from a total of 1501 utensils), corresponding to c.
-17,12% of the total retouched tools assemblage.
-This typological index is as high as both Vale Boi’s and Fonte Santa’s and like both of these sites, Gönnersdorf is an open-air residential site.
-However, these data can be misleading, as Casal do Felipe (5,91%), Terra do Manuel 1940-42 (1,03%), Terra do Manuel 1988-89 (2,13%) and Vale Comprido – Encosta (0,9%) are also open-air multifunctional sites.
-These sites show the same functionality yet, much smaller representations of scaled pieces.
+Regarding site functionality, if we look at the *loci* with higher counts of scaled pieces (Vale Boi’s Terrace and Slope and Fonte Santa), all are open-air residential sites.
+One example outside of this region is the Magdalenian site of Gönnersdorf in Southwestern Germany, where according to Sano (2012) scaled pieces were the second most frequent tool types within the site’s retouched tools assemblage (257 from a total of 1501 utensils, corresponding to c.
+17,12%).
+This frequency is as high as both Vale Boi’s and Fonte Santa’s, and like both of these sites Gönnersdorf is an open-air residential site.
+However, these data can be misleading, as Casal do Felipe (5,91%), Terra do Manuel 1940-42 (1,03%), Terra do Manuel 1988-89 (2,13%) and Vale Comprido – Encosta (0,9%) are also open-air multifunctional sites with much smaller representations of scaled pieces.
 
@@
 # Acknowledgements
+We would like to thank a diversity of institutions that provided funding to the work in Vale Boi in the last decade.
+These are Archaeological Association of Algarve, Fundaç~ao para a Ciência e Tecnologia, Instituto Portugu^es de Arqueologia, National Geographic Society, and Wenner-Gren Foundation for Anthropological Research.
 
@@
 ```
+
+<!--chapter:end:paper.Rmd-->
+
+---
+title: "SOM: The role of lithic bipolar technology in Western Iberia’s Upper Paleolithic: the case of Vale Boi (southern Portugal)"
+author: 
+  - Pedro Horta (ICArEHB, University of Algarve, Faro, Portugal) pedroehorta@gmail.com
+  - João Cascalheira (ICArEHB, University of Algarve, Faro, Portugal) jmcascalheira@ualg.pt
+  - Nuno Bicho (ICArEHB, University of Algarve, Faro, Portugal) nbicho@ualg.pt
+date: '`r Sys.Date()`'
+output:
+   bookdown::word_document2:
+    fig_caption: yes
+    reference_docx: templates/template2.docx
+bibliography: ScaledPiecesVB.bib
+csl: "../templates/springer-socpsych-author-date.csl"
+---
+
+```{r setup, include=FALSE}
+
+knitr::opts_chunk$set(
+  collapse = TRUE,
+  warning = FALSE,
+  message = FALSE,
+  echo = FALSE,
+  comment = "#>",
+  dpi = 300,
+  fig.width = 9,
+  fig.path = "../figures/")
+  cache = TRUE
+
+library(knitr)
+library(readr)
+
+```
+
+## Technological attributes used in the analysis
+
+```{r techno_attributes, echo=FALSE}
+
+techno_attributes <- read_csv("../data/raw_data/technological_attributes.csv")
+
+techno_attributes[is.na(techno_attributes)] <- " "
+
+knitr::kable(techno_attributes, caption= "Technological attributes used in the analysis of scaled pieces included in the study.")
+
+```
+
+\newline
+
+## Morpho-functional attributes used in the analysis
+
+```{r morpho_attributes, echo=FALSE}
+
+morpho_attributes <- read_csv("../data/raw_data/functional_attributes.csv")
+
+morpho_attributes[is.na(morpho_attributes)] <- " "
+
+knitr::kable(morpho_attributes, caption= "Morpho-functional attributes used in the analysis of each platform of scaled pieces included in the study.
+*Adapted from Gonzalez-Urquijo and Ibanez-Estévez [-@gonzalez-urquijo_metodologialisis_1994].
+**Adapted from de la Peña [-@de_la_pena_sobre_2011].")
+
+```
+
+\newline
+
+# References 
+<div id="refs"></div>
+
+# Colophon
+
+This report was generated on `r Sys.time()` using the following computational environment and dependencies:
+
+```{r colophon, cache = FALSE}
+# which R packages and versions?
+devtools::session_info()
+```
+
+The current Git commit details are:
+
+```{r}
+# what commit is this file at? You may need to change the path value
+# if your Rmd is not in analysis/paper/
+git2r::repository("../..")
+```
+
+<!--chapter:end:SOM.Rmd-->

### 2018-10-04 R/functions.R

@@ -91,10 +91,12 @@ mean.plot <- function(dataset, x, y, z){
 
   ggbarplot(dataset, x = x, y = y,
             add = c("mean_sd", "jitter"), size = 1,
-            color = z, palette = c("#00AFBB", "#E7B800", "#FC4E07"),
+            color = z,
             position = position_dodge(0.8),
             xlab = "") +
-  scale_x_discrete(limits=c("Gravettian", "Proto-Solutrean","Solutrean", "Magdalenian"))
+  scale_x_discrete(limits=c("Gravettian", "Proto-Solutrean","Solutrean", "Magdalenian")) +
+  scale_fill_manual(values = wes_palette(name="Rushmore")[c(4,1,3)]) +
+  scale_color_manual(values=wes_palette(name="Rushmore")[c(4,1,3)])
 
 }
 
@@ -108,10 +110,11 @@ box.plot <- function(df, x_var, y_var, x_label, y_label){
 
   b_plot <- b_plot +
     geom_jitter(position=position_jitter(width=.2, height=0), aes(color = RawMaterial), size = 2) +
-    scale_color_manual(values = c("#00AFBB", "#E7B800", "#FC4E07"), name = "Raw Material") +
     labs(x = x_label, y = y_label)+
     scale_fill_discrete(guide=FALSE)+
-    theme_classic()
+    theme_classic() +
+    scale_fill_manual(values = wes_palette(name="Rushmore")[c(5,3,1)]) +
+    scale_color_manual(values=wes_palette(name="Rushmore")[c(5,3,1)])
 
   return(b_plot)
 }
@@ -358,10 +361,11 @@ plot.morpho.var <- function(x){
   dist <- ggbarplot(dist_perc, x = "ScarDistribution", y = "pct",
                     position = position_dodge(0.9),
                     fill = "RawMaterial", color = "RawMaterial",
-                    palette = c("#00AFBB", "#E7B800"),
                     orientation = "horiz",
                     xlab = FALSE,
-                    ylab = FALSE)
+                    ylab = FALSE) +
+    scale_fill_manual(values = wes_palette(name="Rushmore")[c(1,3)]) +
+    scale_color_manual(values=wes_palette(name="Rushmore")[c(1,3)])
 
 
 
@@ -374,10 +378,11 @@ plot.morpho.var <- function(x){
   arr <- ggbarplot(arr_perc, x = "ScarArrangement", y = "pct",
                     position = position_dodge(0.9),
                     fill = "RawMaterial", color = "RawMaterial",
[…]

### 2018-10-07 analysis/paper/paper.Rmd
submitted

@@
 author:
-  - Pedro Horta (ICArEHB, University of Algarve, Faro, Portugal) pedroehorta@gmail.com
-  - João Cascalheira (ICArEHB, University of Algarve, Faro, Portugal) jmcascalheira@ualg.pt
-  - Nuno Bicho (ICArEHB, University of Algarve, Faro, Portugal) nbicho@ualg.pt
-date: '`r Sys.Date()`'
+  - Pedro Horta* (ICArEHB, University of Algarve, Faro, Portugal)
+  - João Cascalheira (ICArEHB, University of Algarve, Faro, Portugal)
+  - Nuno Bicho (ICArEHB, University of Algarve, Faro, Portugal)
+  - "*corresponding author | email: pedroehorta@gmail.com | phone: +351 911 005 506"
+
 output: 
@@
 bibliography: ScaledPiecesVB.bib
-csl: "../templates/springer-humanities-author-date.csl"
-abstract: |
+csl: "../templates/springer-socpsych-author-date.csl"
+---
+
+\newline
+
+# Abstract
+
   Scaled or splintered pieces are one of the most common lithic artifacts type in Upper Paleolithic assemblages throughout Europe, especially in its westernmost regions.
-Despite this, even after one century of being identified there is still no consensus on how to define, analyze or even interpret these tools.
+Despite this, and even after one century of being identified there is still no consensus on how to define, analyze or interpret these tools.
 In western Iberia there is a clear lack of comprehensive studies regarding this type of artifacts at a regional scale.
@@
 
----
+\newline
+
+# Keywords
+
+Stone tools; Bipolar technology; Upper Paleolithic; Western Europe
 
@@
 
+#Please see the functions.R file in /R directory for other libraries used for this compendium
+
 library(knitr)
+library(citr)
 library(tidyverse)
@@
 library(epitools)
+library(wesanderson)
+
 #library(ScaledPiecesVB)
@@
 
-```{r VBLocation, echo = FALSE, fig.cap="Location of the site of Vale Boi"}
+```{r VBLocation, echo = FALSE, fig.cap="Location of the site of Vale Boi. Map data are from stamenmap, using the ggmap package Kahle and Wickham [-@kahle_ggmap:_2013]"}
 
@@
 
-knitr::include_graphics("../figures/VBMap.png")
+knitr::include_graphics("../figures/location_map.png")
 
@@
 All analyses and data processing were accomplished in R (version `r getRversion()`) [@r_core_team_r:_2013].
-Following recent concerns on the reproducibility of archaeological analysis we include the entire R code used for all the analysis and visualizations contained in this paper in our supplemental online material (SOM) at https://dx.doi.org/… To produce those files we followed the procedures described by Marwick et al.
+Following recent concerns on the reproducibility of archaeological analysis we include the entire R code used for all the analysis and visualizations contained in this paper in our supplemental online material (SOM) at https://dx.doi.org/10.17605/OSF.IO/WPXGH To produce those files we followed the procedures described by Marwick et al.
 [-@marwick_packaging_2017] for the creation of research compendiums to enhance the reproducibility of research.
@@
 
-assemb <- cross.tb(techno_data, "RawMaterial", var_list)
+assemb <- as.tibble(cross.tb(techno_data, "RawMaterial", var_list))
 
-knitr::kable(assemb, caption= "Frequencies of scaled pieces by raw material and chronology")
+assemb <- assemb %>%
+  select(" " = "Variable", "Chert", "Quartz", "Chalcedony", "Total" = "Overall") %>% 
+  slice(2:5)
+
+assemb <- assemb[c(1,3,4,2),]
+
+knitr::kable(assemb, caption= "Frequencies of scaled pieces used in this study, by raw material and chronological period. Percentages are shown in parentheses")
 
@@
 
-knitr::kable(Gravettian, caption= "Technological attribute frequencies by raw materials for the Gravettian sample. Percentages are shown in parentheses.")
+knitr::kable(Gravettian, caption= "Technological attributes frequencies by raw materials for the Gravettian sample. Percentages are shown in parentheses.")
 
@@
 
-knitr::kable(`Proto-Solutrean`, caption= "Technological attribute frequencies by raw materials for the Proto-Solutrean sample. Percentages are shown in parentheses.")
+knitr::kable(`Proto-Solutrean`, caption= "Technological attributes frequencies by raw materials for the Proto-Solutrean sample. Percentages are shown in parentheses.")
 
@@
 
-knitr::kable(Solutrean, caption= "Technological attribute frequencies by raw materials for the Solutrean sample. Percentages are shown in parentheses.")
+knitr::kable(Solutrean, caption= "Technological attributes frequencies by raw materials for the Solutrean sample. Percentages are shown in parentheses.")
 
@@
 
-knitr::kable(Magdalenian, caption= "Technological attribute frequencies by raw materials for the Magdalenian sample. Percentages are shown in parentheses.")
+knitr::kable(Magdalenian, caption= "Technological attributes frequencies by raw materials for the Magdalenian sample. Percentages are shown in parentheses.")
 
@@
 
-```{r meanplots, echo = FALSE, fig.cap="Barplots of means for Area (Length x Width) and Thickness of scaled pieces by raw material and across the four chronological phases.
+```{r meanplots, echo = FALSE, fig.cap="Barplots of means for Area (Length x Width) and Thickness of scaled pieces, by raw material and across the four chronological phases.
 Error bars represent standard deviations."}
@@
 
-```{r plotbyplatforms, echo = FALSE, fig.cap="Boxplot of Area and Thickness distribution for each raw material."}
+```{r plotbyplatforms, echo = FALSE, fig.cap="Boxplot of Area (Length x Width) and Thickness distribution for each raw material."}
 
@@
 ```{r morphoattributes, echo = FALSE, fig.cap="Frequency of morphological attributes for each raw material.
-Damaged platforms were grouped so that each artifact was only counted once and to avoid wrong comparisons between active a hammered platforms.
+Opposed damaged platforms were grouped so that each artifact was only counted once and to avoid wrong comparisons between active a hammered platforms.
 A - Distribution of damage; B - Arrangement of scars; C - Extension of scars; D - Delineation of damaged platforms; E - Facial distribution of scars; F - Angle of damaged platforms."}
@@
 
-knitr::kable(scaledpiecesportugal, caption= "Frequencies of scaled pieces in portuguese Upper Paleolithic sites")
+knitr::kable(scaledpiecesportugal, caption= "Frequencies of scaled pieces and bipolar cores in portuguese Upper Paleolithic sites.")
 
@@
 # Acknowledgements
+We would like to thank a diversity of institutions that provided funding to the work in Vale Boi in the last decade.
+These are Archaeological Association of Algarve, Fundaç~ao para a Ciência e Tecnologia, Instituto Português de Arqueologia, National Geographic Society, and Wenner-Gren Foundation for Anthropological Research.
 

### 2018-10-07 R/functions.R

@@ -21,43 +21,68 @@ Map <- function(){
   require(maptools)
   require(ggplot2)
   require(ggmap)
-  require(devtools)
   require(legendMap)
+  require(gridExtra)
 
-  VBLocation <- c(lon = -8.808621, lat = 37.089902)
+  VBLocation <-  c(lon = -8.808621, lat = 37.089902)
 
-  VBMap <- get_googlemap(center = c(lon = -8.101284, lat = 39.486585),
-                         zoom = 7,
-                         scale = 2,
-                         maptype = "roadmap")
+  western_algarve <- c(left = -9.3, bottom = 36.8, right = -7.5, top = 38)
+  myMap <- get_stamenmap(western_algarve, zoom = 10, maptype = "terrain")
 
-  VBMap <-
-    ggmap(VBMap) +
+  close_up_map <- ggmap(myMap) +
     geom_point(data = data.frame(t(VBLocation)),
                aes(lon,
                    lat),
-               size = 5,
-               colour = "black") +
+               size = 10,
+               colour = "red") +
     geom_text(data = data.frame(t(VBLocation)), aes(x = lon, y = lat, label = "Vale Boi"),
-              size = 4, colour = "black", vjust = -1, hjust = 0.5)
+              size = 5, colour = "black", vjust = -1.5, hjust = 0.5)
 
-
-  VBMap <- VBMap +
-    legendMap::scale_bar(lon = -11.3,
-                         lat = 37,
-                         distance_lon = 50,
-                         distance_lat = 8,
-                         distance_legend = 20,
-                         legend_colour = "white",
+  close_up_map <- close_up_map +
+    legendMap::scale_bar(lon = -9.2,
+                         lat = 36.85,
+                         distance_lon = 10,
+                         distance_lat = 3,
+                         distance_legend = 6,
                          dist_unit = "km",
                          orientation = TRUE,
-                         arrow_length = 50,
-                         arrow_distance = 50,
-                         arrow_north_size = 3)
+                         arrow_length = 10,
+                         arrow_distance = 110,
+                         arrow_north_size = 5)
 
-  ggsave("../figures/VBMap.png")
+  iberia <- c(left = -12, bottom = 35, right = 2, top = 44)
+  myMap <- get_stamenmap(iberia, zoom = 7, maptype = "terrain")
+
+  general_map <- ggmap(myMap) +
+    geom_point(data = data.frame(t(VBLocation)),
+               aes(lon,
+                   lat),
+               size = 6,
+               colour = "red") +
[…]

### 2018-12-02 README.Rmd
first changes after reviews

@@
 _The role of lithic bipolar technology in Western Iberia’s Upper Paleolithic: the case of Vale Boi (southern Portugal)_.
-Name of journal/book <https://doi.org/xxx/xxx>
+Journal of Paleolithic Archaeology <https://doi.org/xxx/xxx>
 
@@
 _The role of lithic bipolar technology in Western Iberia’s Upper Paleolithic: the case of Vale Boi (southern Portugal)_.
-Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`.
+SocArXiv , Accessed `r format(Sys.Date(), "%d %b %Y")`.
 Online at <https://doi.org/xxx/xxx>
@@
 
-This repository is organized as an R package using rrtools by Ben Marwick which can be installed from github with:
+This repository is organized as an R package using rrtools by [Ben Marwick](https://github.com/benmarwick) which can be installed from github with:
+
+```{r eval=FALSE}
+
+git clone https://github.com/benmarwick/rrtools.git
+
+```
+
+The functions are provided as a package because this makes it simpler to reuse the functions many times in the paper.
+It also makes it easier for others to use and adapt these functions on their own data.
+Nevertheless, this package has been written explicitly for this project and may not yet be suitable for more general purpose use.
 
@@
 Once the download is complete, open the ScaledPieces.Rproj in RStudio to begin working with the package and compendium files.
+
+If you want to re-run all the analyses in R, you can start by installing the compendium package with this line at the R prompt:
+
+```{r eval = FALSE}
+
+devtools::install_github("jmcascalheira/ScaledPiecesVB")
+
+```
 

### 2018-12-02 analysis/paper/paper.Rmd
first changes after reviews

@@
 output: 
-    bookdown::word_document2: #subsistute by pdf_document2 or hmtl_document2 as necessary
+    bookdown::word_document2: #substitute by pdf_document2 or hmtl_document2 as necessary
       fig_caption: yes
@@
 
-source('~/R_DATA/compendiums/ScaledPiecesVB/R/functions.R')
-
 ```
@@
 library(wesanderson)
-
-#library(ScaledPiecesVB)
+library(ScaledPiecesVB)
 

### 2018-12-02 R/functions.R

@@ -2,6 +2,13 @@
 ###########################################################################
 # READ DATA ----------------------------------------------------------------
 
+
+#' read.data
+#' @importFrom readr read_csv
+#' @export
+#'
+
+
 read.data <- function(){
   #tech_attributes <- readcsv("../data/raw_data/technology_attributes.csv")
   #morpho_attributes <- read.csv("../data/raw_data/morpho_attributes.csv")
@@ -15,6 +22,17 @@ read.data <- function(){
 
 ###########################################################################
 # MAPS --------------------------------------------------------------------
+
+#' Map
+#'
+#' @importFrom ggmap    ggmap    get_stamenmap
+#' @import legendMap
+#' @import maptools
+#' @importFrom gridExtra    grid.arrange
+#' @importFrom grDevices    dev.off    png
+#' @importFrom stats    aggregate    coef    lm    na.omit    quantile    reshape
+#' @importFrom utils    read.csv    read.table
+#'@export
 
 Map <- function(){
 
@@ -88,6 +106,15 @@ Map <- function(){
 ###########################################################################
 # CROSS TABLES ------------------------------------------------------------
 
+#' cross.tb
+#'
+#' @importFrom tab    tabmulti
+#' @importFrom dplyr    mutate_if
+#'@export
+
+
+
+
 cross.tb <- function(dataset,x,y) {
 
 require(tab)
@@ -112,6 +139,13 @@ dataset <- as.data.frame(dataset)
 ###########################################################################
 # MEAN PLOTS --------------------------------------------------------------
 
+
+#' mean.plot
+#'
+#' @importFrom ggpubr ggbarplot
+#'@export
+
+
 mean.plot <- function(dataset, x, y, z){
 
   ggbarplot(dataset, x = x, y = y,
@@ -128,6 +162,13 @@ mean.plot <- function(dataset, x, y, z){
 
 ###########################################################################
 # BOXPLOT WITH JITTER -----------------------------------------------------
+
+
+#' box.plot
+#'
+#' @importFrom ggplot2    ggplot    geom_boxplot
+#'@export
+
 
 box.plot <- function(df, x_var, y_var, x_label, y_label){
   b_plot<-ggplot(df,aes_string(x= x_var, y=y_var)) +
@@ -149,6 +190,12 @@ box.plot <- function(df, x_var, y_var, x_label, y_label){
[…]

### 2018-12-03 analysis/paper/paper.Rmd
small changes in load.libraries function

@@
 
-#Please see the functions.R file in /R directory for other libraries used for this compendium
+library(ScaledPiecesVB)
 
-library(knitr)
-library(citr)
-library(tidyverse)
-library(broom)
-library(FactoMineR)
-library(factoextra)
-library(ggpubr)
-library(ggplot2)
-library(ggrepel)
-library(maptools)
-library(ggmap)
-library(devtools)
-library(legendMap)
-library(tab)
-library(sjstats)
-library(powerAnalysis)
-library(epitools)
-library(wesanderson)
-library(ScaledPiecesVB)
+load.packages()
 

### 2018-12-03 R/functions.R

@@ -1,5 +1,31 @@
 
 ###########################################################################
+# LOAD PACKAGES ----------------------------------------------------------
+
+load.packages <- function {
+  library(knitr)
+  library(citr)
+  library(dplyr)
+  library(readr)
+  library(tibble)
+  library(broom)
+  library(FactoMineR)
+  library(factoextra)
+  library(ggpubr)
+  library(ggplot2)
+  library(ggrepel)
+  library(maptools)
+  library(ggmap)
+  library(devtools)
+  library(legendMap)
+  library(tab)
+  library(sjstats)
+  library(powerAnalysis)
+  library(epitools)
+  library(wesanderson)
+}
+
+###########################################################################
 # READ DATA ----------------------------------------------------------------
 
 

### 2018-12-03 R/functions.R
small changes in functions.R

@@ -2,7 +2,7 @@
 ###########################################################################
 # LOAD PACKAGES ----------------------------------------------------------
 
-load.packages <- function {
+load.packages <- function() {
   library(knitr)
   library(citr)
   library(dplyr)

### 2018-12-03 R/functions.R
functions description added

@@ -2,6 +2,17 @@
 ###########################################################################
 # LOAD PACKAGES ----------------------------------------------------------
 
+
+#' load.packages
+#'
+#' Load all necessary libraries
+#'
+#' @example
+#' load.packages()
+#'
+#' @export
+
+
 load.packages <- function() {
   library(knitr)
   library(citr)
@@ -30,9 +41,13 @@ load.packages <- function() {
 
 
 #' read.data
-#' @importFrom readr read_csv
-#' @export
 #'
+#' Read in csv files
+#'
+#' @example
+#' read.data()
+#'
+#' @export
 
 
 read.data <- function(){
@@ -51,22 +66,14 @@ read.data <- function(){
 
 #' Map
 #'
-#' @importFrom ggmap    ggmap    get_stamenmap
-#' @import legendMap
-#' @import maptools
-#' @importFrom gridExtra    grid.arrange
-#' @importFrom grDevices    dev.off    png
-#' @importFrom stats    aggregate    coef    lm    na.omit    quantile    reshape
-#' @importFrom utils    read.csv    read.table
-#'@export
+#' Saves a PNG image of a close-up and a general map of Vale Boi
+#'
+#' @example
+#' Map()
+#'
+#' @export
 
 Map <- function(){
-
-  require(maptools)
-  require(ggplot2)
-  require(ggmap)
-  require(legendMap)
-  require(gridExtra)
 
   VBLocation <-  c(lon = -8.808621, lat = 37.089902)
 
@@ -134,17 +141,13 @@ Map <- function(){
 
 #' cross.tb
 #'
-#' @importFrom tab    tabmulti
-#' @importFrom dplyr    mutate_if
+#' Create cross-tables with counts and percentages inside parentheses
+#'
+#'
 #'@export
 
 
-
-
 cross.tb <- function(dataset,x,y) {
-
-require(tab)
-require(tidyverse)
 
 dataset <- dataset %>% mutate_if(is.character,as.factor)
 dataset <- as.data.frame(dataset)
@@ -168,7 +171,8 @@ dataset <- as.data.frame(dataset)
 
 #' mean.plot
 #'
-#' @importFrom ggpubr ggbarplot
+#' Creates a barplot representing the mean, standard deviations and jitter points
+#'
 #'@export
 
 
@@ -192,7 +196,9 @@ mean.plot <- function(dataset, x, y, z){
 
 #' box.plot
 #'
[…]

### 2018-12-03 R/functions.R
small changes in functions.R

@@ -400,6 +400,13 @@ run.Anova<-function(df, y_var){
   return(compare_aov)
 }
 
+
+#' anova.test
+#'
+#' ANOVA test
+#'
+#'@export
+
 anova.test <- function(df, y_var){
   compare_aov = run.Anova(df, y_var)
   return(compare_aov)

### 2018-12-03 R/functions.R
fix1 anova.test problem

@@ -401,9 +401,14 @@ run.Anova<-function(df, y_var){
 }
 
 
+# ANOVA test (cont.)
+
 #' anova.test
 #'
-#' ANOVA test
+#' ANOVA test (cont.)
+#'
+#' @example
+#' anova.test()
 #'
 #'@export
 

### 2018-12-03 analysis/paper/paper.Rmd
fix2 anova test

@@
 
-aov_area <- anova.test(techno_data, "Area")
+aov_area <- aov.test(techno_data, "Area")
 cohens_area <- cohens.test(aov_area)
@@
 
-aov_thickness <- anova.test(techno_data, "Thickness")
+aov_thickness <- aoa.test(techno_data, "Thickness")
 cohens_thickness <- cohens.test(aov_thickness)

### 2018-12-03 R/functions.R

@@ -403,16 +403,16 @@ run.Anova<-function(df, y_var){
 
 # ANOVA test (cont.)
 
-#' anova.test
+#' aov.test
 #'
 #' ANOVA test (cont.)
 #'
 #' @example
-#' anova.test()
+#' aov.test()
 #'
 #'@export
 
-anova.test <- function(df, y_var){
+aov.test <- function(df, y_var){
   compare_aov = run.Anova(df, y_var)
   return(compare_aov)
 }

### 2018-12-03 analysis/paper/paper.Rmd
fix mean.plot problem

@@
 
-aov_thickness <- aoa.test(techno_data, "Thickness")
+aov_thickness <- aov.test(techno_data, "Thickness")
 cohens_thickness <- cohens.test(aov_thickness)
@@
 
-area <- mean.plot(techno_data, "Chronology", "Area", "RawMaterial")
-thickness <- mean.plot(techno_data, "Chronology", "Thickness", "RawMaterial")
+area <- mean..bar.plot(techno_data, "Chronology", "Area", "RawMaterial")
+thickness <- mean..bar.plot(techno_data, "Chronology", "Thickness", "RawMaterial")
 

### 2018-12-03 R/functions.R

@@ -166,17 +166,18 @@ dataset <- as.data.frame(dataset)
 }
 
 ###########################################################################
-# MEAN PLOTS --------------------------------------------------------------
+# MEAN PLOT --------------------------------------------------------------
 
 
-#' mean.plot
+#' mean.bar.plot
 #'
-#' Creates a barplot representing the mean, standard deviations and jitter points
+#' Creates a barplot representing the mean, standard deviations
+#' and jitter points
 #'
-#'@export
+#' @export
 
 
-mean.plot <- function(dataset, x, y, z){
+mean.bar.plot <- function(dataset, x, y, z){
 
   ggbarplot(dataset, x = x, y = y,
             add = c("mean_sd", "jitter"), size = 1,

### 2018-12-03 analysis/paper/paper.Rmd
bar.plot.mean

@@
 
-area <- mean..bar.plot(techno_data, "Chronology", "Area", "RawMaterial")
-thickness <- mean..bar.plot(techno_data, "Chronology", "Thickness", "RawMaterial")
+area <- bar.plot.mean(techno_data, "Chronology", "Area", "RawMaterial")
+thickness <- bar.plot.mean(techno_data, "Chronology", "Thickness", "RawMaterial")
 

### 2018-12-03 R/functions.R

@@ -169,7 +169,7 @@ dataset <- as.data.frame(dataset)
 # MEAN PLOT --------------------------------------------------------------
 
 
-#' mean.bar.plot
+#' bar.plot.mean
 #'
 #' Creates a barplot representing the mean, standard deviations
 #' and jitter points
@@ -177,7 +177,7 @@ dataset <- as.data.frame(dataset)
 #' @export
 
 
-mean.bar.plot <- function(dataset, x, y, z){
+bar.plot.mean <- function(dataset, x, y, z){
 
   ggbarplot(dataset, x = x, y = y,
             add = c("mean_sd", "jitter"), size = 1,

### 2018-12-03 R/functions.R
morpho plot

@@ -527,13 +527,13 @@ opposed.plat.morpho <- function(){
 ###########################################################################
 # PLOT ALL VARIABLES ---------------------------------------
 
-#' plot.morpho.var
+#' morpho.var.plot
 #'
 #' Horizontal bar plots for frequency of each variable
 #'
 #'@export
 
-plot.morpho.var <- function(x){
+morpho.var.plot <- function(x){
 
   dist_perc <- x %>%
     group_by(RawMaterial, ScarDistribution) %>%

### 2018-12-03 analysis/paper/paper.Rmd
morpho plot
inline statistics change

@@
 
+cohens_area$cohens.f
+
 ```
@@
 In general, mean values tend to be similar between raw materials, with some differences occurring within the Proto-Solutrean and Magdalenian assemblages, most certainly as a result of the small samples analysed for each of these periods (Figure \@ref(fig:meanplots)).
-Across techno-complexes, however, no significant statistical differences were detected when using an ANOVA test for both Area (*F* (`r aov_area$df[1]`, `r aov_area$df[2]`) = `r aov_area$statistic[1]`, *p* = `r aov_area$p.value[1]`, *d* = `r cohens_area$cohens.f`), and Thickness (*F* (`r aov_thickness$df[1]`, `r aov_thickness$df[2]`) = `r aov_thickness$statistic[1]`, *p* = `r aov_thickness$p.value[1]`, *d* = `r cohens_thickness$cohens.f`).
+Across techno-complexes, however, significant statistical differences were detected when using an ANOVA test for Thickness (*F* (`r aov_thickness$df[1]`, `r aov_thickness$df[2]`) = `r aov_thickness$statistic[1]`, *p* = `r aov_thickness$p.value[1]`, *d* = `r cohens_thickness$cohens.f`), but not for Area (*F* (`r aov_area$df[1]`, `r aov_area$df[2]`) = `r aov_area$statistic[1]`, *p* = `r aov_area$p.value[1]`, *d* = `r cohens_area$cohens.f`).
 
@@
 A chi-square test with modified alpha level (Bonferroni correction) to adjust for multiple testing and reduce type-I error, shows significant differences in the quartz and chert Scar Arrangement categories (*X^2^* [`r chi_arrangement$chi_test$parameter`, *N* = `r sum(chi_arrangement$chi_test$observed)`] = `r  round(chi_arrangement$chi_test$statistic, 2)`, *p* = `r round(chi_arrangement$chi_test$p.value, 3)`, *phi* = `r round(chi_arrangement$chi_size_effect$phi, 2)`), and in Scar Faciality categories (*X^2^* [`r chi_faciality$chi_test$parameter`, *N* = `r sum(chi_faciality$chi_test$observed)`] = `r  round(chi_faciality$chi_test$statistic,2)`, *p* = `r round(chi_faciality$chi_test$p.value, 3)`, *phi* = `r round(chi_faciality$chi_size_effect$phi, 2)`).
-The effect size statistic (Phi) suggests, however, a low practical significance for Scar Arrangement and a large pratical significance for Scar Faciality.
+The effect size statistic (Phi) suggests, however, a medium practical significance for both variables.
 
@@
 
-plot.morpho.var(morpho_data_opposed_platforms)
+morpho.var.plot(morpho_data_opposed_platforms)
 

### 2018-12-03 R/functions.R

@@ -396,7 +396,7 @@ CHI <- function(x, y) {
 # ANOVA test
 
 run.Anova<-function(df, y_var){
-  frm<-as.formula(sprintf("%s~%s", y_var, "RawMaterial"))
+  frm<-as.formula(sprintf("%s~%s", y_var, "Chronology"))
   compare_aov<-aov(frm, data=df)
   return(compare_aov)
 }

### 2018-12-04 analysis/paper/paper.Rmd
small changes in ANOVA and CHISquare statistics

--- 2018-12-03 analysis/paper/paper.Rmd
@@
 
+aov_thickness$p.value
+
 ```

### 2018-12-09 analysis/paper/paper.Rmd
data dictionaries altered

@@
 output: 
-    bookdown::word_document2: #substitute by pdf_document2 or hmtl_document2 as necessary
+    bookdown::word_document2:
       fig_caption: yes
@@
 
-Bipolar technology is generically classified into two types of lithic artifacts - bipolar cores and scaled pieces (also known as splintered pieces or *pièce esquilée*).
+Bipolar technology is generically classified into two types of lithic artifacts - bipolar cores and scaled pieces (also known as splintered pieces or *pièce esquilée*), this latter being the main focus of this paper.
 Their distinction, however, has not always been consensual [@hayden_confusion_1980].
-Besides the fact that bipolar cores are often confused or lumped with scaled pieces [@villa_neandertals_2018], the definition of scaled piece has suffered significant changes ever since its initial identification in the early 20th century and, to this day, there still seems to be no clear definition for this type of artifact.
+Besides the fact that bipolar cores are often confused or lumped with scaled pieces [@villa_neandertals_2018;@de_la_pena_sobre_2011], the definition of scaled piece has suffered significant changes ever since its initial identification in the early 20th century and, to this day, there still seems to be no clear definition for this type of artifact.
 The first definition of scaled pieces was proposed by Bardon and Bouyssonie [-@bardon_outils_1906], describing them as a result of bipolar knapping through direct percussion, with the “core” rested in a hard surface, originating splintering in both ends of the tool.
@@
 The issue surrounding this problem lies within the functional equifinality of these artifacts.
-Contrary to bipolar cores that are most of the times easily identified as cores, scaled pieces have been associated with two distinct types of activities: (1) as intermediate pieces or wedges for working hard raw materials (e.g.
-bone, ivory, antler); and (2) as cores for the extraction of chips and small bladelets [@brantingham_early_2004].
+Contrary to bipolar cores that are exclusively part of a technological reduction sequence with the goal of blank extraction, scaled pieces have been associated with two distinct types of activities: (1) as intermediate pieces or wedges for working hard organic raw materials (e.g.
+bone, ivory, antler); and (2) as cores for the extraction of chips, small flakes and bladelets [@brantingham_early_2004].
 Although largely debated, the ambiguity of this classification has, however, been ignored in some of the most recent literature, with some authors not acknowledging that bipolar evidence in stone tools might also result from other activities other than only reduction strategies [see e.g.
@@
 @bicho_technological_2000; @carvalho_talhe_1998; @zilhao_o_1997].
-While scarce, the majority of portuguese paleolithic studies [e.g.
+While scarce, the majority of Portuguese Paleolithic studies [e.g.
 @almeida_terminal_2000; @goncalves_variabilite_2012] have not presented, so far, any context-specific interpretations for the presence of scaled pieces.
-Most references broadly interpret these as bipolar cores for the extraction of small bladelets and chips to be  used in composite tools [e.g.
+Most references broadly interpret these as bipolar cores for the extraction of small flakes, bladelets and chips to be used in composite tools [e.g.
 @zilhao_o_1997], or as intermediate pieces for working hard materials [@gibaja_tecnologiy_2007; @marreiros_as_2009].
 Zilhão [-@zilhao_o_1997], for example, argues that the presence of scaled pieces throughout the Upper Paleolithic sequence of central Portugal is inversely proportional to the presence of “carinated cores”, and thus it is likely that the former should have worked as a flexible substitute for the latter.
-Carvalho [-@carvalho_talhe_1998] agrees with this interpretation and considers it also valid for the portuguese early Neolithic.
+Carvalho [-@carvalho_talhe_1998] agrees with this interpretation and considers it also valid for the Portuguese early Neolithic.
 
-While we agree that some of the elements might have been used as cores, these interpretations seem overly simplistic, and the authors seem to have little to no evidence to back their claims other than the inverse relationship mentioned above.
+While we agree that some of the elements might have been used as cores, these interpretations seem simplistic, and the authors seem to have little to no analytical evidence to back their hypothesis, other than the inverse relationship mentioned above.
 This results, in part, from the lack of comprehensive studies regarding this type of artifacts at a regional scale.
-Additionally, the contexts from where most of these artifacts were recovered did not have good organic preservation and, thus, no direct association between these tools and the exploitation of hard-organic raw materials is possible, and, in face of the nonexistence of any absolute dates, their precise chronological attribution is also, most of the times, unreliable.
+Additionally, the contexts from where most of these artifacts were recovered did not have good organic preservation or dedicated use-wear studies and, thus, no direct association between these tools and the exploitation of hard organic raw materials is possible, and, in face of the nonexistence of any absolute dates, their precise chronological attribution is also, most of the times, unreliable.
 
@@
 The archaeological site of Vale Boi is located in the western edge of the Algarve region (southern Portugal) (Figure \@ref(fig:VBLocation)).
-The site can be found in a small valley following a river north-south for 2 km until it reaches the Atlantic Ocean.
+The site can be found in a small valley following a river north-south for 2 km, until it reaches the Atlantic Ocean.
 Archaeological deposits occupy an estimated area of over 10 000 sq.
-meters across a stepped slope marked at the top by a 10 meter-high limestone cliff face [@bicho_ledge:_2012; @bicho_paleolithique_2010].
+meters across a stepped slope, marked at the top by a 10 meter-high limestone cliff face [@bicho_ledge_2012; @bicho_paleolithique_2010].
 
@@
 The Terrace area is located in the lower part of the hill.
-In this area, the longest archaeological sequence of the site can be found, including the complete Upper Paleolithic sequence but also three Holocene horizons, corresponding to Neolithic, Mesolithic and Epipaleolithic occupations [@bicho_ledge:_2012; @cascalheira_cross-scale_2017].
+In this area, the longest archaeological sequence of the site can be found, including the complete Upper Paleolithic sequence but also three Holocene horizons, corresponding to Neolithic, Mesolithic and Epipaleolithic occupations [@bicho_ledge_2012; @cascalheira_cross-scale_2017].
 From within the lower levels of this area, an Early Gravettian occupation was discovered, dated to c.
@@
 Like in the previous areas, remains are well preserved, and high frequencies of lithic artifacts, malacological and mammalogical fauna and bone tools were recovered.
-No habitation features were identified in this area and based on the conditions and type of artifacts found, it has been suggested that this area would have mostly functioned as a midden deposit [@bicho_ledge:_2012; @bicho_paleolithique_2010].
+No habitation features were identified in this area and based on the conditions and type of artifacts found, it has been suggested that this area would have mostly functioned as a midden deposit [@bicho_ledge_2012; @bicho_paleolithique_2010].
 
@@
 Chert is the most abundant knapped rock type in Vale Boi.
-Throughout the several occupations, chert was considered the best rock type for knapping, which is evident by the way it was explored with much more complex strategies than either greywacke or quartz.
+Throughout the several occupations, chert was considered the preferred rock type for knapping, which is evident by the way it was explored with much more complex strategies than either greywacke or quartz.
 While it can be said that chert exploration strategies were more elaborate, the most common strategies were still simple unidirectional or bidirectional reduction sequences, producing mostly flakes [@marreiros_flake_2012].
 Elongated products are found at low frequencies across all chronologies.
-When present, retouched tools are mostly notches, denticulates, end-scrapers, and scaled pieces.
+When present, retouched tools are mostly notches, denticulates, end-scrapers, and, although not retouched artifacts *per* *se* scaled pieces.
 One possible reason for these simplistic strategies seems to be an overall low knapping quality, since nodules are quite small, and frequently show a high degree of tectonically derived fractures [@pereira_territory_2016].
@@
 The first is a thick grain, low-quality type, mostly inadequate for knapping.
-Still, this kind of quartz is present in large quantities and is most likely associated with stone boiling and grease rendering activities [@bicho_ledge:_2012; @manne_early_2014; @manne_intensive_2012; @manne_upper_2010; @manne_vale_2009].
+Still, this kind of quartz is present in large quantities and is most likely associated with stone boiling and grease rendering activities [@bicho_ledge_2012; @manne_early_2014; @manne_intensive_2012; @manne_upper_2010; @manne_vale_2009].
 The second type of quartz is characterized by finer-grain small pebbles with yellowish cortex.
@@
 Greywacke is the third most frequent raw material at Vale Boi, showing up in the site mostly in the form of large slabs, in which the identification of concavely shaped impact marks is thought to result from their use as anvils.
-The importance of greywacke anvils is indicated by its high frequency, with hundreds of slabs found in many levels throughout all occupations [@bicho_ledge:_2012; @manne_intensive_2012].
+The importance of greywacke anvils is indicated by its high frequency, with hundreds of slabs found in many levels throughout all occupations [@bicho_ledge_2012; @manne_intensive_2012].
 Greywacke was also knapped for flake extraction, using expedient unidirectional reduction strategies [@cascalheira_influencia_2013; @marreiros_as_2009].
@@
 
-We define scaled pieces as artifacts of variable size and morphology, showing traces of crushing and splitering of edges at opposite ends.
-These traces are caused by direct percussion at one end and subsequent crushing of the other because it is rested on a hard surface, and can be bifacially or unifacially distributed.
-As will become clear from our results, some of the analyzed artifacts did not present splitering/crushing traces in opposed platforms, but we still classified them as scaled pieces.
+Based on previous works [e.g.
+@hayden_confusion_1980] here we define scaled pieces as artifacts of variable size and morphology, showing traces of crushing and splitering of edges at opposite ends, caused by direct percussion at one end, and subsequent crushing of the other for being rested on a hard surface.
+Scaled pieces present always two opposite surfaces, just like a flake or blade would.
+Crushing traces can be bifacially or unifacially distributed.
+Some scaled pieces do not present crushing in opposed platforms, but are still classified as scaled pieces here.
 This detail has been previously referred by Villa et al.
-[-@villa_neandertals_2018] who noted that some edges may present a flat (instead of intensively shattered) platforms.
+[-@villa_neandertals_2018] who noted that some edges may present a flat (instead of intensively shattered) platform.
+Our criteria does not include a particular type of blank because, as we will show in later sections of the paper, these artifacts can originate from flakes, blades or debris.
 
-Attribute analysis was split into two main groups, each corresponding to two distinct types of features: (1) technological attributes and (2) morpho-functional attributes.
+On the contrary, following mostly Hayden [-@hayden_confusion_1980] and Leaf [-@leaf_variation_1979], bipolar cores are not marked by bifacially opposed surfaces.
+Their shape is more blocky and angular, showing evidence of, at least, one flaking surface with two opposed platforms (the striking platform and the base) with typical crushing and flake removal on one or both ends.
+The striking platform is the surface that is struck with a hammerstone to produce blanks.
+Tipically, it exhibits little crushing except near the point of impact.
+Large flake scars tend to originate at the striking platform.
+The base is the surface that rests on the anvil, from where small flake scars can also originate.
+
+All lithic artifacts recovered from the Slope area of Vale Boi that fit the scaled piece definition presented below were considered for this study, independent of raw material or technological class.
+
+Attribute analysis was split into two main groups, each corresponding to two distinct types of features: (1) technological attributes (Table \@ref(tab:techno_attributes)) and (2) morpho-functional attributes (Table \@ref(tab:morpho_attributes)).
 In the first group, a series of variables traditionally used in lithic studies [e.g.
-@andrefsky_jr_lithics:_2005; @inizan_technology_1999] were recorded (see Supplementary Materials for full list of variables), aiming to characterize patterns of blank choice for the application of bipolar technology.
+@andrefsky_jr_lithics:_2005; @inizan_technology_1999] were recorded, aiming to characterize patterns of blank choice for the application of bipolar technology.
 For the second group of variables, a macroscopic approach building upon the work of de la Peña [-@de_la_pena_sobre_2011] and Fischer et.
-al [-@fischer_macro_1984] was adopted (see Supplementary Materials for full list of variables).
+al [-@fischer_macro_1984] was adopted.
 Studies by Sano [-@sano_functional_2012], and Gibaja et al.
@@
 This, together with the large presence of quartz artifacts in our sample, prevented us to pursue a microscopic use-wear approach for this study in particular.
+
+\newline
+
+```{r techno_attributes, echo=FALSE}
+
+techno_attributes <- read_csv("../data/raw_data/technological_attributes.csv")
+
+morpho_attributes[is.na(techno_attributes)] <- " "
+
+knitr::kable(techno_attributes, caption = "Technological attributes used for the analysis of scaled pieces included in this study.")
+
+```
+
+\newline
+
+```{r morpho_attributes, echo=FALSE}
+
+morpho_attributes <- read_csv("../data/raw_data/functional_attributes.csv")
+
+morpho_attributes[is.na(morpho_attributes)] <- " "
+
+knitr::kable(morpho_attributes, caption = "Morpho-functional attributes used for the analysis of each damaged platform of scaled pieces included in this study.
+*Adapted from Gonzalez-Urquijo and Ibanez-Estévez [-@gonzalez-urquijo_metodologialisis_1994].
+**Adapted from de la Peña [-@de_la_pena_sobre_2011].")
+
+```
 
@@
 All analyses and data processing were accomplished in R (version `r getRversion()`) [@r_core_team_r:_2013].
-Following recent concerns on the reproducibility of archaeological analysis we include the entire R code used for all the analysis and visualizations contained in this paper in our supplemental online material (SOM) at https://dx.doi.org/10.17605/OSF.IO/WPXGH To produce those files we followed the procedures described by Marwick et al.
+Following recent concerns on the reproducibility of archaeological analysis we include the entire R code used for all the analysis and visualizations contained in this paper in our online research compendiumm at https://dx.doi.org/10.17605/OSF.IO/WPXGH.
+To produce those files we followed the procedures described by Marwick et al.
 [-@marwick_packaging_2017] for the creation of research compendiums to enhance the reproducibility of research.
@@
 Overall, the technological analysis revealed some trends that lasted throughout the Upper Paleolithic, in agreement with the general patterns of lithic technology recorded for the site.
-Across all techno-complexes, blanks used were either flakes or unclassifiable fragments, but the choice seems to be different for each raw material.
+Across all techno-complexes, blanks used were either flakes (based on the recognition of dorsal and ventral surfaces) or unclassifiable fragments, but the choice seems to be different for each raw material.
 For quartz, in the Gravettian assemblage, the blank types are almost equally split between flakes and unclassifiable pieces (Table \@ref(tab:technogravettian)).
@@
 
-Other characteristics of the assemblage are the low frequency of retouch (n = `r techno_data %>% filter(RetouchedBlank == "Yes") %>% summarise(n())`), of fire alterations (n = `r techno_data %>% filter(Fire != "No_Traces") %>% summarise(n())`), and of the presence of original striking platforms (n = `r techno_data %>% filter(ButtType != "Absent") %>% summarise(n())`).
+Other characteristics of the assemblage are the low frequency of retouch found in the non-damaged edges (n = `r techno_data %>% filter(Retouch == "Yes") %>% summarise(n())`), fire alterations (n = `r techno_data %>% filter(Fire != "No_Traces") %>% summarise(n())`), and the presence of original striking platforms (n = `r techno_data %>% filter(ButtType != "Absent") %>% summarise(n())`).
 Still, when present, retouch is located in the lateral part of the artifacts, similar to what would define a side-scraper.
@@
 # Set variables
-var_list <- c("Blank","CrossSection","Profile","Blankshape","Cortex","ButtType","Profile")
+var_list <- c("Blank","AxisCoincident","CrossSection","SideSection","Profile","BlankShape","Cortex","ButtType","Retouch","Fire")
 
@@
 techno_data <- techno_data %>% 
-  mutate(Area = AxisWidth*AxisLength)
+  mutate(Area = TypologicalAxisWidth*TypologicalAxisLength)
 
@@
 aov_thickness <-  tidy(aov_thickness)
-
-aov_thickness$p.value
 
@@
 On the other end, a small number of chert and chalcedony artifacts (n = `r techno_data %>% filter(RawMaterial != "Quartz" & DamagedPlatforms == "3" | DamagedPlatforms == "4_or_more") %>% summarize(n())`) exhibit multiple functional axis, with three or four damaged platforms (Figure \@ref(fig:nplatforms)).
-This seems to indicate different strategies of curation for coarse and fine grain raw materials, with fine grain materials being rotated when the first used axis becomes too small and/or the edges get useless for that specific activity.
-However, when plotted against metric data (Figure \@ref(fig:plotbyplatforms)), results for the Area variable reveal that pieces with four damaged platforms are among the largest in the whole assemblages, and that there is not a visible difference between the thickness of the pieces comprising each group.
-This seems to attest that the use of several axis in the same piece is not correlated with curation occuring in later phases of artifact use but instead to a probable difference in raw material performance.
+This seems to indicate different strategies of curation for coarse (quartz) and fine grain (chert and chalcedony) raw materials, with fine grain materials being rotated when the first used axis becomes too small and/or the edges get useless for that specific activity.
+However, when plotted against metric data (Figure \@ref(fig:plotbyplatforms)), results for the Area variable reveal that pieces with four damaged platforms are among the largest in the whole assemblages, and that there is not a visible difference between the thicknesses of the pieces comprising each group.
+This seems to attest that the use of several axes in the same piece is not correlated with curation occuring in later phases of artifact use but instead to a probable difference in raw material performance.
 
@@
 
+chi_arrangement$chi_test
+
 ```
@@
 In three (Scar Arrangement, Scar Extension and Edge Delineation) of the six represented variables, the category 'Other' is one of the most frequent (above 35%) for both quartz and chert, revealing a very high variability for the combination of attributes within each variable.
-A chi-square test with modified alpha level (Bonferroni correction) to adjust for multiple testing and reduce type-I error, shows significant differences in the quartz and chert Scar Arrangement categories (*X^2^* [`r chi_arrangement$chi_test$parameter`, *N* = `r sum(chi_arrangement$chi_test$observed)`] = `r  round(chi_arrangement$chi_test$statistic, 2)`, *p* = `r round(chi_arrangement$chi_test$p.value, 3)`, *phi* = `r round(chi_arrangement$chi_size_effect$phi, 2)`), and in Scar Faciality categories (*X^2^* [`r chi_faciality$chi_test$parameter`, *N* = `r sum(chi_faciality$chi_test$observed)`] = `r  round(chi_faciality$chi_test$statistic,2)`, *p* = `r round(chi_faciality$chi_test$p.value, 3)`, *phi* = `r round(chi_faciality$chi_size_effect$phi, 2)`).
+A chi-square test with modified alpha level (Bonferroni correction) to adjust for multiple testing and reduce type-I error, shows significant differences in the quartz and chert Scar Arrangement categories (*X^2^* [`r chi_arrangement$chi_test$parameter`, *N* = `r sum(chi_arrangement$chi_test$observed)`] = `r  round(chi_arrangement$chi_test$statistic, 2)`, *p* = < 0.001, *phi* = `r round(chi_arrangement$chi_size_effect$phi, 2)`), and in Scar Faciality categories (*X^2^* [`r chi_faciality$chi_test$parameter`, *N* = `r sum(chi_faciality$chi_test$observed)`] = `r  round(chi_faciality$chi_test$statistic,2)`, *p* = `r round(chi_faciality$chi_test$p.value, 3)`, *phi* = `r round(chi_faciality$chi_size_effect$phi, 2)`).
 The effect size statistic (Phi) suggests, however, a medium practical significance for both variables.
@@
 Bipolar technology clearly had an important role on the adaptive systems of the first modern humans in Western Iberia, as well as in other european regions [see e.g.
-@villa_neandertals_2018].
+@villa_neandertals_2018; @de_la_pena_sobre_2011; @sano_functional_2012; @zilhao_o_1997].
+In Europe there is a considerable rise in the use of bipolar technologies after the Middle to Upper Paleolithic transition, and some authors consider scaled pieces as one of the most common lithic morphotypes in european Upper Paleolithic assemblages [@de_la_pena_sobre_2011].
 A rise in bipolar technologies cannot be dissociated from the diverse set of factors that made Anatomically Modern Humans thrive.
@@
 @carvalho_talhe_1998; @de_la_pena_sobre_2011; @leblanc_wedges_1992; @shott_bipolar_1989; @shott_bipolar_1999; @zilhao_o_1997].
-Based on our data we find that this concept is somehow flawed, especially when applied to Vale Boi.
+Based on our data we find that this concept does not seem to fit in Vale Boi.
 First, with the single exception of the early Gravettian (during which very small backed bladelets are present - see Marreiros et al [-@marreiros_lithic_2018]), the site’s technology shows no evidence of manufacture or use of tools with such small dimensions as the ones recorded from the scars of scaled pieces (5-8 mm).
-Second, although the most common interpretation is that these chips/bladelets would be inserted in composite throwing tools, the functionality of chips of ~7mm is quite arguable, especially regarding the impact that these would have in a perforation or extension of a bleeding scenario, as opposed to using regular bladelets.
-Third, although Vale Boi has a very rich assemblage of organic tools [@evora_osseous_2013], there is no evidence of bone or antler tools with grooves that might be used for the insertion of stone implements.
+Second, although the most common interpretation is that these chips/bladelets would be inserted in composite throwing tools, there is no evidence of bone or antler tools with grooves that might be used for the insertion of stone implements even though Vale Boi has a very rich assemblage of organic tools [@evora_osseous_2013].
+However, we still do not discard the hypothesis that these may have been made from perishable organic materials.
+Third, although previous ethnographic studies support the use of small flakes, bladelets and chips for these types of implements and other activities such as scraping, boring and cutting, we currently have no use-wear data dtah can support this hypothesis at the site.
 Finally, as previously mentioned, Vale Boi stone tool technological analyses revealed that every phase of the chaîne operatoire is present, which means that knapping was mostly occurring at the site, which is attested by the high number of chips found.
@@
 By putting all these factors in context, the hypothesis that these pieces were used for carcass processing, bone marrow extracting [@manne_intensive_2012] and other similar activities, such as working hard, organic materials with the goal of producing tools seems, thus far, to be the most reasonable interpretation.
+The use of scaled pieces as wedges would allow to further enhancing the effectiveness of these activities and has been identified in modern human occupations all over the world [@igreja_functional_2013; @langejans_middle_2012; @leblanc_wedges_1992; @sano_functional_2012; @shott_bipolar_1989].
+In fact, this technique would be extremely useful in periods of greater environmental stress and in periods when communities needed higher mobility.
+Despite this, we do not discard the possibility that some scaled pieces might have been used for bipolar reduction strategies.
+In addition to the amount of stone anvils at the site, bipolar cores are also present across all techno-complexes [@marreiros_as_2009].
+There may even be the case that blanks from bipolar cores were transformed into scaled pieces by using them for further bipolar or wedging activities.
+Unfotunately, due to the nature of the damage present in scaled pieces this is a very hard point to prove.
+Still, open hypothesis clearly attest the value and versatility of bypolar technologies for early modern human groups in the region.
 
-In Table \@ref(tab:scaledpiecesportugal) we present current data on Upper Paleolithic bipolar technology in Portugal.
+Taking this into consideration, in Table \@ref(tab:scaledpiecesportugal) we present current data on Upper Paleolithic bipolar technology in Portugal.
+It is important to note that aside from Vale Boi, all scaled pieces and bipolar cores counts in this table were made according to each author's typological definitions and classifications, and not the ones we propose for this paper.
 Although the number of scaled pieces and bipolar cores is considerably low in most of the sites, a significant number of *loci* present some kind of bipolar technology evidence.
@@
 Bipolar technology is quite common in Upper Paleolithic contexts of Western Iberia, but current interpretations are based on empirical observations, rather than on more solid analytical evidence.
+This, of course, stems in part from the lack of dedicated studies on bipolar technology in the region, an issue that can be also argued to be true for all of the European Upper Paleolithic.
 We argue that all factors mentioned in this study must be considered while interpreting each site in this region.
-While in Vale Boi there is clear evidence that the use of these artifacts would not be for bipolar knapping, in other sites the picture may be different.
+While in Vale Boi there is clear evidence that the use of these artifacts would not be primarly for bipolar knapping, in other sites the picture may be different.
 
@@
 We would like to thank a diversity of institutions that provided funding to the work in Vale Boi in the last decade.
-These are Archaeological Association of Algarve, Fundaç~ao para a Ciência e Tecnologia, Instituto Português de Arqueologia, National Geographic Society, and Wenner-Gren Foundation for Anthropological Research.
+These are Fundação para a Ciência e Tecnologia (Grants PTDC/HAH/64184/2006, SFRH/BD/65527/2009 and PTDC/HIS-ARQ/117540/2010), National Geographic Society (Grant #8045-06), and Wenner-Gren Foundation for Anthropological Research (Grant #8290).
+Thanks also to the three anonymous reviewers for their valuable comments that helped to improve the content of the paper.
 

### 2018-12-09 R/functions.R

@@ -34,6 +34,7 @@ load.packages <- function() {
   library(powerAnalysis)
   library(epitools)
   library(wesanderson)
+  library(dataMeta)
 }
 
 ###########################################################################

### 2018-12-09 analysis/paper/paper.Rmd
revised

@@
   - Nuno Bicho (ICArEHB, University of Algarve, Faro, Portugal)
-  - "*corresponding author | email: pedroehorta@gmail.com | phone: +351 911 005 506"
+  - "*corresponding author | email: pedroehorta@gmail.com"
 
@@
 Their distinction, however, has not always been consensual [@hayden_confusion_1980].
-Besides the fact that bipolar cores are often confused or lumped with scaled pieces [@villa_neandertals_2018;@de_la_pena_sobre_2011], the definition of scaled piece has suffered significant changes ever since its initial identification in the early 20th century and, to this day, there still seems to be no clear definition for this type of artifact.
+Besides the fact that bipolar cores are often confused or lumped with scaled pieces [@villa_neandertals_2018; @de_la_pena_sobre_2011], the definition of scaled piece has suffered significant changes ever since its initial identification in the early 20th century and, to this day, there still seems to be no clear definition for this type of artifact.
 The first definition of scaled pieces was proposed by Bardon and Bouyssonie [-@bardon_outils_1906], describing them as a result of bipolar knapping through direct percussion, with the “core” rested in a hard surface, originating splintering in both ends of the tool.
 Since then, several other definitions were adopted and adapted by researchers for a very diverse set of contexts across the world [e.g.
-@hayden_confusion_1980; @macdonald_debert:_1985; @octobon_contribution_1938; @shott_bipolar_1989; @sonneville-bordes_lexique_1956].
+@hayden_confusion_1980; @macdonald_debert_1985; @octobon_contribution_1938; @shott_bipolar_1989; @sonneville-bordes_lexique_1956].
 
@@
 
-```{r VBLocation, echo = FALSE, fig.cap="Location of the site of Vale Boi. Map data are from stamenmap, using the ggmap package Kahle and Wickham [-@kahle_ggmap:_2013]"}
+```{r VBLocation, echo = FALSE, fig.cap="Location of the site of Vale Boi. Map data are from Stamenmap (http://maps.stamen.com), using the ggmap package Kahle and Wickham [-@kahle_ggmap_2013]."}
 
@@
 Vale Boi’s lithic studies have revealed a general tendency for stable technological and functional patterns throughout the Upper Paleolithic [e.g.
-@bicho_ledge:_2012; @bradtmoller_lithic_2016; @cascalheira_cross-scale_2017; @cascalheira_influencia_2013; @cascalheira_vale_2012; @gibaja_funcion_2011; @marreiros_as_2009; @marreiros_lithic_2013; @marreiros_lithic_2015; @marreiros_lithic_2018].
+@bicho_ledge_2012; @bradtmoller_lithic_2016; @cascalheira_cross-scale_2017; @cascalheira_influencia_2013; @cascalheira_vale_2012; @gibaja_funcion_2011; @marreiros_as_2009; @marreiros_lithic_2013; @marreiros_lithic_2015; @marreiros_lithic_2018].
 This can be explained, partially, by raw material availability.
@@
 
-Attribute analysis was split into two main groups, each corresponding to two distinct types of features: (1) technological attributes (Table \@ref(tab:techno_attributes)) and (2) morpho-functional attributes (Table \@ref(tab:morpho_attributes)).
+Attribute analysis was split into two main groups, each corresponding to two distinct types of features: (1) technological attributes (Table \@ref(tab:technoattributes)) and (2) morpho-functional attributes (Table \@ref(tab:morphoattributes)).
 In the first group, a series of variables traditionally used in lithic studies [e.g.
-@andrefsky_jr_lithics:_2005; @inizan_technology_1999] were recorded, aiming to characterize patterns of blank choice for the application of bipolar technology.
+@andrefsky_jr_lithics_2005; @inizan_technology_1999] were recorded, aiming to characterize patterns of blank choice for the application of bipolar technology.
 For the second group of variables, a macroscopic approach building upon the work of de la Peña [-@de_la_pena_sobre_2011] and Fischer et.
@@
 
-```{r techno_attributes, echo=FALSE}
+```{r technoattributes, echo=FALSE}
 
@@
 
-morpho_attributes[is.na(techno_attributes)] <- " "
+techno_attributes[is.na(techno_attributes)] <- " "
 
-knitr::kable(techno_attributes, caption = "Technological attributes used for the analysis of scaled pieces included in this study.")
+knitr::kable(techno_attributes, caption= "Technological attributes used for the analysis of scaled pieces included in this study.")
 
@@
 
-```{r morpho_attributes, echo=FALSE}
+---
+nocite: | 
+  [@debenath_handbook_1994]
+---
+
+\newline
+
+```{r morphoattributes, echo=FALSE}
 
@@
 
-knitr::kable(morpho_attributes, caption = "Morpho-functional attributes used for the analysis of each damaged platform of scaled pieces included in this study.
+knitr::kable(morpho_attributes, caption= "Morpho-functional attributes used for the analysis of each damaged platform of scaled pieces included in this study.
 *Adapted from Gonzalez-Urquijo and Ibanez-Estévez [-@gonzalez-urquijo_metodologialisis_1994].
@@
 
-All analyses and data processing were accomplished in R (version `r getRversion()`) [@r_core_team_r:_2013].
+All analyses and data processing were accomplished in R (version `r getRversion()`) [@r_core_team_r_2013].
 Following recent concerns on the reproducibility of archaeological analysis we include the entire R code used for all the analysis and visualizations contained in this paper in our online research compendiumm at https://dx.doi.org/10.17605/OSF.IO/WPXGH.
@@
 
-knitr::kable(assemb, caption= "Frequencies of scaled pieces used in this study, by raw material and chronological period. Percentages are shown in parentheses")
+knitr::kable(assemb, caption= "Frequencies of scaled pieces used in this study, by raw material and chronological period. Percentages are shown in parentheses.")
 
@@
 Since the majority of the analyzed pieces presented a rectangular outline, we use the area of a rectangle (i.e.
-Length x Width) as an approximation for the overall dimensions of the artifacts.
+typological Length x Width) as an approximation for the overall dimensions of the artifacts.
 Area calculations revealed a maximum of `r round(max(techno_data$Area, na.rm = TRUE), digits = 2)` mm and a minimum of `r round(min(techno_data$Area, na.rm = TRUE), digits = 2)` mm.
@@
 This seems to indicate different strategies of curation for coarse (quartz) and fine grain (chert and chalcedony) raw materials, with fine grain materials being rotated when the first used axis becomes too small and/or the edges get useless for that specific activity.
-However, when plotted against metric data (Figure \@ref(fig:plotbyplatforms)), results for the Area variable reveal that pieces with four damaged platforms are among the largest in the whole assemblages, and that there is not a visible difference between the thicknesses of the pieces comprising each group.
-This seems to attest that the use of several axes in the same piece is not correlated with curation occuring in later phases of artifact use but instead to a probable difference in raw material performance.
+However, when plotted against metric data (Figure \@ref(fig:plotbyplatforms)), results for the Area variable reveal that pieces with four damaged platforms are among the largest in all assemblages, and that there is not a visible difference between the thicknesses of the pieces comprising each group.
+This seems to attest that the use of several axes in the same piece is not correlated with curation occuring in later phases of artifact use, but instead to a probable difference in raw material performance.
 
@@
 
-chi_arrangement$chi_test
-
 ```
@@
 For this reason the results presented below focus only on overall and between raw materials variability.
-Figure \@ref(fig:morphoattributes) shows the relative frequencies of all qualitative variables recorded for each damaged platform.
+Figure \@ref(fig:morphoattributesfig) shows the relative frequencies of all qualitative variables recorded for each damaged platform.
 To simplify representation, and to avoid wrong comparisons between hammered and active platforms, opposed damaged platforms were grouped into pairwise categories.
@@
 
-```{r morphoattributes, echo = FALSE, fig.cap="Frequency of morphological attributes for each raw material.
+```{r morphoattributesfig, echo = FALSE, fig.cap="Frequency of morphological attributes for each raw material.
 Opposed damaged platforms were grouped so that each artifact was only counted once and to avoid wrong comparisons between active a hammered platforms.
@@
 These dimensions present an amount of inertia slightly greater than that obtained by the 0.95-quantile of random distributions (37.74% against 34.37%).
-Still, it is a rather small percentage, somehow attesting the high variability suggested by our interpretation of Figure \@ref(fig:morphoattributes), and suggesting that patterns identified by previous studies [e.g.
+Still, it is a rather small percentage, somehow attesting the high variability suggested by our interpretation of Figure \@ref(fig:morphoattributesfig), and suggesting that patterns identified by previous studies [e.g.
 @de_la_pena_qualitative_2015; @de_la_pena_sobre_2011], in which certain combination of attributes were used to discriminate scaled pieces functionalities, are difficult to apply to the assemblage used in this study.
@@
 Despite this, we do not discard the possibility that some scaled pieces might have been used for bipolar reduction strategies.
-In addition to the amount of stone anvils at the site, bipolar cores are also present across all techno-complexes [@marreiros_as_2009].
+In addition to the amount of stone anvils at the site, bipolar cores are also present across all techno-complexes [@cascalheira_tecnologia_2010; @marreiros_as_2009].
 There may even be the case that blanks from bipolar cores were transformed into scaled pieces by using them for further bipolar or wedging activities.

### 2018-12-12 analysis/paper/paper.Rmd
packrat added

@@
 Despite this, and even after one century of being identified there is still no consensus on how to define, analyze or interpret these tools.
-In western Iberia there is a clear lack of comprehensive studies regarding this type of artifacts at a regional scale.
-In this paper we present a first techno-morphological analysis of a sample of scaled pieces from the Upper Paleolithic site of Vale Boi.
+In western Iberia, there is a clear lack of comprehensive studies regarding this type of artifacts at a regional scale.
+In this paper, we present a first techno-morphological analysis of a sample of scaled pieces from the Upper Paleolithic site of Vale Boi.
 Our first aim was to build upon existing analytical models in order to identify function and possible reduction strategies for these artifacts.
@@
 By comparing our data with other author’s we found that current models could not be applied to the archaeological record, as the attribute variability is too high.
-Furthermore, in this region we found that higher frequencies of bipolar technology can be found related to residential sites due to both functional and cultural patterns.
+Furthermore, in this region, we found that higher frequencies of bipolar technology can be found related to residential sites due to both functional and cultural patterns.
 While we still cannot define a specific function for these artifacts (intermediate pieces or wedges for working hard raw materials or cores for the extraction of chips and small bladelets), it is clear that they had a major role in the variability and intensification of resource exploitation during the Upper Paleolithic in western Iberia.
@@
 Their distinction, however, has not always been consensual [@hayden_confusion_1980].
-Besides the fact that bipolar cores are often confused or lumped with scaled pieces [@villa_neandertals_2018; @de_la_pena_sobre_2011], the definition of scaled piece has suffered significant changes ever since its initial identification in the early 20th century and, to this day, there still seems to be no clear definition for this type of artifact.
-The first definition of scaled pieces was proposed by Bardon and Bouyssonie [-@bardon_outils_1906], describing them as a result of bipolar knapping through direct percussion, with the “core” rested in a hard surface, originating splintering in both ends of the tool.
+Besides the fact that bipolar cores are often confused or lumped with scaled pieces [@villa_neandertals_2018; @de_la_pena_sobre_2011], the definition of scaled piece has suffered significant changes ever since its initial identification in the early 20th century and, to this day, there still seems to be no world-wide accepted clear definition for this type of artifact.
+The first definition of scaled pieces was proposed by Bardon and Bouyssonie [-@bardon_outils_1906], describing them as a result of bipolar knapping through direct percussion, with the “core” resting on a hard surface, originating splintering in both ends of the tool.
 Since then, several other definitions were adopted and adapted by researchers for a very diverse set of contexts across the world [e.g.
@@
 @binford_indian_1963; @de_la_pena_sobre_2011; @flood_moth_1980; @igreja_functional_2013; @leblanc_wedges_1992; @lucas_les_2004; @shott_bipolar_1999; @tixier_typologie_1963].
-The issue surrounding this problem lies within the functional equifinality of these artifacts.
-Contrary to bipolar cores that are exclusively part of a technological reduction sequence with the goal of blank extraction, scaled pieces have been associated with two distinct types of activities: (1) as intermediate pieces or wedges for working hard organic raw materials (e.g.
+The issue surrounding this problem lies in the functional equifinality of these artifacts.
+Contrary to bipolar cores, that are exclusively part of a technological reduction sequence with the goal of blank extraction [e.g.
+@binford_indian_1963; @crabtree_introduction_1972; @leaf_variation_1979], scaled pieces have been associated with two distinct types of activities: (1) as intermediate pieces or wedges for working hard organic raw materials (e.g.
 bone, ivory, antler); and (2) as cores for the extraction of chips, small flakes and bladelets [@brantingham_early_2004].
@@
 
-Most current studies on scaled pieces apply use-wear methods [e.g.
+Many current studies on scaled pieces apply use-wear methods [e.g.
 @bader_lithic_2015; @de_la_pena_interpretation_2015; @de_la_pena_new_2014; @de_la_pena_qualitative_2015; @de_la_pena_sobre_2011; @gibaja_tecnologiy_2007; @igreja_functional_2013; @lucas_les_2004; @sano_functional_2012], focusing on the identification of polishes and use-wear patterned stigmas, through both micro and macroscopic analysis of splintered surfaces and its comparison with experimental assemblages.
@@
 
-While we agree that some of the elements might have been used as cores, these interpretations seem simplistic, and the authors seem to have little to no analytical evidence to back their hypothesis, other than the inverse relationship mentioned above.
+While we agree that some of the elements might have been used as cores, these interpretations seem simplistic, those studies seem to have little to no analytical evidence to back those hypotheses, other than the inverse relationship mentioned above.
 This results, in part, from the lack of comprehensive studies regarding this type of artifacts at a regional scale.
-Additionally, the contexts from where most of these artifacts were recovered did not have good organic preservation or dedicated use-wear studies and, thus, no direct association between these tools and the exploitation of hard organic raw materials is possible, and, in face of the nonexistence of any absolute dates, their precise chronological attribution is also, most of the times, unreliable.
+Additionally, the contexts from where most of these artifacts were recovered did not have good organic preservation or dedicated use-wear studies and, thus, no direct association between these tools and the exploitation of hard organic raw materials is possible, and, in face of the nonexistence of any absolute dates, their precise chronological attribution is also, frequently, unreliable.
 
@@
 The archaeological site of Vale Boi is located in the western edge of the Algarve region (southern Portugal) (Figure \@ref(fig:VBLocation)).
-The site can be found in a small valley following a river north-south for 2 km, until it reaches the Atlantic Ocean.
+The site can be found in a small valley following a river north-south for 2 km until it reaches the Atlantic Ocean.
 Archaeological deposits occupy an estimated area of over 10 000 sq.
-meters across a stepped slope, marked at the top by a 10 meter-high limestone cliff face [@bicho_ledge_2012; @bicho_paleolithique_2010].
+meters across a stepped slope marked at the top by a 10 meter-high limestone cliff face [@bicho_ledge_2012; @bicho_paleolithique_2010].
 
@@
 Elongated products are found at low frequencies across all chronologies.
-When present, retouched tools are mostly notches, denticulates, end-scrapers, and, although not retouched artifacts *per* *se* scaled pieces.
+When present, retouched tools are mostly notches, denticulates, end-scrapers, and, although not retouched artifacts *per* *se*, scaled pieces.
 One possible reason for these simplistic strategies seems to be an overall low knapping quality, since nodules are quite small, and frequently show a high degree of tectonically derived fractures [@pereira_territory_2016].
@@
 Based on previous works [e.g.
-@hayden_confusion_1980] here we define scaled pieces as artifacts of variable size and morphology, showing traces of crushing and splitering of edges at opposite ends, caused by direct percussion at one end, and subsequent crushing of the other for being rested on a hard surface.
-Scaled pieces present always two opposite surfaces, just like a flake or blade would.
+@hayden_confusion_1980] here we define scaled pieces as artifacts of variable size and morphology, showing traces of crushing and splintering of edges at opposite ends, caused by direct percussion at one end, and subsequent crushing of the other for being rested on a hard surface.
+Scaled pieces present always two opposite surfaces, just like a flake or blade would, but at least some of them show signs of crushing.
 Crushing traces can be bifacially or unifacially distributed.
-Some scaled pieces do not present crushing in opposed platforms, but are still classified as scaled pieces here.
+Some scaled pieces do not present crushing in opposed platforms but are still classified as scaled pieces here.
 This detail has been previously referred by Villa et al.
@@
 The striking platform is the surface that is struck with a hammerstone to produce blanks.
-Tipically, it exhibits little crushing except near the point of impact.
+Typically, it exhibits little crushing except near the point of impact.
 Large flake scars tend to originate at the striking platform.
@@
 Studies by Sano [-@sano_functional_2012], and Gibaja et al.
-[-@gibaja_tecnologiy_2007] indicate that, since use-wear traces are formed before the splintering, the removal of small chips removes most of the polishes and use-wear traces left by the contact with the static element (either a stone anvil or a hard-organic materials).
+[-@gibaja_tecnologiy_2007] indicate that, since use-wear traces are formed before the splintering, the removal of small chips removes most of the polishes and use-wear traces left by the contact with the static element (either a stone anvil or hard-organic materials).
 This, together with the large presence of quartz artifacts in our sample, prevented us to pursue a microscopic use-wear approach for this study in particular.
@@
 ```
-
-\newline
-
----
-nocite: | 
-  [@debenath_handbook_1994]
----
 
@@
 Following the work of de la Peña [-@de_la_pena_qualitative_2015; -@de_la_pena_sobre_2011] and Gonzalez-Urquijo and Ibanez-Estévez [-@gonzalez-urquijo_metodologialisis_1994], we expected that these attributes would be indicative of which function the artifacts had.
-For instance, according to de la Peña [-@de_la_pena_qualitative_2015; -@de_la_pena_sobre_2011], for pieces used as wedges the delineation of the damaged platforms are constantly asymmetrical and only the hammered edge clearly shows the typical *écaille* retouch.
+For instance, according to de la Peña [-@de_la_pena_qualitative_2015; -@de_la_pena_sobre_2011], for pieces used as wedges, the delineation of the damaged platforms are constantly asymmetrical and only the hammered edge clearly shows the typical *écaille* retouch.
 Furthermore, these pieces would have irregular shapes, variable scar size, and irregular scar distribution.
@@
 All analyses and data processing were accomplished in R (version `r getRversion()`) [@r_core_team_r_2013].
-Following recent concerns on the reproducibility of archaeological analysis we include the entire R code used for all the analysis and visualizations contained in this paper in our online research compendiumm at https://dx.doi.org/10.17605/OSF.IO/WPXGH.
+Following recent concerns on the reproducibility of archaeological analysis, we include the entire R code used for all the analysis and visualizations contained in this paper in our online research compendium at https://dx.doi.org/10.17605/OSF.IO/WPXGH.
 To produce those files we followed the procedures described by Marwick et al.
 [-@marwick_packaging_2017] for the creation of research compendiums to enhance the reproducibility of research.
-The files provided contain all the raw data used in our analysis as well as a custom R package [@wickham_r_2015] holding the code use for all analysis and to produce all tables and figures.
+The files provided contain all the raw data used in our analysis as well as a custom R package [@wickham_r_2015] holding the code used for all analysis and to produce all tables and figures.
 To enable maximum re-use, our code is released under the MIT license, our data as CC-0, and our figures as CC-BY, [for more information see @marwick_computational_2016].
@@
 Technological and morphological characteristics of the flake blanks present very similar patterns across time and among raw materials.
-The blanks sought after would have straight profiles, parallel edges, no cortical surfaces and trapezoidal or triangular shaped cross-sections.
+The blanks sought after would have straight profiles, parallel edges, no cortical surfaces, and trapezoidal or triangular shaped cross-sections.
 
@@
 The absence of the original blank platforms is to be expected in this type of artifact, mostly due to the functional use of the pieces, and consequent removal of the platform, rather than an actual choice.
-The large absence of platforms may, also, be the result of the use of the technological axis as main functional axis.
+The large absence of platforms may, also, be the result of the use of the technological axis as the main functional axis.
 In fact, when identifiable, technological and typological axes coincide in `r inline.perc(techno_data, AxisCoincident, "Yes")`% of the cases.
@@
 typological Length x Width) as an approximation for the overall dimensions of the artifacts.
-Area calculations revealed a maximum of `r round(max(techno_data$Area, na.rm = TRUE), digits = 2)` mm and a minimum of `r round(min(techno_data$Area, na.rm = TRUE), digits = 2)` mm.
-For thickness the maximum is `r round(max(techno_data$Thickness, na.rm = TRUE), digits = 2)` mm and the minimum is `r round(min(techno_data$Thickness, na.rm = TRUE), digits = 2)` mm.
-In general, mean values tend to be similar between raw materials, with some differences occurring within the Proto-Solutrean and Magdalenian assemblages, most certainly as a result of the small samples analysed for each of these periods (Figure \@ref(fig:meanplots)).
+Area calculations revealed a maximum of `r round(max(techno_data$Area, na.rm = TRUE), digits = 2)` mm^2^ and a minimum of `r round(min(techno_data$Area, na.rm = TRUE), digits = 2)` mm^2^.
+For thickness, the maximum is `r round(max(techno_data$Thickness, na.rm = TRUE), digits = 2)` mm and the minimum is `r round(min(techno_data$Thickness, na.rm = TRUE), digits = 2)` mm.
+In general, mean values tend to be similar between raw materials, with some differences occurring within the Proto-Solutrean and Magdalenian assemblages, most certainly as a result of the small samples analyzed for each of these periods (Figure \@ref(fig:meanplots)).
 Across techno-complexes, however, no significant statistical differences were detected when using an ANOVA test for both Area (*F* (`r aov_area$df[1]`, `r aov_area$df[2]`) = `r aov_area$statistic[1]`, *p* = `r aov_area$p.value[1]`, *d* = `r cohens_area$cohens.f`), and Thickness  (*F* (`r aov_thickness$df[1]`, `r aov_thickness$df[2]`) = `r aov_thickness$statistic[1]`, *p* = `r aov_thickness$p.value[1]`, *d* = `r cohens_thickness$cohens.f`).
@@
 In every assemblage quartz pieces were exclusively used in one single axis, exhibiting only two damaged platforms.
-On the other end, a small number of chert and chalcedony artifacts (n = `r techno_data %>% filter(RawMaterial != "Quartz" & DamagedPlatforms == "3" | DamagedPlatforms == "4_or_more") %>% summarize(n())`) exhibit multiple functional axis, with three or four damaged platforms (Figure \@ref(fig:nplatforms)).
+On the other end, a small number of chert and chalcedony artifacts (n = `r techno_data %>% filter(RawMaterial != "Quartz" & DamagedPlatforms == "3" | DamagedPlatforms == "4_or_more") %>% summarize(n())`) exhibit multiple functional axes, with three or four damaged platforms (Figure \@ref(fig:nplatforms)).
 This seems to indicate different strategies of curation for coarse (quartz) and fine grain (chert and chalcedony) raw materials, with fine grain materials being rotated when the first used axis becomes too small and/or the edges get useless for that specific activity.
-However, when plotted against metric data (Figure \@ref(fig:plotbyplatforms)), results for the Area variable reveal that pieces with four damaged platforms are among the largest in all assemblages, and that there is not a visible difference between the thicknesses of the pieces comprising each group.
-This seems to attest that the use of several axes in the same piece is not correlated with curation occuring in later phases of artifact use, but instead to a probable difference in raw material performance.
+However, when plotted against metric data (Figure \@ref(fig:plotbyplatforms)), results for the Area variable reveal that pieces with four damaged platforms are among the largest in all assemblages and that there is not a visible difference between the thicknesses of the pieces comprising each group.
+This seems to attest that the use of several axes in the same piece is not correlated with curation occurring in later phases of artifact use, but instead to a probable difference in raw material performance.
 
@@
 Similarly to what was registered for the technological attributes, most of the differences occur between raw materials rather than between chronologies.
-For this reason the results presented below focus only on overall and between raw materials variability.
+For this reason, the results presented below focus only on overall and between raw materials variability.
 Figure \@ref(fig:morphoattributesfig) shows the relative frequencies of all qualitative variables recorded for each damaged platform.
@@
 
-In three (Scar Arrangement, Scar Extension and Edge Delineation) of the six represented variables, the category 'Other' is one of the most frequent (above 35%) for both quartz and chert, revealing a very high variability for the combination of attributes within each variable.
+In three (Scar Arrangement, Scar Extension, and Edge Delineation) of the six represented variables, the category 'Other' is one of the most frequent (above 35%) for both quartz and chert, revealing a very high variability for the combination of attributes within each variable.
 A chi-square test with modified alpha level (Bonferroni correction) to adjust for multiple testing and reduce type-I error, shows significant differences in the quartz and chert Scar Arrangement categories (*X^2^* [`r chi_arrangement$chi_test$parameter`, *N* = `r sum(chi_arrangement$chi_test$observed)`] = `r  round(chi_arrangement$chi_test$statistic, 2)`, *p* = < 0.001, *phi* = `r round(chi_arrangement$chi_size_effect$phi, 2)`), and in Scar Faciality categories (*X^2^* [`r chi_faciality$chi_test$parameter`, *N* = `r sum(chi_faciality$chi_test$observed)`] = `r  round(chi_faciality$chi_test$statistic,2)`, *p* = `r round(chi_faciality$chi_test$p.value, 3)`, *phi* = `r round(chi_faciality$chi_size_effect$phi, 2)`).
@@
 
-For quartz artifacts the most common patterns are the combination of central/total scar distributions, both platforms with angles wider than 45º, and a combination of either unifacial/unifacial or unifacial/bifacial scar distribution.
+For quartz artifacts, the most common patterns are the combination of central/total scar distributions, both platforms with angles wider than 45º, and a combination of either unifacial/unifacial or unifacial/bifacial scar distribution.
 Chert pieces, on the other hand, more typically present opposed platforms with damage occupying the whole platform width, and a weaker presence of unifacial/unifacial scar edge facial distribution.
@@
 ```{r morphoattributesfig, echo = FALSE, fig.cap="Frequency of morphological attributes for each raw material.
-Opposed damaged platforms were grouped so that each artifact was only counted once and to avoid wrong comparisons between active a hammered platforms.
+Opposed damaged platforms were grouped so that each artifact was only counted once and to avoid wrong comparisons between active and hammered platforms.
 A - Distribution of damage; B - Arrangement of scars; C - Extension of scars; D - Delineation of damaged platforms; E - Facial distribution of scars; F - Angle of damaged platforms."}
@@
 The first two dimensions of the MCA express 20.92% of the total dataset inertia, meaning that only that percentage of the total variability is explained by the plane combining the first two dimensions.
-An inspection to the screeplot presented in Figure \@ref(fig:screeplot) suggests restricting the analysis to the description of the first 4 dimensions.
+An inspection of the screeplot presented in Figure \@ref(fig:screeplot) suggests restricting the analysis to the description of the first 4 dimensions.
 These dimensions present an amount of inertia slightly greater than that obtained by the 0.95-quantile of random distributions (37.74% against 34.37%).
 Still, it is a rather small percentage, somehow attesting the high variability suggested by our interpretation of Figure \@ref(fig:morphoattributesfig), and suggesting that patterns identified by previous studies [e.g.
-@de_la_pena_qualitative_2015; @de_la_pena_sobre_2011], in which certain combination of attributes were used to discriminate scaled pieces functionalities, are difficult to apply to the assemblage used in this study.
+@de_la_pena_qualitative_2015; @de_la_pena_sobre_2011], in which certain combinations of attributes were used to discriminate scaled pieces functionalities, are difficult to apply to the assemblage used in this study.
 
@@
 
-```{r screeplot, echo = FALSE, fig.cap="Mutiple Correspondence Analysis screeplot."}
+```{r screeplot, echo = FALSE, fig.cap="Multiple Correspondence Analysis screeplot."}
 
@@
 
-Bipolar technology clearly had an important role on the adaptive systems of the first modern humans in Western Iberia, as well as in other european regions [see e.g.
+Bipolar technology clearly had an important role on the adaptive systems of the first modern humans in Western Iberia, as well as in other European regions [see e.g.
 @villa_neandertals_2018; @de_la_pena_sobre_2011; @sano_functional_2012; @zilhao_o_1997].
-In Europe there is a considerable rise in the use of bipolar technologies after the Middle to Upper Paleolithic transition, and some authors consider scaled pieces as one of the most common lithic morphotypes in european Upper Paleolithic assemblages [@de_la_pena_sobre_2011].
+In Europe, there is a considerable rise in the use of bipolar technologies after the Middle to Upper Paleolithic transition, and some authors consider scaled pieces as one of the most common lithic morphotypes in European Upper Paleolithic assemblages [@de_la_pena_sobre_2011].
 A rise in bipolar technologies cannot be dissociated from the diverse set of factors that made Anatomically Modern Humans thrive.
@@
 Other indicators are both the constant presence of thermally altered quartz fragments that might be linked to stone boiling activities, and a large number of greywacke slabs with impact marks revealing their use as anvils [@cascalheira_cross-scale_2017].
-For carcass processing and bone marrow extraction the use of scaled pieces would allow for a better control of bone fracture.
+For carcass processing and bone marrow extraction, the use of scaled pieces would allow for a better control of bone fracture.
 In fact, this should be significantly better than using a hard hammer directly on the bone, since the latter technique may either over fracture the bone or even crush it due to the lack of precision.
@@
 
-In addition, scaled pieces in Vale Boi may also be associated to bone tool production, given that Vale Boi is one of the portuguese Upper Paleolithic sites with significant evidence for onsite production and use of bone tools [@evora_osseous_2013].
+In addition, scaled pieces in Vale Boi may also be associated to bone tool production, given that Vale Boi is one of the Portuguese Upper Paleolithic sites with significant evidence for onsite production and use of bone tools [@evora_osseous_2013].
 According to both Leblanc [-@leblanc_wedges_1992] and Shott [-@shott_bipolar_1999] wedges are needed to work bone, antler, and wood, and stone wedges are preferable than other types of material.
@@
 Based on our data we find that this concept does not seem to fit in Vale Boi.
-First, with the single exception of the early Gravettian (during which very small backed bladelets are present - see Marreiros et al [-@marreiros_lithic_2018]), the site’s technology shows no evidence of manufacture or use of tools with such small dimensions as the ones recorded from the scars of scaled pieces (5-8 mm).
+First, with the single exception of the early Gravettian (during which very small backed bladelets are present - see Marreiros et al.
+[-@marreiros_lithic_2018]), the site’s technology shows no evidence of manufacture or use of tools with such small dimensions as the ones recorded from the scars of scaled pieces (5-8 mm).
 Second, although the most common interpretation is that these chips/bladelets would be inserted in composite throwing tools, there is no evidence of bone or antler tools with grooves that might be used for the insertion of stone implements even though Vale Boi has a very rich assemblage of organic tools [@evora_osseous_2013].
 However, we still do not discard the hypothesis that these may have been made from perishable organic materials.
-Third, although previous ethnographic studies support the use of small flakes, bladelets and chips for these types of implements and other activities such as scraping, boring and cutting, we currently have no use-wear data dtah can support this hypothesis at the site.
-Finally, as previously mentioned, Vale Boi stone tool technological analyses revealed that every phase of the chaîne operatoire is present, which means that knapping was mostly occurring at the site, which is attested by the high number of chips found.
+Third, although previous ethnographic studies [e.g.
+@flenniken_replicative_1981-1; @shott_bipolar_1989] support the use of small flakes, bladelets, and chips for these types of implements and other activities such as scraping, boring and cutting, we currently have no use-wear data that can support this hypothesis at the site.
+Finally, as previously mentioned, Vale Boi stone tool technological analyses revealed that every phase of the *chaîne* *operatoire* is present, which means that knapping was mostly occurring at the site, which is attested by the high number of chips found.
 Considering that most knapping activities originate chipage, the need of a specific tool for the sole purpose of extracting them does not seem viable.
@@
 There may even be the case that blanks from bipolar cores were transformed into scaled pieces by using them for further bipolar or wedging activities.
-Unfotunately, due to the nature of the damage present in scaled pieces this is a very hard point to prove.
-Still, open hypothesis clearly attest the value and versatility of bypolar technologies for early modern human groups in the region.
+Unfortunately, due to the nature of the damage present in scaled pieces, this is a very hard point to prove.
+Still, open hypothesis clearly attests the value and versatility of bipolar technologies for early modern human groups in the region.
 
@@
 Although the number of scaled pieces and bipolar cores is considerably low in most of the sites, a significant number of *loci* present some kind of bipolar technology evidence.
-Moreover, it is clear a discrepancy when comparing the presence of scaled pieces versus bipolar cores, the latter being well under represented than the first.
+Moreover, it is clear a discrepancy when comparing the presence of scaled pieces versus bipolar cores, the latter being well underrepresented than the first.
 It is also clear that scaled pieces show on average a higher representation in Vale Boi’s assemblages than on most other sites.
-Further, of the three bipolar cores in all sites, 2 of these come from Vale Boi.
+Further, of the three bipolar cores in all sites, two of these come from Vale Boi.
 While currently there are no definitive data on the representation of scaled pieces within the retouched tool assemblage of the Slope area, it is fairly safe to assume that it should be a particularly high value.
@@
 
-knitr::kable(scaledpiecesportugal, caption= "Frequencies of scaled pieces and bipolar cores in portuguese Upper Paleolithic sites.")
+knitr::kable(scaledpiecesportugal, caption= "Frequencies of scaled pieces and bipolar cores in Portuguese Upper Paleolithic sites.")
 
@@
 We argue that all factors mentioned in this study must be considered while interpreting each site in this region.
-While in Vale Boi there is clear evidence that the use of these artifacts would not be primarly for bipolar knapping, in other sites the picture may be different.
+While in Vale Boi there is clear evidence that the use of these artifacts would not be primarily for bipolar knapping, in other sites the picture may be different.
 
@@
 
-# Acknowledgements
+# Acknowledgments
 We would like to thank a diversity of institutions that provided funding to the work in Vale Boi in the last decade.
@@
 Thanks also to the three anonymous reviewers for their valuable comments that helped to improve the content of the paper.
+
+---
+nocite: | 
+  [@debenath_handbook_1994]
+  [@horta_tecnologia_2016]
+  [@zilhao_o_1995]
+---
 

### 2018-12-12 R/functions.R

@@ -34,7 +34,6 @@ load.packages <- function() {
   library(powerAnalysis)
   library(epitools)
   library(wesanderson)
-  library(dataMeta)
 }
 
 ###########################################################################

### 2018-12-13 analysis/paper/paper.Rmd
packrat final, preprint submision

@@
 @binford_indian_1963; @de_la_pena_sobre_2011; @flood_moth_1980; @igreja_functional_2013; @leblanc_wedges_1992; @lucas_les_2004; @shott_bipolar_1999; @tixier_typologie_1963].
-The issue surrounding this problem lies in the functional equifinality of these artifacts.
+The issue surrounding this problem lies within the functional equifinality of these artifacts.
 Contrary to bipolar cores, that are exclusively part of a technological reduction sequence with the goal of blank extraction [e.g.
@@
 
-As in many other regions and stone age periods across the world [e.g.
+As in many other regions and Stone Age periods across the world [e.g.
 @diez-martin_were_2009; @igreja_functional_2013; @langejans_middle_2012; @white_fabricators_1968], evidence of bipolar technologies are quite ubiquitous in European Upper Paleolithic contexts.
@@
 
-While we agree that some of the elements might have been used as cores, these interpretations seem simplistic, those studies seem to have little to no analytical evidence to back those hypotheses, other than the inverse relationship mentioned above.
+While we agree that some of the elements might have been used as cores, these interpretations seem simplistic, and those studies seem to have little to no analytical evidence to back those hypotheses, other than the inverse relationship mentioned above.
 This results, in part, from the lack of comprehensive studies regarding this type of artifacts at a regional scale.
@@
 @hayden_confusion_1980] here we define scaled pieces as artifacts of variable size and morphology, showing traces of crushing and splintering of edges at opposite ends, caused by direct percussion at one end, and subsequent crushing of the other for being rested on a hard surface.
-Scaled pieces present always two opposite surfaces, just like a flake or blade would, but at least some of them show signs of crushing.
+Scaled pieces present always two opposite surfaces, just like a flake or blade would, but at least one of them shows signs of crushing.
 Crushing traces can be bifacially or unifacially distributed.
-Some scaled pieces do not present crushing in opposed platforms but are still classified as scaled pieces here.
+Some scaled pieces do not present crushing in opposed platforms, but are still classified as scaled pieces here.
 This detail has been previously referred by Villa et al.
@@
 
-All lithic artifacts recovered from the Slope area of Vale Boi that fit the scaled piece definition presented below were considered for this study, independent of raw material or technological class.
+All lithic artifacts recovered from the Slope area of Vale Boi that fit the scaled piece definition presented above were considered for this study, independent of raw material or technological class.
 
@@
 By putting all these factors in context, the hypothesis that these pieces were used for carcass processing, bone marrow extracting [@manne_intensive_2012] and other similar activities, such as working hard, organic materials with the goal of producing tools seems, thus far, to be the most reasonable interpretation.
-The use of scaled pieces as wedges would allow to further enhancing the effectiveness of these activities and has been identified in modern human occupations all over the world [@igreja_functional_2013; @langejans_middle_2012; @leblanc_wedges_1992; @sano_functional_2012; @shott_bipolar_1989].
+The use of scaled pieces as wedges would allow for further enhancement of the effectiveness of these activities and has been identified in modern human occupations all over the world [@igreja_functional_2013; @langejans_middle_2012; @leblanc_wedges_1992; @sano_functional_2012; @shott_bipolar_1989].
 In fact, this technique would be extremely useful in periods of greater environmental stress and in periods when communities needed higher mobility.
@@
 Unfortunately, due to the nature of the damage present in scaled pieces, this is a very hard point to prove.
-Still, open hypothesis clearly attests the value and versatility of bipolar technologies for early modern human groups in the region.
+Still, open hypotheses clearly attest the value and versatility of bipolar technologies for early modern human groups in the region.
 

### 2018-12-13 R/functions.R

@@ -32,13 +32,11 @@ load.packages <- function() {
   library(tab)
   library(sjstats)
   library(powerAnalysis)
-  library(epitools)
   library(wesanderson)
 }
 
 ###########################################################################
 # READ DATA ----------------------------------------------------------------
-
 
 #' read.data
 #'
